### 2020-04-19 manuscript/ms.Rmd
start writing manuscript

---
title: "Quantifying social distancing behaviour across Australia in response to COVID-19"
author:
  - Nick Golding$^1$, ...$^2$ & ...$^3$
  - $^1$ Telethon Kids Institute and Curtin University, Perth WA, Australia
  - $^2$ ...
  - $^3$ ...
bibliography: distancing.bib
output:
  pdf_document:
  word_document:
    fig_caption: true
    highlight: null
header-includes: \usepackage{float}
---

## Gist (to be split up into a paper format)

Social distancing behaviour - advised and enforced by governments around the world - has been a cornerstone of the global response to the COVID-19 pandemic. Quantifying how the behaviour of a population has changed in response to these public health measures, is critical to monitoring the ongoing effectiveness of these measures, identifying which are most impactful, and as inputs to modelling to forecast how the measures will aimpact ont he course of the outbreak.

A variety of data sources have are available that can quantify different aspects of social distancing behaviour at a population level. Aggregated user-tracking data from technology companies provide timeseries of: how often users leave home, the lengths of journeys they undertake when they do, the types of locations they visited, and the locations of those places.

Whilst these multiple data sources provide varied perspectives on population behaviour, they vary significantly in the subset of the population they reflect and the aspects of behaviour being considered. For example, data from route-planning apps is more likely to reflect unfamiliar journeys than trips to places regularly visited, and data on the types of places being visited does not indicate how far users travel to visit them. The different, and often non-overlapping spatial scales and time frames of these data sources also hinders direct comparison and synthesis of them.

Moreover, no single one of these types of data fully encompasses the range of behaviours that affect transmission of COVID-19 - those that can be sumamrised in a mathematical modelling framework via the mixing parameter of a transmission model.

The aim of this work is to synthesise these varied measures of population mixing behaviour into a small, quantified set of latent social distancing factors that reflect the underlying changes in behaviour. We consider these latent social distancing factors in a semi-mechanistic way; reflecting knowledge and hypotheses of the events that triggered and influenced these behavioural trends. To do this, we construct a statistical semi-mechanistic latent factor model that considers dates that events, policy measures, and public announcements are made, and estimates how the behaviour of the population responded to each events by calibration against all of the available metrics of population mixing behaviour.

We apply this model to 9 separate data streams provided by from two technology companies for Australia, with spatial data varying between national, state and city-levels. This reveals two distinct latent population behavioural factors: a strong social distancing effect in response to national announcements and policies, preceeded by a small and short-lived rise in mixing apparently driven by the population 'stocking up' on supplies.

To add:
 - relative effectiveness of three different policies
 - relative effect size by state

## Abstract

 - Social distancing 
 - Data available
 - Synthesis needed
 - Findings
 - Use for monitoring

## Introduction

 - COVID-19 background
 - Social distancing policies
 - Population mixing/mobility studies
 - Timeline of data availability

## Methods

 - General model description
   - latent factor model structure (weights)
   - construction of latent factors
   - observation model (weekends and holidays)
 - Data availability by source
 - Announcement timeline for Australia
 - Specific model fitting for Australia
    - interaction between distancing and weekends
    - greta HMC
    - residuals checking

## Results

Figures:
 - Two latent factors nationally
 - Latent factors superimposed on different data sources
 - Loadings for each data type (ordination-style plot of similarities between them? After taking absolute value)
 - Relative impact of different announcements (summary of ranking in posterior?)
 - Variation in latent factors by state

## Discussion

 - Use for monitoring adherence.
 - Use in modelling studies.
 - Caveat that the population is biased to users of these technologies.
 - Potential to include more data sources.

### 2020-04-19 manuscript/ms.Rmd
add notation to methods

@@
 
- - General model description
-   - latent factor model structure (weights)
-   - construction of latent factors
-   - observation model (weekends and holidays)
+### General model description
+
+#### Latent factor model
+
+We construct a mechanistic latent-factor model that relates a series of underlying ('latent') population behaviours to an array of data streams, and then infer its parameters using statistical inference.
+This section, gives a general formulation of the model.
+Details of the specific implementation and data sources for Australia are given in a later section.
+
+For each data stream $i$, we observe a time series of data values $\mathbf{y}_{i}$, comprising the values of the metric at each time $t$.
+We model the overall trend $\mathbf{m}_i$ in each of these data streams as a function of $K$ different temporal latent factors $\mathbf{z}_k$.
+
+Each datastream trend is modelled as a linear function of the latent behavioural factors, with intercept term $\alpha_i$ and loading values $\omega_{i,k}$ against each latent factor:
+
+$$
+m_{i,t} = \alpha_i + \sum_{k=1}^K \omega_{i,k} z_{k,t}
+$$
+The latent factors $\mathbf{z}_k$ quantify general patterns of behaviour, rather than any one specific measurement of behaviour.
+As such, they are unitless, relative values which for statistical convenience we constrain to be scaled between 0 and 1.
+The sign and magnitude of the loadings $\omega_{i,k}$ therefore determine whether each behaviour corresponds to an increase or decrease in that metric, and by what magnitude.
+
+#### Mechanistic latent factors
+
+  We model each of these latent behavioural factors in a semi-mechanistic fashion.
+We compute the degree of behavioural change as a function of the timing, magnitude and early/late adoption of behavioural responses to events.
+For each latent factor $k$, we model the impact of each event $j$ as a function of: $\tau_{j,k}$ the time when the event took place; $\epsilon_{j,k}$ the relative impact of the event on behaviour; and $\kappa_{j,k}$, the variability among the population in how rapidly individuals responded to these events.
+
+  For each latent behaviour $k$, these parameters describe the behavioural reponse to each event via a logistic distribution.
+Where the behaviour occurs at a single point in time, these parameters give rise to a logistic distribution over the times at which members of the population exhibit this behaviour, where the fraction of the population exhibiting behaviour $k$ in response to event $j$ at time $t$ is given by the probability density function of the logistic distribution:
+
+$$
+p_{j,k}(t) = \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{\kappa_{j,k} (1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
+$$
+
+  Where the latent behavioural state is described by individuals switching from one behavioural state to another, the fraction of the population that have made the switch by time $t$ is given by the cumulative density function of the logistic distribution:
+
+$$
+p_{j,k}(t) = \frac{1}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
+$$
+
+  Since the latent factors are scaled between 0 and 1, we scale these probabilities such that their maximum value is the relative event impact parameter $\epsilon_{j,k}$, and the sum over these parameeters is constrained to 1.
+Therefore, where the behaviour is a function of only a single event, $\epsilon_{j,k}$ is fixed at 1.
+The contribution of each event to the behaviour is therefore given by:
+
+<!-- MAKE THIS HAVE DENSITY EPSILON AT THE MEAN -->
+<!-- HAS DENSITY 1/(4 * kappa) AT MEAN -->
+<!-- SO DIVIDE BY THAT AND MULTIPLY BY EPSILON -->
+<!-- MUTIPLY BY epsilon * 4 * kappa -->
+
+<!-- CHECK THIS! -->
+$$
+f_{j,k}(t) = 4 \kappa_{j,k} \epsilon_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{\kappa_{j,k} (1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
+$$
+And the cumulative contribution by:
+
+$$
+f_{j,k}(t) = \frac{\epsilon_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
+$$
+
+#### Data observatrion models
+
+   - observation model (weekends, holidays, and interactions)
+
  - Data availability by source

### 2020-04-19 manuscript/ms.Rmd
expand on the behavioural probability distribution stuff

@@
 
-  For each latent behaviour $k$, these parameters describe the behavioural reponse to each event via a logistic distribution.
-Where the behaviour occurs at a single point in time, these parameters give rise to a logistic distribution over the times at which members of the population exhibit this behaviour, where the fraction of the population exhibiting behaviour $k$ in response to event $j$ at time $t$ is given by the probability density function of the logistic distribution:
+  For each latent behaviour $k$, these parameters describe the behavioural reponse to each event via a rescaled logistic probability distribution, with distribution location parameter $\tau_{j,k}$ and scale parameter $\kappa_{j,k}$, and a rescaling factor $\epsilon_{j,k}$.
+This logistic probability distribution can be used in two ways useful for modelling behaviour related to COVID-19.
+For behaviours that individuals enact at a specific point in time - such as increased grocery shopping, or travelling to a particular residence in preparation for social distancing - the probability density function gives the fraction of the population that enacts these behaviours at a given time.
+For behavioural states that individuals 'switch' to enacting at some point in time - such as increased social distancing behaviour - the cumulative density function gives the fraction of the population that have switched to this new behavioural state by a given time.
+
+  These probabilities are then rescaled such that their maximum value is the relative event impact parameter $\epsilon_{j,k}$.
+This parameter is proportional to the fraction of the population that enacts a behaviour $k$ in response to event $j$.
+However, the data streams considered only provide information on *relative* behavioural changes among a subset of the population, there is therefore no information against which to calibrate an estimate of the *absolute* fraction of the population enacting a behaviour.
+Parameters $\epsilon_{j,k}$ are therefore constrained so that their sum over all events for a given behaviour is equal to 1; if only a single event is considered, $\epsilon_{j,k}$ is fixed at 1.
+The contribution of each event $j$ to an episodic behaviour is therefore given by:
 
 $$
-p_{j,k}(t) = \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{\kappa_{j,k} (1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
+f_{j,k}(t) = 4 \epsilon_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{(1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
 $$
 
-  Where the latent behavioural state is described by individuals switching from one behavioural state to another, the fraction of the population that have made the switch by time $t$ is given by the cumulative density function of the logistic distribution:
+and the contribution to a state-switching behaviour is given by:
 
 $$
-p_{j,k}(t) = \frac{1}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
-$$
-
-  Since the latent factors are scaled between 0 and 1, we scale these probabilities such that their maximum value is the relative event impact parameter $\epsilon_{j,k}$, and the sum over these parameeters is constrained to 1.
-Therefore, where the behaviour is a function of only a single event, $\epsilon_{j,k}$ is fixed at 1.
-The contribution of each event to the behaviour is therefore given by:
-
-<!-- MAKE THIS HAVE DENSITY EPSILON AT THE MEAN -->
-<!-- HAS DENSITY 1/(4 * kappa) AT MEAN -->
-<!-- SO DIVIDE BY THAT AND MULTIPLY BY EPSILON -->
-<!-- MUTIPLY BY epsilon * 4 * kappa -->
-
-<!-- CHECK THIS! -->
-$$
-f_{j,k}(t) = 4 \kappa_{j,k} \epsilon_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{\kappa_{j,k} (1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
-$$
-And the cumulative contribution by:
-
-$$
-f_{j,k}(t) = \frac{\epsilon_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
+g_{j,k}(t) = \frac{\epsilon_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
 $$

### 2020-04-19 R/functions.R
switch to using Google csv

@@ -334,20 +334,23 @@ p_values <- Vectorize(one_p_value, c("n_crisis", "n_baseline"))
 # - remove the grocery and pharmacy category
 # (affected by panic buying, not interventions)
 google_mobility <- function() {
-  file <-
-    "https://raw.githubusercontent.com/goldingn/google-location-coronavirus/AUS/2020-04-11-au_state.tsv"
-  data <- readr::read_tsv(file) %>%
+  url <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
+  data <- readr::read_csv(url, ) %>%
+    filter(country_region == "Australia") %>%
+    tidyr::pivot_longer(ends_with("_percent_change_from_baseline"),
+                        names_to = "category",
+                        values_to = "trend") %>%
+    mutate(source = "Google") %>%
     dplyr::select(
-      state = sub_region_name,
+      state = sub_region_1,
       category = category,
       date = date,
       trend = trend
     ) %>%
-    filter(category != "Grocery & pharmacy") %>%
-    mutate(category = ifelse(category == "Workplace",
-                             "Workplaces",
-                             category)) %>%
-    mutate(source = "Google")
+    mutate(category = str_remove_all(category, "_percent_change_from_baseline"),
+           category = str_replace_all(category, "_", " "),
+    ) %>%
+    filter(category != "grocery and pharmacy")
   
   data
 }

### 2020-04-19 R/latent_social_distancing.R

@@ -155,5 +155,4 @@ title(main = "Social distancing effect",
       col.main = "darkseagreen4")
 
 # - do multiple states with hierarchical weights
-# - account for public holidays in each state
-# - get national estiamte with population weights
+# - get national estimate with population weights

### 2020-04-19 manuscript/ms.Rmd
add observation model

--- 2020-04-19 manuscript/ms.Rmd
@@
 We consider these latent social distancing factors in a semi-mechanistic way; reflecting knowledge and hypotheses of the events that triggered and influenced these behavioural trends.
-To do this, we construct a statistical semi-mechanistic latent factor model that considers dates that events, policy measures, and public announcements are made, and estimates how the behaviour of the population responded to each events by calibration against all of the available metrics of population mixing behaviour.
+To do this, we construct a statistical semi-mechanistic latent factor model that considers dates that events, policy measures, and public announcements are made, and estimates how the behaviour of the population responded to each event by calibration against all of the available metrics of population mixing behaviour.
 
@@
 
- - General model description
-   - latent factor model structure (weights)
-   - construction of latent factors
-   - observation model (weekends and holidays)
+### General model description
+
+#### Latent factor model
+
+We construct a mechanistic latent-factor model that relates a series of underlying ('latent') population behaviours to an array of data streams, and then infer its parameters using statistical inference.
+This section, gives a general formulation of the model.
+Details of the specific implementation and data sources for Australia are given in a later section.
+
+For each data stream $i$, we observe a time series of data values $\mathbf{y}_{i}$, comprising the values of the metric at each time $t$.
+We model the overall trend $\mathbf{m}_i$ in each of these data streams as a function of $K$ different temporal latent factors $\mathbf{z}_k$.
+
+Each datastream trend is modelled as a linear function of the latent behavioural factors, with loading values $\omega_{i,k}$ against each latent factor:
+
+$$
+m_{i,t} = \sum_{k=1}^K \omega_{i,k} z_{k,t}
+$$
+The latent factors $\mathbf{z}_k$ quantify general patterns of behaviour, rather than any one specific measurement of behaviour.
+As such, they are unitless, relative values which for statistical convenience we constrain to be scaled between 0 and 1.
+The sign and magnitude of the loadings $\omega_{i,k}$ therefore determine whether each behaviour corresponds to an increase or decrease in that metric, and by what magnitude.
+
+#### Mechanistic latent factors
+
+  We model each of these latent behavioural factors in a semi-mechanistic fashion.
+We compute the degree of behavioural change as a function of the timing, magnitude and early/late adoption of behavioural responses to triggers - such as announcements or imposed restrictions.
+For each latent factor $k$, we model the impact of each trigger $j$ as a function of: $\tau_{j,k}$ the time when the trigger occurred; $\epsilon_{j,k}$ the relative impact of the event on behaviour; and $\kappa_{j,k}$, the variability among the population in how rapidly individuals responded to these events.
+
+  For each latent behaviour $k$, these parameters describe the behavioural reponse to each trigger via a rescaled logistic probability distribution, with distribution location parameter $\tau_{j,k}$ and scale parameter $\kappa_{j,k}$, and a rescaling factor $\epsilon_{j,k}$.
+This logistic probability distribution can be used in two ways for two different types of behaviour related to COVID-19.
+For behavioural events - behaviours that individuals engage in at a specific point in time (such as bulk grocery shopping, or travelling to a particular residence in preparation for social distancing) the probability density function gives the fraction of the population that engages in the behaviours at a given time.
+For behavioural shifts - where individuals 'switch' from one behavioural state to another at some point in time (such as shifting from their baseline state to one of increased social distancing behaviour) the cumulative density function gives the fraction of the population that have shifted to this new behavioural state by a given time.
+
+  These probabilities are then rescaled such that their maximum value is the relative event impact parameter $\epsilon_{j,k}$.
+This parameter is proportional to the fraction of the population that enacts a behaviour $k$ in response to event $j$.
+However, the data streams considered only provide information on *relative* behavioural changes among a subset of the population - there is therefore no information against which to calibrate an estimate of the *absolute* fraction of the population engaging in a behaviour.
+Parameters $\epsilon_{j,k}$ are therefore constrained so that their sum over all events for a given behaviour is equal to 1.
+If only a single event is considered, $\epsilon_{j,k}$ is fixed at 1.
+With this rescaling, the contribution to latent factor $k$ of behavioural events associated with trigger $j$ is given by:
+
+$$
+f_{j,k}(t) = 4 \epsilon_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{(1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
+$$
+
+and the contribution of behavioural shifts associated with trigger $j$ is given by:
+
+$$
+g_{j,k}(t) = \frac{\epsilon_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
+$$
+The latent factors can then be computed by summing the contributions of all $J$ triggers. For a latent factor representing behavioural events this would be:
+$$
+z_{k,t} = \sum_{j=1}^J  f_{j,k}(t)
+$$
+and for behavioural shifts:
+
+$$
+z_{k,t} = \sum_{j=1}^J g_{j,k}(t)
+$$
+
+#### Observation models
+
+We assume that the temporal trend for each datastream is driven by these latent behavioural states.
+To complete the model, we define an observation model for the datastreams, comprising sampling error and drivers of patterns in the data stream that are not driven by COVID-19-associated behaviours - such as weekends and public holidays.
+We assume that each data stream observation $y_{i,t}$ is subject to independent Gaussian error with variance $\sigma^2$ (i.e.
+common to all data sources) and expected value $\mu_{i,t}$, given by the modelled trend $m_{i,t}$ plus a linear model to account for non-COVID-19 effects:
+<!-- \begin{align} -->
+$$
+y_{i,t} \sim N(\mu_{i,t}, \sigma^2) \\ 
+\mu_{i,t} = m_{i,t} + \alpha_i + \mathbf{x}_t \beta_i
+$$
+<!-- \end{align} -->
+
+where $\alpha_i$ is an intercept term, $\mathbf{x}_t$ is a data vector giving the values of the $M$ covariates (e.g.
+indicator variables for whether $t$ falls on a weekend or public holiday), and $\beta_i$ is a corresponding vector of datastream-specific regression coefficients for the main effects of these covariates.
+
+####
+
+When presented in matrix form, the model can be seen to be a standard linear latent factor model:
+CHANGE e TO \epsilon AND RENAME \epsilon
+$$
+\mathbf{y} =  \mathbf{x} \beta + \mathbf{z} \omega + \mathbf{e}\\
+e_{i,t} \sim N(0, \sigma^2)
+$$
+where $\mathbf{x}$ is an $N$-by-$M+1$ design matrix of covariates, including a column of ones to represent the intercept terms, $\beta$ is an $M-1$-by-$N$ matrix of regression coefficients, $\mathbf{z}$ is a $T$-by-$K$ matrix of behavioural latent factors and $\omega$ is a $K$-by-$N$ matrix of loadings between the latent factors and data streams.
+
  - Data availability by source
@@
     - interaction between distancing and weekends
+
+<!-- $$ -->
+<!-- y_{i,t} \sim N(\mu_{i,t}, \sigma^2) \\  -->
+<!-- \mu_{i,t} = \alpha_i + m_{i,t} + \mathbf{x}_t \beta_i + m_{i,t} \mathbf{x}_t \gamma_i -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- y_{i,t} \sim N(\mu_{i,t}, \sigma^2) \\  -->
+<!-- \mu_{t} = \alpha + \mathbf{m}_{t} + \mathbf{x}_t \beta + (\mathbf{m}_{t} \odot \mathbf{x}_t) \gamma \\  -->
+<!-- \mathbf{m} = \mathbf{z} \omega -->
+<!-- $$ -->
+
+<!-- $\gamma_i$ is a second vector of coefficints for the interactions between these covariates and the trend.
+These interaction terms allow the model to reflect the fact that the impact of of these other drivers may diminish or be exacerbated by COVID-19 -related behaviours, for example a reduced difference in behaviours between weekends and weekdays when the population is exhibiting strong social distancing behaviour.
+-->
+
     - greta HMC

### 2020-04-20 manuscript/ms.Rmd
rename some variables

@@
 We compute the degree of behavioural change as a function of the timing, magnitude and early/late adoption of behavioural responses to triggers - such as announcements or imposed restrictions.
-For each latent factor $k$, we model the impact of each trigger $j$ as a function of: $\tau_{j,k}$ the time when the trigger occurred; $\epsilon_{j,k}$ the relative impact of the event on behaviour; and $\kappa_{j,k}$, the variability among the population in how rapidly individuals responded to these events.
+For each latent factor $k$, we model the impact of each trigger $j$ as a function of: $\tau_{j,k}$ the time when the trigger occurred; $\lambda_{j,k}$ the relative impact of the event on behaviour; and $\kappa_{j,k}$, the variability among the population in how rapidly individuals responded to these events.
 
-  For each latent behaviour $k$, these parameters describe the behavioural reponse to each trigger via a rescaled logistic probability distribution, with distribution location parameter $\tau_{j,k}$ and scale parameter $\kappa_{j,k}$, and a rescaling factor $\epsilon_{j,k}$.
+  For each latent behaviour $k$, these parameters describe the behavioural reponse to each trigger via a rescaled logistic probability distribution, with distribution location parameter $\tau_{j,k}$ and scale parameter $\kappa_{j,k}$, and a rescaling factor $\lambda_{j,k}$.
 This logistic probability distribution can be used in two ways for two different types of behaviour related to COVID-19.
@@
 
-  These probabilities are then rescaled such that their maximum value is the relative event impact parameter $\epsilon_{j,k}$.
+  These probabilities are then rescaled such that their maximum value is the relative event impact parameter $\lambda_{j,k}$.
 This parameter is proportional to the fraction of the population that enacts a behaviour $k$ in response to event $j$.
 However, the data streams considered only provide information on *relative* behavioural changes among a subset of the population - there is therefore no information against which to calibrate an estimate of the *absolute* fraction of the population engaging in a behaviour.
-Parameters $\epsilon_{j,k}$ are therefore constrained so that their sum over all events for a given behaviour is equal to 1.
-If only a single event is considered, $\epsilon_{j,k}$ is fixed at 1.
+Parameters $\lambda_{j,k}$ are therefore constrained so that their sum over all events for a given behaviour is equal to 1.
+If only a single event is considered, $\lambda_{j,k}$ is fixed at 1.
 With this rescaling, the contribution to latent factor $k$ of behavioural events associated with trigger $j$ is given by:
@@
 $$
-f_{j,k}(t) = 4 \epsilon_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{(1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
+f_{j,k}(t) = 4 \lambda_{j,k} \frac{e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}{(1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}) ^ 2}
 $$
@@
 $$
-g_{j,k}(t) = \frac{\epsilon_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
+g_{j,k}(t) = \frac{\lambda_{j,k}}{1 + e^{-(t - \tau_{j,k}) / \kappa_{j,k}}}
 $$
@@
 When presented in matrix form, the model can be seen to be a standard linear latent factor model:
-CHANGE e TO \epsilon AND RENAME \epsilon
 $$
-\mathbf{y} =  \mathbf{x} \beta + \mathbf{z} \omega + \mathbf{e}\\
-e_{i,t} \sim N(0, \sigma^2)
+\mathbf{y} =  \mathbf{x} \beta + \mathbf{z} \omega + \epsilon\\
+\epsilon_{t,i} \sim N(0, \sigma^2)
 $$
-where $\mathbf{x}$ is an $N$-by-$M+1$ design matrix of covariates, including a column of ones to represent the intercept terms, $\beta$ is an $M-1$-by-$N$ matrix of regression coefficients, $\mathbf{z}$ is a $T$-by-$K$ matrix of behavioural latent factors and $\omega$ is a $K$-by-$N$ matrix of loadings between the latent factors and data streams.
+where $\mathbf{x}$ is an $N$-by-$M+1$ design matrix of covariates, including a column of ones to represent the intercept terms, $\beta$ is an $M-1$-by-$N$ matrix of regression coefficients, $\mathbf{z}$ is a $T$-by-$K$ matrix of behavioural latent factors, $\omega$ is a $K$-by-$N$ matrix of loadings between the latent factors and data streams, and $\epsilon$ is a $T$-by-$N$ matrix of independent Gaussian errors.
 

### 2020-04-20 R/latent_social_distancing.R

@@ -29,11 +29,11 @@ n_interventions <- n_distinct(interventions)
 n_states <- n_distinct(na.omit(data$state))
 
 # model the impacts of those interventions on social distancing factor k are
-# length of the tails for early- and late-adopters; e is the relative
+# length of the tails for early- and late-adopters; lambda is the relative
 # contribution of the three interventions
 k <- 1 / uniform(0, 16, dim = n_interventions)
 props <- uniform(0, 1, dim = n_interventions)
-e <- props / sum(props)
+lambda <- props / sum(props)
 
 # get regression weights for each category
 trend_weights <- normal(0, 10, dim = n_categories)
@@ -50,7 +50,7 @@ date_num <- dates - first_date
 intervention_date_num <- interventions$date - first_date
 lags <- outer(date_num, intervention_date_num, FUN = "-")
 lag_delays <- ilogit(sweep(lags, 2, 2 * k, FUN = "*"))
-epsilon <- 1 - sum(e) + lag_delays %*% e
+epsilon <- 1 - sum(lambda) + lag_delays %*% lambda
 
 # get a vector of weekendiness for each day (lowest on Sundays, average of 0)
 doy <- lubridate::wday(dates)
@@ -103,7 +103,7 @@ distribution(aus_data$trend) <- normal(mean = trends[idx],
                                        sd = sigma_obs)
 
 # fit model
-m <- model(e, k, trend_weights, weekend_weights, weekend_trend_weights)
+m <- model(lambda, k, trend_weights, weekend_weights, weekend_trend_weights)
 draws <- mcmc(m, chains = 10)
 
 # check convergence

### 2020-04-20 R/functions.R
add apple datastreams to model

@@ -337,21 +337,21 @@ google_mobility <- function() {
   url <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
   data <- readr::read_csv(url, ) %>%
     filter(country_region == "Australia") %>%
-    tidyr::pivot_longer(ends_with("_percent_change_from_baseline"),
-                        names_to = "category",
-                        values_to = "trend") %>%
-    mutate(source = "Google") %>%
+    tidyr::pivot_longer(
+      ends_with("_percent_change_from_baseline"),
+      names_to = "category",
+      values_to = "trend"
+    ) %>%
     dplyr::select(
       state = sub_region_1,
       category = category,
       date = date,
       trend = trend
     ) %>%
-    mutate(category = str_remove_all(category, "_percent_change_from_baseline"),
-           category = str_replace_all(category, "_", " "),
-    ) %>%
-    filter(category != "grocery and pharmacy")
-  
+    mutate(
+      category = str_remove_all(category, "_percent_change_from_baseline"),
+      category = str_replace_all(category, "_", " ")
+    )
   data
 }
 
@@ -382,25 +382,59 @@ apple_mobility <- function() {
   
   url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev8/v1/en-us/applemobilitytrends-2020-04-17.csv"
   data <- readr::read_csv(url) %>%
-    tidyr::pivot_longer(cols = starts_with("2020-"),
-                        names_to = "date",
-                        values_to = "trend") %>%
-    mutate(date = lubridate::date(date)) %>%
-    filter(region %in% ideal_regions)
+    tidyr::pivot_longer(
+      cols = starts_with("2020-"),
+      names_to = "date",
+      values_to = "trend"
+    ) %>%
+    mutate(
+      date = lubridate::date(date)
+    ) %>%
+    filter(
+      region %in% ideal_regions
+    )
   
   # add on a state label, rename the transportation type to 'category' to match
   # google, and add on the data source
   data <- data %>%
-    mutate(state = case_when(region == "Sydney" ~ "New South Wales",
-                             region == "Melbourne" ~ "Victoria",
-                             region == "Brisbane" ~ "Queensland",
-                             region == "Perth" ~ "Western Australia",
-                             TRUE ~ as.character(NA))) %>%
-    rename(category = transportation_type) %>%
-    mutate(source = "Apple")
+    mutate(
[…]

### 2020-04-20 R/latent_social_distancing.R

@@ -7,26 +7,26 @@ library(lubridate)
 library(greta)
 source("R/functions.R")
 
-# load Google mobility data, intervention dates, and public holiday dates
-data <- google_mobility()
+# load mobility streams, and potential drivers
+mobility <- all_mobility()
 interventions <- intervention_dates()
 holidays <- holiday_dates()
 populations <- state_populations()
 
 # # subset to all-Australia
-# data <- data %>%
+# mobility <- mobility %>%
 #   filter(is.na(state)) %>%
 #   dplyr::select(-state)
 
-# get vectors of date ranges and categories to model
-first_date <- min(data$date)
-last_date <- max(data$date)
+# get vectors of date ranges and datastreams to model
+first_date <- min(mobility$date)
+last_date <- max(mobility$date)
 dates <- seq(first_date, last_date, by = 1)
-categories <- unique(data$category)
-n_categories <- n_distinct(data$category)
+datastreams <- unique(mobility$datastream)
+n_datastreams <- n_distinct(mobility$datastream)
 n_dates <- n_distinct(dates)
 n_interventions <- n_distinct(interventions)
-n_states <- n_distinct(na.omit(data$state))
+n_states <- n_distinct(na.omit(mobility$state))
 
 # model the impacts of those interventions on social distancing factor k are
 # length of the tails for early- and late-adopters; lambda is the relative
@@ -35,12 +35,12 @@ k <- 1 / uniform(0, 16, dim = n_interventions)
 props <- uniform(0, 1, dim = n_interventions)
 lambda <- props / sum(props)
 
-# get regression weights for each category
-trend_weights <- normal(0, 10, dim = n_categories)
-weekend_weights <- normal(0, 10, dim = n_categories)
-weekend_trend_weights <- normal(0, 10, dim = n_categories)
-trend_intercepts <- normal(0, 10, dim = n_categories)
-holiday_weights <- normal(0, 10, dim = n_categories)
+# get regression weights for each datastream
+trend_weights <- normal(0, 10, dim = n_datastreams)
+weekend_weights <- normal(0, 10, dim = n_datastreams)
+weekend_trend_weights <- normal(0, 10, dim = n_datastreams)
+trend_intercepts <- normal(0, 10, dim = n_datastreams)
+holiday_weights <- normal(0, 10, dim = n_datastreams)
 
[…]

### 2020-04-20 R/functions.R
include apple data and second latent factor

@@ -380,7 +380,7 @@ apple_mobility <- function() {
     "Darwin"
   )
   
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev8/v1/en-us/applemobilitytrends-2020-04-17.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev9/v1/en-us/applemobilitytrends-2020-04-18.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -393,6 +393,11 @@ apple_mobility <- function() {
     filter(
       region %in% ideal_regions
     )
+  
+  # transform to percentage change
+  data <- data %>%
+    mutate(trend = trend - 100)
+  
   
   # add on a state label, rename the transportation type to 'category' to match
   # google, and add on the data source
@@ -415,13 +420,13 @@ all_mobility <- function() {
   # load datasets and label their datastreams separately
   google <- google_mobility() %>%
     mutate(
-      datastream = str_c("google: time spent at ", category)
+      datastream = str_c("google: time at\n", category)
     ) %>%
     dplyr::select(-category)
   
   apple <- apple_mobility() %>%
     mutate(
-      datastream = str_c("apple: direction requests for ", transportation_type)
+      datastream = str_c("apple: directions for\n", transportation_type)
     ) %>%
     dplyr::select(
       -geo_type,
@@ -596,6 +601,17 @@ holiday_dates <- function() {
     mutate(date = lubridate::date(date))
 }
 
+# define a latent factor for a single symmetric distribution of behavioural events
+latent_behavioural_event <- function(date_num, tau, kappa = normal(3, 1, truncation = c(0, Inf)), lambda = 1) {
+  # work with logs for numerical stability
+  # e <- exp(-lag / kappa)
+  # lambda * e / kappa * (1 + e) ^ 2
+  lag <- date_num - tau
+  le <- -lag / kappa
+  log_bump <- log(4 * lambda) + le - 2 * log1pe(le)
+  exp(log_bump)
+}
+
 state_populations <- function() {
   tibble::tribble(
     ~state, ~population,
@@ -611,8 +627,19 @@ state_populations <- function() {
 }
 
 # summarise the posterior for a vector greta array
-summarise_vec_posterior <- function(vector, draws) {
[…]

### 2020-04-20 R/latent_social_distancing.R

@@ -1,22 +1,18 @@
 # model overall movement changes in Australia, with parameteric latent factor
 # model on Google mobility data
 
+source("R/functions.R")
 library(dplyr)
 library(readr)
 library(lubridate)
 library(greta)
-source("R/functions.R")
+library(RColorBrewer)
 
 # load mobility streams, and potential drivers
 mobility <- all_mobility()
 interventions <- intervention_dates()
 holidays <- holiday_dates()
 populations <- state_populations()
-
-# # subset to all-Australia
-# mobility <- mobility %>%
-#   filter(is.na(state)) %>%
-#   dplyr::select(-state)
 
 # get vectors of date ranges and datastreams to model
 first_date <- min(mobility$date)
@@ -25,32 +21,64 @@ dates <- seq(first_date, last_date, by = 1)
 datastreams <- unique(mobility$datastream)
 n_datastreams <- n_distinct(mobility$datastream)
 n_dates <- n_distinct(dates)
-n_interventions <- n_distinct(interventions)
 n_states <- n_distinct(na.omit(mobility$state))
 
-# model the impacts of those interventions on social distancing factor k are
-# length of the tails for early- and late-adopters; lambda is the relative
-# contribution of the three interventions
-k <- 1 / uniform(0, 16, dim = n_interventions)
-props <- uniform(0, 1, dim = n_interventions)
-lambda <- props / sum(props)
+# model the social distancing latent factoras a funciton of various triggers
+# kappa are length of the tails for early- and late-adopters; lambda are the
+# relative contribution of the triggers; tau are dates of triggers (known)
+n_triggers <- n_distinct(interventions)
+kappa_distancing <- normal(2, 1, truncation = c(0, Inf), dim = n_triggers)
+props_distancing <- uniform(0, 1, dim = n_triggers)
+lambda_distancing <- props_distancing / sum(props_distancing)
+
+date_num <- as.numeric(dates - first_date)
+tau_distancing <- as.numeric(interventions$date - first_date)
+lags_distancing <- outer(date_num, tau_distancing, FUN = "-")
+lag_delays <- ilogit(sweep(lags_distancing, 2, kappa_distancing, FUN = "/"))
+distancing <- lag_delays %*% lambda_distancing
+
[…]

### 2020-04-21 R/functions.R
add facebook national datastreams

@@ -1,334 +1,45 @@
 library(readr)
-library(fs)
-library(purrr)
 library(dplyr)
 library(stringr)
-library(sf)
-library(raster)
 
 # read in and tidy up Facebook movement data
-load_fb_data <- function() {
+facebook_mobility <- function() {
   
-  # load all datasets
-  data <- load_fb_csvs() 
-  
-  # fill in centroid coordinates for LGAs known to be missing them
-  #   Meander Valley (M) 146.4871 -41.62647
-  #   Hobart (C) 147.2873 -42.89699
-  #   Tasman (M) 147.8344 -43.05465
-  # see find_centroid()
-  data <- data %>%
-    mutate(start_y = case_when(
-      start_polygon_names == "Meander Valley (M)" ~ -41.62647,
-      start_polygon_names == "Hobart (C)" ~ -42.89699,
-      start_polygon_names == "Tasman (M)" ~ -43.05465,
-      TRUE ~ start_y
-    )) %>%
-    mutate(start_x = case_when(
-      start_polygon_names == "Meander Valley (M)" ~ 146.4871,
-      start_polygon_names == "Hobart (C)" ~  147.2873,
-      start_polygon_names == "Tasman (M)" ~ 147.8344,
-      TRUE ~ start_x
-    ))%>%
-    mutate(end_y = case_when(
-      end_polygon_names == "Meander Valley (M)" ~ -41.62647,
-      end_polygon_names == "Hobart (C)" ~ -42.89699,
-      end_polygon_names == "Tasman (M)" ~ -43.05465,
-      TRUE ~ end_y
-    )) %>%
-    mutate(end_x = case_when(
-      end_polygon_names == "Meander Valley (M)" ~ 146.4871,
-      end_polygon_names == "Hobart (C)" ~  147.2873,
-      end_polygon_names == "Tasman (M)" ~ 147.8344,
-      TRUE ~ end_x
-    ))
-    
-  # fill in the state from the name stack, and scrape suffixes from the lga
-  # names
-  data <- data %>%
-    mutate(
-      start_state = str_split_fixed(start_name_stack, ' // ', n = 2)[, 1],
-      end_state = str_split_fixed(end_name_stack, ' // ', n = 2)[, 1]
+  file <- "data/fb_data/au_gadm_mobility_statistics.20200416.csv"
+  data <- readr::read_csv(file) %>%
+    dplyr::select(
+      state = polygon_name,
+      date = ds,
[…]

### 2020-04-21 R/latent_social_distancing.R

@@ -3,12 +3,11 @@
 
 source("R/functions.R")
 library(dplyr)
-library(readr)
 library(lubridate)
 library(greta)
 library(RColorBrewer)
 
-# load mobility streams, and potential drivers
+# load mobility datastreams, and potential drivers
 mobility <- all_mobility()
 interventions <- intervention_dates()
 holidays <- holiday_dates()

### 2020-04-21 R/functions.R
convert fb staying still metric to % change; and use a back-to-work latent factor in place of an intercept

@@ -18,10 +18,29 @@ facebook_mobility <- function() {
       names_to = "metric",
       values_to = "trend"
     ) %>%
+    mutate(date = lubridate::date(date))
+  
+  # set the staying home variable against a baseline of the first two weeks
+  baseline <- data %>%
+    filter(date < lubridate::date("2020-03-15")) %>%
+    group_by(state, metric) %>%
+    summarise(baseline = mean(trend))
+  
+  data <- data %>%
+    left_join(baseline) %>%
+    mutate(
+      corrected = (trend - baseline) / abs(baseline)
+    ) %>%
+    mutate(
+      trend = ifelse(metric == "staying still", corrected, trend)
+    ) %>%
+    select(
+      -corrected,
+      -baseline,
+      ) %>%
     mutate(
       trend = trend * 100,
-    ) %>%
-    mutate(date = lubridate::date(date))
+    )
   
   # add a composite national trend with population weights
   relative_population <- state_populations() %>%
@@ -36,8 +55,10 @@ facebook_mobility <- function() {
     ungroup() %>%
     mutate(state = NA)
   
-  bind_rows(national_data,
-            data)
+  bind_rows(
+    national_data,
+    data
+  )
   
 }
 
@@ -319,6 +340,28 @@ holiday_dates <- function() {
   ) %>%
     mutate(date = lubridate::date(date))
 }
+
+# define a latent factor for state-switching behaviour. kappa are lengths of the
+# tails for early- and late-adopters; lambda are the relative contribution of
+# the triggers; tau are dates of triggers
+latent_behaviour_switch <- function(date_num,
+                                    tau,
+                                    kappa = normal(2, 1, truncation = c(0, Inf), dim = length(tau)),
+                                    lambda = uniform(0, 1, dim = length(tau))) {
+  
+  if (length(tau) == 1) {
+    lag <- date_num - tau
+    result <- ilogit(lag / kappa)
+  } else {
+    lambda <- lambda / sum(lambda)
+    lags <- outer(date_num, tau, FUN = "-")
+    lag_delays <- ilogit(sweep(lags, 2, kappa, FUN = "/"))
+    result <- lag_delays %*% lambda
+  }
+  result
+  
+}
+
 
 # define a latent factor for a single symmetric distribution of behavioural events
 latent_behavioural_event <- function(date_num, tau, kappa = normal(3, 1, truncation = c(0, Inf)), lambda = 1) {

### 2020-04-21 R/latent_social_distancing.R

@@ -22,38 +22,19 @@ n_datastreams <- n_distinct(mobility$datastream)
 n_dates <- n_distinct(dates)
 n_states <- n_distinct(na.omit(mobility$state))
 
-# model the social distancing latent factoras a funciton of various triggers
-# kappa are length of the tails for early- and late-adopters; lambda are the
-# relative contribution of the triggers; tau are dates of triggers (known)
-n_triggers <- n_distinct(interventions)
-kappa_distancing <- normal(2, 1, truncation = c(0, Inf), dim = n_triggers)
-props_distancing <- uniform(0, 1, dim = n_triggers)
-lambda_distancing <- props_distancing / sum(props_distancing)
-
+# model the social distancing latent factor as a function of behavioural
+# switches triggered by the different interventions
 date_num <- as.numeric(dates - first_date)
-tau_distancing <- as.numeric(interventions$date - first_date)
-lags_distancing <- outer(date_num, tau_distancing, FUN = "-")
-lag_delays <- ilogit(sweep(lags_distancing, 2, kappa_distancing, FUN = "/"))
-distancing <- lag_delays %*% lambda_distancing
+trigger_date_num <- as.numeric(interventions$date - first_date)
+distancing <- latent_behaviour_switch(date_num, trigger_date_num)
 
 # (add terms for the potential waning of distancing, around an unknown date)
 
-# build the pre-distancing surge in mobility 
-
-# prior that timing of pre-distancing bump peak is around time of first
-# restriction - but we aren't sure about that.
-
-# one near the first restriction
-tau_bump <- normal(min(tau_distancing), 1)
+# build the pre-distancing surge in mobility with a prior that it peaks around
+# the time of the first restriction
+tau_bump <- normal(min(trigger_date_num), 1)
 kappa_bump <- normal(3, 1, truncation = c(0, Inf))
 bump <- latent_behavioural_event(date_num, tau_bump, kappa_bump)
-
-# # another bump before this one
-# tau_bump_diff <- normal(10, 7, truncation = c(0, Inf))
-# tau_bump2 <- tau_bump - tau_bump_diff
-# kappa_bump2 <- normal(3, 1, truncation = c(0, Inf))
-# bump2 <- latent_behavioural_event(date_num, tau_bump2, kappa_bump2)
-
 
 # sims <- calculate(bum2, nsim = 100)[[1]][, , 1]
[…]

### 2020-04-21 R/functions.R
move covariates into latent factor part of the model

@@ -374,6 +374,16 @@ latent_behavioural_event <- function(date_num, tau, kappa = normal(3, 1, truncat
   exp(log_bump)
 }
 
+latent_spline <- function (x, knots = 3) {
+  # build a spline latent factor for weekly variation constrained to 0-1
+  bases <- splines::bs(x, df = knots)
+  weights <- normal(0, 1, dim = knots)
+  spline <- bases %*% weights
+  spline <- spline - min(spline)
+  spline <- spline / max(spline)
+  spline
+}
+
 state_populations <- function() {
   tibble::tribble(
     ~state, ~population,

### 2020-04-21 R/latent_social_distancing.R

@@ -22,95 +22,59 @@ n_datastreams <- n_distinct(mobility$datastream)
 n_dates <- n_distinct(dates)
 n_states <- n_distinct(na.omit(mobility$state))
 
-# model the social distancing latent factor as a function of behavioural
-# switches triggered by the different interventions
+# social distancing latent factor as a function of behavioural switches
+# triggered by the major interventions
 date_num <- as.numeric(dates - first_date)
 trigger_date_num <- as.numeric(interventions$date - first_date)
 distancing <- latent_behaviour_switch(date_num, trigger_date_num)
 
 # (add terms for the potential waning of distancing, around an unknown date)
 
-# build the pre-distancing surge in mobility with a prior that it peaks around
-# the time of the first restriction
+# latent factor for pre-distancing surge in mobility with a prior that it peaks
+# around the time of the first restriction
 tau_bump <- normal(min(trigger_date_num), 1)
 kappa_bump <- normal(3, 1, truncation = c(0, Inf))
 bump <- latent_behavioural_event(date_num, tau_bump, kappa_bump)
 
-# sims <- calculate(bum2, nsim = 100)[[1]][, , 1]
-# plot(sims[1, ] ~ dates, type = "n", ylim = range(sims))
-# for (i in 1:100) {
-#   lines(sims[i, ] ~ dates, lwd = 1)
-# }
+# behaviour-switching latent factor for back to work period. schools go back
+# from Jan 28 to Feb 3-5, so set mean to Feb 1
+back_to_school_datenum <- as.numeric(lubridate::date("2020-02-01") - first_date)
+tau_back_to_work <- normal(back_to_school_datenum, 5)
+back_to_work <- latent_behaviour_switch(date_num, tau_back_to_work)
 
-# add behaviour switching latent factor for back to work period
-# schools go back from Jan 28 to Feb 3-5, so set the mean to Feb 1
-b2s_datenum <- as.numeric(lubridate::date("2020-02-01") - first_date)
-tau_b2w <- normal(b2s_datenum, 5)
-b2w <- latent_behaviour_switch(date_num, tau_b2w)
+# spline latent factor for weekly variation and expand out to dates
+day_weights <- latent_spline(1:7)
+doy <- lubridate::wday(dates)
+weekday <- day_weights[doy]
 
-# add latent factor for weekend effect
-# add (fixed) latent factor for each weekend
[…]

### 2020-04-21 R/latent_social_distancing.R
fix up plots and outputting

@@ -107,10 +107,13 @@ min(n_effs)
 # plot latent factors
 colours <- c("Greens", "Oranges", "Purples", "Blues", "PuRd", "Reds")
 
+png("outputs/figures/latent_factors.png",
+    width = 2000, height = 2000,
+    pointsize = 40)
 par(mfrow = n2mfrow(n_latents))
 for (i in 1:n_latents) {
   plot_latent_factor(
-    factor = z[, i],
+    factor = latents[, i],
     draws = draws,
     dates = dates,
     key_dates = interventions$date,
@@ -118,20 +121,25 @@ for (i in 1:n_latents) {
     latent_names[i]
   )
 }
+dev.off()
 
 # plot datastreams and latent factor fit
 
 # simulate with error variance
 errors <- normal(0, 1, dim = dim(trends))
-latent_fit <- trends + sweep(errors, 2, sigma_obs, FUN = "+")
+latent_fit <- trends + sweep(errors, 2, sigma_obs, FUN = "*")
 sim <- calculate(latent_fit, values = draws, nsim = 1000)[[1]]
 trend_mean <- apply(sim, 2:3, mean)
 trend_lower <- apply(sim, 2:3, quantile, 0.025)
 trend_upper <- apply(sim, 2:3, quantile, 0.975)
 
 
+png("outputs/figures/datastream_model_fit.png",
+    width = 3000, height = 2500,
+    pointsize = 50)
 
-par(mfrow = n2mfrow(n_datastreams))
+par(mfrow = n2mfrow(n_datastreams),
+    mar = c(2, 2, 4, 2))
 for (i in seq_len(n_datastreams)) {
   datastream_i <- datastreams[i]
   plot_data <- aus_mobility %>%
@@ -149,11 +157,11 @@ for (i in seq_len(n_datastreams)) {
        xlim = range(dates),
        type = "n",
        ylim = ylim,
-       ylab = "trend",
+       ylab = "",
        xlab = "")
   add_gridlines(interventions$date)
   add_mean_ci(est, dates_plot,
-              col = grey(0.95),
+              col = grey(0.9),
               border_col = grey(0.8),
               line_col = grey(0.4),
               lwd = 2)
@@ -165,24 +173,26 @@ for (i in seq_len(n_datastreams)) {
   title(main = datastream_i)
   
 }
+dev.off()
 
 # plot the loadings (removing the intercept column)
 latent_loadings <- loadings[-1, ]
-loadings_draws <- calculate(latent_loadings, values = draws)
[…]

### 2020-04-21 R/functions.R
add hierarchical model on states

@@ -367,15 +367,15 @@ latent_behaviour_switch <- function(date_num,
 latent_behavioural_event <- function(date_num, tau, kappa = normal(3, 1, truncation = c(0, Inf)), lambda = 1) {
   # work with logs for numerical stability
   # e <- exp(-lag / kappa)
-  # lambda * e / kappa * (1 + e) ^ 2
+  # 4 * lambda * e / (1 + e) ^ 2
   lag <- date_num - tau
   le <- -lag / kappa
   log_bump <- log(4 * lambda) + le - 2 * log1pe(le)
   exp(log_bump)
 }
 
+# a b-spline constrained to 0-1
 latent_spline <- function (x, knots = 3) {
-  # build a spline latent factor for weekly variation constrained to 0-1
   bases <- splines::bs(x, df = knots)
   weights <- normal(0, 1, dim = knots)
   spline <- bases %*% weights

### 2020-04-21 R/latent_social_distancing.R

@@ -7,8 +7,12 @@ library(lubridate)
 library(greta)
 library(RColorBrewer)
 
-# load mobility datastreams, and potential drivers
-mobility <- all_mobility()
+# load mobility datastreams, keeping only state-level data
+mobility <- all_mobility() %>%
+  filter(!is.na(state)) %>%
+  filter(!is.na(trend)) %>%
+  arrange(state, datastream, date) %>%
+  mutate(state_datastream = str_c(state, datastream, sep = " "))
 interventions <- intervention_dates()
 holidays <- holiday_dates()
 populations <- state_populations()
@@ -17,10 +21,25 @@ populations <- state_populations()
 first_date <- min(mobility$date)
 last_date <- max(mobility$date)
 dates <- seq(first_date, last_date, by = 1)
-datastreams <- unique(mobility$datastream)
-n_datastreams <- n_distinct(mobility$datastream)
 n_dates <- n_distinct(dates)
-n_states <- n_distinct(na.omit(mobility$state))
+
+datastreams <- mobility %>%
+  group_by(datastream) %>%
+  summarise() %>%
+  pull(datastream)
+n_datastreams <- length(datastreams)
+
+states <- mobility %>%
+  group_by(state) %>%
+  summarise() %>%
+  pull(state)
+n_states <- length(states)
+
+state_datastreams <- mobility %>%
+  group_by(state_datastream) %>%
+  summarise() %>%
+  pull(state_datastream)
+n_state_datastreams <- length(state_datastreams)
 
 # social distancing latent factor as a function of behavioural switches
 # triggered by the major interventions
@@ -71,28 +90,45 @@ latent_names <- c("Social distancing",
                   "Public holidays",
                   "Week/Social distancing interaction")
 
-# project latent factors onto datastreams - adding an intercept column
-loadings <- normal(0, 10, dim = c(n_latents + 1, n_datastreams))
+# project latent factors onto state-datastreams - adding an intercept column
+
+# hierarchical prior on loadings, so states have similar values, within for each
+# latent factor
+
+# get index to state in the combined state-datastreams (not all states have all
+# datastreams, so this handles mismatch)
+col_index <- mobility %>%
+  group_by(state_datastream) %>%
+  summarise(datastream = first(datastream)) %>%
+  pull(datastream) %>%
+  match(datastreams)
+
[…]

### 2020-04-21 R/latent_social_distancing.R
remove intercept

@@ -104,15 +104,16 @@ col_index <- mobility %>%
   match(datastreams)
 
 # a mean and standard deviation for each latent factor and datastream
-means <- normal(0, 10, dim = c(n_latents + 1, n_datastreams))
-sds <- normal(0, 1, dim = c(n_latents + 1, n_datastreams),
+means <- normal(0, 10, dim = c(n_latents, n_datastreams))
+sds <- normal(0, 1, dim = c(n_latents, n_datastreams),
               truncation = c(0, Inf))
 
 # hierarchical decentring with a 3D array squished into two dimensions
-loadings_raw <- normal(0, 1, dim = c(n_latents + 1, n_state_datastreams))
+loadings_raw <- normal(0, 1, dim = c(n_latents, n_state_datastreams))
 loadings <- means[, col_index] + loadings_raw * sds[, col_index]
 
-z <- cbind(ones(n_dates), latents)
+# z <- cbind(ones(n_dates), latents)
+z <- latents
 trends <- z %*% loadings
 
 # extract expected trend for each observation and define likelihood
@@ -229,7 +230,7 @@ for (j in seq_len(n_states)) {
 
 
 # plot the national mean loadings (removing the intercept column)
-latent_loadings <- means[-1, ]
+latent_loadings <- means  # [-1, ]
 loadings_sim <- calculate(latent_loadings, values = draws, nsim = 1000)[[1]]
 loadings_mean <- apply(loadings_sim, 2:3, mean)
 loadings_lower <- apply(loadings_sim, 2:3, quantile, 0.025)

### 2020-04-22 R/functions.R
update apple and fb datastreams

@@ -5,7 +5,7 @@ library(stringr)
 # read in and tidy up Facebook movement data
 facebook_mobility <- function() {
   
-  file <- "data/fb_data/au_gadm_mobility_statistics.20200416.csv"
+  file <- "data/fb_data/au_gadm_mobility_statistics.20200420.csv"
   data <- readr::read_csv(file) %>%
     dplyr::select(
       state = polygon_name,
@@ -111,7 +111,7 @@ apple_mobility <- function() {
     "Hobart",
     "Darwin"
   )
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev10/v1/en-us/applemobilitytrends-2020-04-19.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev11/v1/en-us/applemobilitytrends-2020-04-20.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),

### 2020-04-22 R/functions.R
add waning latent; split holiday effects by state; improve colours

@@ -399,22 +399,10 @@ state_populations <- function() {
 }
 
 # summarise the posterior for a vector greta array
-summarise_vec_posterior <- function(vector, draws, sigma = NULL) {
-  
-  # if sigma isn't NULL, add that standard deviation to the CIs, and call this
-  # function recursively
-  if (!is.null(sigma)) {
-    mean_est <- summarise_vec_posterior(vector, draws)
-    lower_est <- summarise_vec_posterior(vector - 1.96 * sigma, draws)
-    upper_est <- summarise_vec_posterior(vector + 1.96 * sigma, draws)
-    est <- cbind(mean_est[, 1], lower_est[, 2], upper_est[, 3])
-    return(est)
-  }
-  
-  vector_draws <- calculate(vector, values = draws)
-  vector_mat <- as.matrix(vector_draws)
-  posterior_mean <- colMeans(vector_mat)
-  posterior_ci <- t(apply(vector_mat, 2, quantile, c(0.025, 0.975)))
+summarise_vec_posterior <- function(vector, draws, nsim = 1000) {
+  vector_sim <- calculate(vector, values = draws, nsim = nsim)[[1]][, , 1]
+  posterior_mean <- colMeans(vector_sim)
+  posterior_ci <- t(apply(vector_sim, 2, quantile, c(0.025, 0.975)))
   cbind(mean = posterior_mean, posterior_ci)
 }
 

### 2020-04-22 R/latent_social_distancing.R

@@ -47,7 +47,15 @@ date_num <- as.numeric(dates - first_date)
 trigger_date_num <- as.numeric(interventions$date - first_date)
 distancing <- latent_behaviour_switch(date_num, trigger_date_num)
 
-# (add terms for the potential waning of distancing, around an unknown date)
+# add term for recent change to social distancing (some proportion either
+# switching back to baseline behaviour or increasing distancing behaviour), at
+# some which could be from halfway since the last intervention to now, up to one
+# week in the future
+last_intervention <- max(trigger_date_num)
+tau_distancing_start <- last_intervention + (n_dates - last_intervention) / 2
+tau_distancing_change <- uniform(tau_distancing_start, n_dates + 7)
+distancing_change <- latent_behaviour_switch(date_num, tau_distancing_change)
+distancing_change <- distancing_change / distancing_change[n_dates]
 
 # latent factor for pre-distancing surge in mobility with a prior that it peaks
 # around the time of the first restriction
@@ -65,56 +73,84 @@ back_to_work <- latent_behaviour_switch(date_num, tau_back_to_work)
 day_weights <- latent_spline(1:7)
 doy <- lubridate::wday(dates)
 weekday <- day_weights[doy]
-
-# latent factor for public holidays - IID on each date
-is_a_holiday <- dates %in% holidays$date
-n_holidays <- sum(is_a_holiday)
-holiday_weights <- uniform(0, 1, dim = n_holidays)
-holiday_weights <- holiday_weights / max(holiday_weights)
-holiday <- zeros(n_dates)
-holiday[is_a_holiday] <- holiday_weights
-
+  
 # combine into latent factor matrix
-latents <- cbind(distancing,
-                 bump,
-                 back_to_work,
-                 weekday,
-                 holiday,
-                 distancing * weekday)
-n_latents <- ncol(latents)
+latents_ntnl <- cbind(bump,
+                      distancing,
+                      distancing_change,
+                      back_to_work,
+                      weekday,
+                      distancing * weekday)
+n_latents_ntnl <- ncol(latents_ntnl)
 
-latent_names <- c("Social distancing",
-                  "The Quilton bump",
+latent_names <- c("Preparation for distancing",
[…]

### 2020-04-22 R/latent_social_distancing.R
tighten prior on timing of back to work effect

@@ -66,8 +66,11 @@ bump <- latent_behavioural_event(date_num, tau_bump, kappa_bump)
 # behaviour-switching latent factor for back to work period. schools go back
 # from Jan 28 to Feb 3-5, so set mean to Feb 1
 back_to_school_datenum <- as.numeric(lubridate::date("2020-02-01") - first_date)
-tau_back_to_work <- normal(back_to_school_datenum, 5)
-back_to_work <- latent_behaviour_switch(date_num, tau_back_to_work)
+tau_back_to_work <- normal(back_to_school_datenum, 1)
+kappa_back_to_work <- normal(1, 0.25, truncation = c(0, Inf))
+back_to_work <- latent_behaviour_switch(date_num,
+                                        tau_back_to_work,
+                                        kappa = kappa_back_to_work)
 
 # spline latent factor for weekly variation and expand out to dates
 day_weights <- latent_spline(1:7)

### 2020-04-22 R/functions.R
improve colours and plotting

@@ -434,9 +434,29 @@ add_gridlines <- function(key_dates, vertical = TRUE, horizontal = TRUE) {
 }
 
 # colours for plotting
-pal <- function(name = "Greens") {
-  idx <- c(2, 5, 6, 7)
-  RColorBrewer::brewer.pal(9, name)[idx]
+pal <- function(colour = "green") {
+  
+  # look up the base colour
+  brewer_cols <- brewer.pal(8, "Set2")
+  # display.brewer.pal(8, "Set2")
+  colour_names <- c(
+    "green",
+    "red",
+    "blue",
+    "pink",
+    "bright green",
+    "yellow",
+    "beige",
+    "grey"
+  )
+  idx <- match(colour, colour_names)
+  base_colour <- brewer_cols[idx]
+  
+  # create a four-colour palette based on this colour
+  pal <- colorRampPalette(c("white", base_colour, "black"))(10)
+  pal[c(4, 5, 6, 7)]
+  
+  
 }
 
 plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0.9, 0.7, 0.5, 0.3)), title = "") {
@@ -455,5 +475,5 @@ plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0
               border_col = cols[2],
               line_col = cols[3])
   title(main = title,
-        col.main = cols[4])
+        col.main = grey(0.3))
 }

### 2020-04-22 R/latent_social_distancing.R

@@ -180,14 +180,14 @@ min(n_effs)
 # ~~~~~~~~~
 # plot fits
 
-# plot (nationwide) latent factors
-colours <- c("Purples", "Greens", "Reds", "Blues", "PuRd", "Oranges")
+# plot the first 4 (nationwide) latent factors
 
+colours <- c("pink", "green", "red", "blue")
 png("outputs/figures/latent_factors.png",
-    width = 2000, height = 2000,
-    pointsize = 40)
-par(mfrow = n2mfrow(n_latents_ntnl))
-for (i in 1:n_latents_ntnl) {
+    width = 2000, height = 1500,
+    pointsize = 50)
+par(mfrow = c(2, 2))
+for (i in 1:4) {
   plot_latent_factor(
     factor = latents_ntnl[, i],
     draws = draws,
@@ -267,8 +267,8 @@ for (j in seq_len(n_states)) {
 }
 
 
-# plot the national mean loadings (removing the intercept column)
-latent_loadings <- means_ntnl  # [-1, ]
+# plot the national mean loadings for the first four latent factors
+latent_loadings <- means_ntnl[1:4, ]
 loadings_sim <- calculate(latent_loadings, values = draws, nsim = 1000)[[1]]
 loadings_mean <- apply(loadings_sim, 2:3, mean)
 loadings_lower <- apply(loadings_sim, 2:3, quantile, 0.025)
@@ -324,10 +324,11 @@ loadings_plot_data %>%
 
 ggsave("outputs/figures/loadings_datastream.png",
        width = 10,
-       height = 5.5)
+       height = 4.5)
 
-# - use nicer pastel colours for plots
-# - use different colours for loadings (save red/green for waning traffic light) 
-# - plot state-by-factor traffic light plots
-# - plot state traffic light plot for waning social distancing
+# - plot state-by-factor loading plots
+# - add legend to loading plots
+# - plot state-datastream traffic light plot for waning social distancing
+# - pull more model code out into functions
 # - programatically find Apple download link
+# - programatically find Google download link

### 2020-04-22 R/functions.R
add social distancing regression early warning plot

@@ -399,10 +399,13 @@ state_populations <- function() {
 }
 
 # summarise the posterior for a vector greta array
-summarise_vec_posterior <- function(vector, draws, nsim = 1000) {
+summarise_vec_posterior <- function(vector,
+                                    draws,
+                                    quantiles = c(0.025, 0.975),
+                                    nsim = 1000) {
   vector_sim <- calculate(vector, values = draws, nsim = nsim)[[1]][, , 1]
   posterior_mean <- colMeans(vector_sim)
-  posterior_ci <- t(apply(vector_sim, 2, quantile, c(0.025, 0.975)))
+  posterior_ci <- t(apply(vector_sim, 2, quantile, quantiles))
   cbind(mean = posterior_mean, posterior_ci)
 }
 

### 2020-04-26 R/functions.R
try using a November baseline

@@ -23,8 +23,9 @@ prep_nbn_mobility <- function() {
     mutate(GB = bytes * 1e-9) %>%
     select(-bytes)
 
-  # compute baseline values for each day of the week Jan 3 - Feb 6 (same as Google)
-  baseline_period <- as.Date(c("2020-01-03", "2020-02-06"))
+  # compute baseline values for each day of the week in Novembr 2019
+  # similarly a non-holiday period, 
+  baseline_period <- as.Date(c("2019-11-01", "2019-11-30"))
   baseline <- data %>%
     mutate(in_baseline = date >= baseline_period[1] &
              date <= baseline_period[2]) %>%
@@ -35,7 +36,7 @@ prep_nbn_mobility <- function() {
   data <- data %>%
     left_join(baseline) %>%
     mutate(trend = 100 * (GB - baseline)/ abs(baseline)) %>%
-    filter(date > baseline_period[1])
+    filter(date > as.Date("2020-02-01"))
   
   # rename states, drop unnecessary columns, and save to disk
   data %>%

### 2020-04-26 R/functions.R
use composite January and pre-Christmas baseline for NBN data

@@ -23,20 +23,48 @@ prep_nbn_mobility <- function() {
     mutate(GB = bytes * 1e-9) %>%
     select(-bytes)
 
-  # compute baseline values for each day of the week in Novembr 2019
-  # similarly a non-holiday period, 
-  baseline_period <- as.Date(c("2019-11-01", "2019-11-30"))
-  baseline <- data %>%
-    mutate(in_baseline = date >= baseline_period[1] &
-             date <= baseline_period[2]) %>%
+  # subset to fully observed days, since there's a trend within the 10-3 period,
+  # so curtailing it early would bias the metric
+  last_reliable_day <- df %>%
+    filter(!is.na(date)) %>%
+    group_by(date) %>%
+    summarise(latest_hour = max(hour, na.rm = TRUE)) %>%
+    filter(latest_hour > 15) %>%
+    pull(date) %>%
+    max()
+    
+  data <- data %>%
+    filter(date <= last_reliable_day)
+  
+  # compute baseline values based on the daily pattern in November 2019
+  # (similarly a non-holiday period), but the man usage from Jan 3 - Feb 6 (the
+  # same as Google)
+  nov_baseline_period <- as.Date(c("2019-11-01", "2019-11-30"))
+  jan_baseline_period <- as.Date(c("2020-01-03", "2020-02-06"))
+  
+  # relevel against November baseline first to remove day effect
+  nov_baseline <- data %>%
+    mutate(in_baseline = date >= nov_baseline_period[1] &
+             date <= nov_baseline_period[2]) %>%
     group_by(direction, state, weekday) %>%
-    summarise(baseline = median(GB[in_baseline]))
+    summarise(nov_baseline = median(GB[in_baseline]))
+  
+  data <- data %>%
+    left_join(nov_baseline) %>%
+    mutate(GB_diff = GB - nov_baseline)
+  
+  # now relevel differences against Jan baseline, to match google
+  jan_baseline <- data %>%
+    mutate(in_baseline = date >= jan_baseline_period[1] &
+             date <= jan_baseline_period[2]) %>%
+    group_by(direction, state) %>%
+    summarise(jan_baseline = median(GB_diff[in_baseline]))
   
   # compute percentage change, and remove pre-baseline data
   data <- data %>%
-    left_join(baseline) %>%
-    mutate(trend = 100 * (GB - baseline)/ abs(baseline)) %>%
-    filter(date > as.Date("2020-02-01"))
[…]

### 2020-04-27 R/functions.R
use daily baseline for facebook staying still datastream

@@ -18,13 +18,14 @@ facebook_mobility <- function() {
       names_to = "metric",
       values_to = "trend"
     ) %>%
-    mutate(date = lubridate::date(date))
+    mutate(date = lubridate::date(date)) %>%
+    mutate(weekday = lubridate::wday(date))  
   
   # set the staying home variable against a baseline of the first two weeks
   baseline <- data %>%
     filter(date < lubridate::date("2020-03-15")) %>%
-    group_by(state, metric) %>%
-    summarise(baseline = mean(trend))
+    group_by(state, metric, weekday) %>%
+    summarise(baseline = median(trend))
   
   data <- data %>%
     left_join(baseline) %>%
@@ -37,6 +38,7 @@ facebook_mobility <- function() {
     select(
       -corrected,
       -baseline,
+      -weekday
       ) %>%
     mutate(
       trend = trend * 100,

### 2020-04-27 R/functions.R
tweak latent factor plot label

@@ -514,7 +514,6 @@ pal <- function(colour = "green") {
   pal <- colorRampPalette(c("white", base_colour, "black"))(10)
   pal[c(4, 5, 6, 7)]
   
-  
 }
 
 plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0.9, 0.7, 0.5, 0.3)), title = "") {
@@ -523,7 +522,7 @@ plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0
   plot(est[, 1] ~ dates,
        type = "n",
        ylim = c(0, 1),
-       ylab = "effect",
+       ylab = "relative effect",
        xlab = "",
        las = 1)
   add_gridlines(key_dates, horizontal = FALSE)

### 2020-04-28 manuscript/ms.Rmd
add NBN data section to mt_nbn branch

@@
 
+<!-- the shifts can be a sum, the bumps should be singular -->
+
 #### Observation models
@@
 where $\mathbf{x}$ is an $N$-by-$M+1$ design matrix of covariates, including a column of ones to represent the intercept terms, $\beta$ is an $M-1$-by-$N$ matrix of regression coefficients, $\mathbf{z}$ is a $T$-by-$K$ matrix of behavioural latent factors, $\omega$ is a $K$-by-$N$ matrix of loadings between the latent factors and data streams, and $\epsilon$ is a $T$-by-$N$ matrix of independent Gaussian errors.
+
+### Data
+
+#### NBN
+
+The National Broadband Network Co.
+(NBN) is the Australian national wholesale open-access data network, providing broadband access to retail service providers.
+The majority of its traffic is through wired broadband connections.
+As such, these connections are static in place, and they offer internet access to a broad mixture of users.
+NBN provided a time series (with a temporal resolution of 15 minutes) to aggregated data relating to mean upload and download traffic volume generated by houshold customers, aggregated spatiall by ABS Statistical Area 2 (SA2) region, and excluding data points generated by less than 600 customers.
 

### 2020-04-28 manuscript/ms.Rmd
Revert "Revert "adding NBN data section""
This reverts commit 366dca4a323aec58fb9c313b5e039fa604643ad9.
Revert "Revert "Revert "adding NBN data section"""
This reverts commit b22ee3fe8d40380b0dde0e335b7852e5be51f1d9.
adding NBN data section
adding NBN data section - updated
adding timeline draft

@@
 
+Full-day NBN data for the period from September 1, 2019, until March 24th, 2020 have been used.
+
+#### Announcement timeline for Australia
+
+Govenrmental controls in response to COVID-19 in Australia started in January 2020.
+On January 20th, 2020, the Australian government imposed restrictions on flights from the Wuhan region of China.
+The first case was identified in Australia shortly after, on January 25th.
+Queensland declared *State of emergency* on January 29th, and travel restrictionsfrom China were imposed on Fenruary 1st.
+Enforced measures were imposed from March 16th, with a ban on non-essential gatherings, with Victoria declaring a *State of emergency* on the same day, and closing schools from March 22nd.
+From the 22th of March, Stage 1 restricitons were mandated (closures of pubs, restaurants, except takeaway, and adivsing on social distancing).
+This was escalated to Stage 2 from March 26th, with mandatory social distancing and closures of museums, galleries, many additional services.
+On March 31st, Stage 3 restrictions were imposed, with enforcdeable stay at home measures.
+
  - Data availability by source

### 2020-04-28 manuscript/ms.Rmd
adding timeline draft

@@
 
-Govenrmental controls in response to COVID-19 in Australia started in January 2020.
+Governmental controls in response to COVID-19 in Australia started in January 2020.
 On January 20th, 2020, the Australian government imposed restrictions on flights from the Wuhan region of China.
 The first case was identified in Australia shortly after, on January 25th.
-Queensland declared *State of emergency* on January 29th, and travel restrictionsfrom China were imposed on Fenruary 1st.
+Queensland declared a *State of emergency* on January 29th, and travel restrictions from China were imposed on February 1st.
 Enforced measures were imposed from March 16th, with a ban on non-essential gatherings, with Victoria declaring a *State of emergency* on the same day, and closing schools from March 22nd.
 From the 22th of March, Stage 1 restricitons were mandated (closures of pubs, restaurants, except takeaway, and adivsing on social distancing).
-This was escalated to Stage 2 from March 26th, with mandatory social distancing and closures of museums, galleries, many additional services.
+This was escalated to Stage 2 from March 26th, with mandatory social distancing and closures of museums, galleries, and additional services.
 On March 31st, Stage 3 restrictions were imposed, with enforcdeable stay at home measures.

### 2020-04-28 R/functions.R
update data URLs

@@ -93,7 +93,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev14/v1/en-us/applemobilitytrends-2020-04-23.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev18/v1/en-us/applemobilitytrends-2020-04-26.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -133,7 +133,7 @@ apple_mobility <- function() {
 citymapper_mobility <- function() {
   
   # get link from: https://citymapper.com/cmi/about
-  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200424.csv"
+  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200427.csv"
   data <- readr::read_csv(url, skip = 3) %>%
     tidyr::pivot_longer(cols = -Date,
                         names_to = "region",

### 2020-04-28 manuscript/ms.Rmd
dump report methods text into manuscript

--- 2020-04-28 manuscript/ms.Rmd
@@
 
-We apply this model to 9 separate data streams provided by from two technology companies for Australia, with spatial data varying between national, state and city-levels.
+We apply this model to 12 separate data streams provided by 4 technology companies for Australia, with spatial data varying between national, state and city-levels.
 This reveals two distinct latent population behavioural factors: a strong social distancing effect in response to national announcements and policies, preceeded by a small and short-lived rise in mixing apparently driven by the population 'stocking up' on supplies.
@@
  - relative effect size by state
+ - impact of the latent factors on R_eff
 
@@
 
+*NB. this is currently a lot of copy-pasta from reports etc. which will be massaged into a report*
+
+A number of datastreams provide information on mobility before and in response to COVID-19 across the states and territories.
+Each of these datastreams represents a different aspect of population mobility, but they show some common trends - reflecting underlying changes in behaviour.
+We use a latent factor statistical model to simultaneously analyse these datastreams and quantify these underlying behavioural factors.
+
+### Mobility datastreams
+
+We currently consider 12 different datastreams, provided by four different technology companies: Apple and Citymapper provide regularly-updated data on direction requests, whilst Facebook and Google provide less regularly-updated data on different measures of mobility from users' GPS data.
+Each datastream is encoded as a percentage change in the mobility metric, relative to a pre-COVID-19 baseline.
+
+#### Apple
+
+Apple provide 3 datastreams from their Apple Maps app of the total number of user requests for directions, one each for driving, walking, and public transport.
+No further details are provided on the nature of these journeys.
+Separate versions of these datastreams are provided for Sydney, Melbourne, Brisbane, Perth, and for all of Australia.
+We use the city-level data as representative of mobility within that state, and do not currently use the national composite data.
+Apple's proportional change data is relative to a baseline of the metric in each location on January 13th 2020.
+Daily counts are calculated using Pacific Standard Time (i.e.
+the time in California), which largely but not entirely overlaps with the subsequent day in Australia.
+We therefore assign mobility data to the subsequent date.
+These data are updated daily and can be accessed at: https://www.apple.com/covid19/mobility.
+
+#### Citymapper
+
+Citymapper provide a composite index based on the numbers of requests for directions using the Citymapper app.
+This is a composite metric of requests for different transport mode, though Citymapper does not provide driving directions and is primarily used for public transport directions, with some use for walking, cycling, and cab directions.
+No further details are provided on the nature of these journeys.
+Separate versions of these datastreams are provided for Sydney, Melbourne, and for all of Australia.
+We use the city-level data as representative of mobility within that state, and do not currently use the national composite data.
+Citymapper's proportional change data are relative to a baseline of the metric between January 6th and February 2nd, 2020.
+Daily counts are calculated using UTC (i.e.
+the time in London), which largely but not entirely overlaps with the same day in Australia.
+We therefore assign mobility data to the date provided.
+These data are updated daily and can be accessed at: https://citymapper.com/CMI.
+
+#### Google
+
+Google provide GPS-derived indices of the amount of time spent (a combination of visits and lengths of stay) in locations of one of 6 types. The following are Google's descriptions of these places:
+- retail and recreation: “places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters”
+- grocery & pharmacy: “places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies”
+- parks: “places like national parks, public beaches, marinas, dog parks, plazas, and public gardens”
+- transit stations: “places like public transport hubs such as subway, bus, and train stations”
+- workplaces: “places of work”
+- residential: “places of residence”
+This visitation information is derived from the aggregated GPS tracks of users with the Location History setting enabled (off by default).
+No further details are provided on the users or the nature of these visits.
+Separate versions of these datastreams are provided for each state and territory, and for all of Australia.
+We do not currently use the national composite data.
+Google's proportional change data are relative to a baseline of the metric between January 3rd and February 6th, 2020.
+Daily counts are calculated using UTC (i.e.
+the time in London), which largely but not entirely overlaps with the same day in Australia.
+We therefore assign mobility data to the date provided.
+These data are updated intermittently.
+Recently the data have been released every week, but providing data only up to about a week before the release date so that data are typically between 1 and 2 weeks out of date.
+The data can be accessed at: https://www.google.com/covid19/mobility/.
+
+#### Facebook
+
+Facebook provide two GPS-derived indices on user movement: how far users travel per day, and the proportion staying in the same location.
+These indices are calculated from GPS location data that is first aggregated to grid cells with dimensions 600x600 metres, to protect user privacy.
+Users remaining in these grid cells throughout the day are considered to have ‘stayed still'.
+For users who have travelled outside that cell, the movement range is the number of unique cells in which the user was present on that date.
+Only Facebook users using the Facebook smartphone app and who have opted in to their GPS locations being recorded are represented in these data.
+No further details are provided on these users or the nature of their journeys.
+Separate versions of these datastreams are provided for each state.
+The movement range datastream is provided by Facebook as a percentage change, relative to a baseline of average values of the metric throughout February 2020.
+The proportion of users staying still is provided as an absolute value, and we convert it into a percentage change from a baseline of the median daily value between March 1st and 15th.
+Daily metrics are calculated using UTC (i.e.
+the time in London), which largely but not entirely overlaps with the same day in Australia.
+The data are described here: https://dataforgood.fb.com/tools/disease-prevention-maps/ and researchers (but not governments) can request access.
+
+#### Representativeness
+
+All of these datastreams are derived from apps that are primarily smartphone-based.
+As such they overrepresent demographics that are more likely to own smartphones and have account with these technology companies.
+Data on the demographics of the users that contributed to these particular datastreams are not available.
+However rough public estimates suggest that 90% of Facebook users are aged between 18 and 64.
+Younger and older demographics are underrepresented, with 4.5% facebook users vs 16% in the general population over the age of 65, and 5.6% vs 20% under the age of 18.
+Given that school-age children are underrepresented in COVID-19 cases in Australia, and that those over 65 are likely to be less mobile, these data may be unintentionally biased towards those contributing most strongly to transmission.
+Apple's and Citymapper's direction requests likely represent deliberate visits to a specific location, and a location that the user either does not usually visit (and therefore requires directions) or would benefit from live traffic updates for.
+Whilst Google's index of the time spent in residential places is likely to be a good proxy for people staying at home, it does not exclude the possibility that this time is spent in another person's household.
+
+#### Access and privacy
+
+The datastreams provided by Apple, Citymapper and Google are all publicly available for the express purpose of supporting public heath bodies in their response to COVID-19.
+The datastreams provided by Facebook are provided to researchers via Facebook's Data for Good  programme.
+All of these datasets are fully anonymised and aggregated at the level of either states and territories or Australia's 4 largest cities, over each day.
+The large-scale aggregation of these datasets ensures the privacy of users - the smallest population is that of the Northern Territory: over 245,000.
+
+### Timeline of Physical Distancing Measures
+
+ - Announcement timeline for Australia
+
 ### General model description
 
+We simultaneously analyse all of these datastreams using a statistical latent factor model.
+The timeseries of values for each datastream is regressed against a set of ‘latent factors'.
+All datastreams share the same latent factors, but have their own regression coefficients.
+Rather than using known values for these common latent factors (as would be the case in a regression analysis), each of the latent factors is itself a function of some unknown parameters that we infer during model fitting.
+This enables us to statistically infer a set of common temporal patterns that are shared - each to a different extent - by all of the datastreams.
+
+We use prior knowledge to define the different latent functions, so that they represent known or hypothesised effects.
+Two of these latent factors are related to COVID-19, and three describe non-COVID-19-related trends.
+The three COVID-19 related latent factors considered are: ‘preparation' - a surge of behaviours is preparation of social distancing, ‘social distancing' - a switch to social distancing behaviour, and ‘waning distancing' - a partial reversion to a non-social-distancing behavioural state.
+The three non-COVID-19-related latent factors are: ‘back to work' - a behavioural switch associated with the start of school term one (and return to work for many parents), ‘day of the week' - different behaviours on each day of the week, including weekday vs weekend behaviours, and ‘public holidays' - different behaviours associated with public holidays.
+The model fits for each datastream incorporate the impacts of all of these latent factors, though we are most interested in the COVID-19 related factors.
+
+Because each datastream weights the different latent factors differently, these latent factors are considered only as relative effects - they have no absolute magnitude or sign.
+We therefore define each of them as functions with values constrained between 0 and 1.
+The functions reflect prior knowledge of the behaviours they represent:
+
+Preparation
+A smooth, symmetric unimodal peak representing behaviours in advance of anticipated social distancing behaviour. The date and width of the peak are estimated from the data.
+
+Social Distancing
+A smooth monotone increasing function representing the population switching to a behavioural state of increased social distance.
+We assume that the switching happens in bursts around each of three major non-pharmaceutical interventions (NPIs) introduced at the national level: restriction of gatherings to 500 or fewer people on March 16th; closure of restaurants, bars, and cafes on March 24th; restriction of gatherings to 2 or fewer people in most states and territories on March 29th.
+The relative magnitudes of the behavioural switches associated with each of these NPIS, as well as the rate of change around each, are estimated from the data.
+
+Waning Distancing
+A smooth monotone increasing function representing a waning of social distancing behaviour as the population switches back to a socially non-distant behavioural state.
+Currently the parameters of this latent factor are fixed - implying a steady increase starting two weeks before the most recent date.
+The relative impact of (and evidence for) this behaviour on each datastream is estimated from the data.
+As more up-to-date data become available for the Google and Facebook datastreams the timing and slope of this effect will be estimated from the data.
+
+Back to Work
+An increasing sigmoid function representing a switch from one behavioural state during the school holidays to another during term one.
+The location (date by which 50% of people have switched) and slope of this sigmoid are estimated from the data.
+
+Day of the Week
+A flexible curve over the day of the week, representing a gradient of behaviour between the peak of activity in the work week, and the trough at the weekend.
+The curve is constrained between 0 and 1, with a value of 1 constrained to fall on a Sunday.
+The three parameters describing the shape of the curve are estimated from the data.
+
+Public Holidays
+A series of independent offsets applied to each public holiday in each state to represent different behavioural states on these days.
+The value offset for each holiday is estimated from the data, and scaled to between 0 and 1.
+
 #### Latent factor model
+
+*Note: this will be overhauled shortly, since in the current version of the model, everything is now a latent factor.*
 
@@
 
-### Data
+### Latent factor mobility model for Australian states and Territories
 
-#### NBN
-
-The National Broadband Network Co.
-(NBN) is the Australian national wholesale open-access data network, providing broadband access to retail service providers.
-The majority of its traffic is through wired broadband connections.
-As such, these connections are static in place, and they offer internet access to a broad mixture of users.
-NBN provided a time series (with a temporal resolution of 15 minutes) to aggregated data relating to mean upload and download traffic volume generated by houshold customers, aggregated spatiall by ABS Statistical Area 2 (SA2) region, and excluding data points generated by less than 600 customers.
-
- - Data availability by source
- - Announcement timeline for Australia
  - Specific model fitting for Australia
     - interaction between distancing and weekends
+    - greta HMC
+    - residuals checking
+
+### Estimating changes to the effective reproduction number due to physical distancing
+
+ - model to estimate timeseries of R_eff with the distancing factor as a covariate
 
@@
 -->
-
-    - greta HMC
-    - residuals checking
 

### 2020-04-28 manuscript/ms.Rmd

@@
 
+#### NBN
+
+The National Broadband Network Co.
+(NBN) is the Australian national wholesale open-access data network, providing broadband access to retail service providers.
+The majority of its traffic is through wired broadband connections.
+As such, these connections are static in place, and they offer internet access to a broad mixture of users.
+NBN provided a time series (with a temporal resolution of 15 minutes) to aggregated data relating to mean upload and download traffic volume generated by houshold customers, aggregated spatiall by ABS Statistical Area 2 (SA2) region, and excluding data points generated by less than 600 customers.
+
+Full-day NBN data for the period from September 1, 2019, until March 24th, 2020 have been used.
+
 #### Representativeness
@@
 
- - Announcement timeline for Australia
+Governmental controls in response to COVID-19 in Australia started in January 2020.
+On January 20th, 2020, the Australian government imposed restrictions on flights from the Wuhan region of China.
+The first case was identified in Australia shortly after, on January 25th.
+Queensland declared a *State of emergency* on January 29th, and travel restrictions from China were imposed on February 1st.
+Enforced measures were imposed from March 16th, with a ban on non-essential gatherings, with Victoria declaring a *State of emergency* on the same day, and closing schools from March 22nd.
+From the 22th of March, Stage 1 restricitons were mandated (closures of pubs, restaurants, except takeaway, and adivsing on social distancing).
+This was escalated to Stage 2 from March 26th, with mandatory social distancing and closures of museums, galleries, and additional services.
+On March 31st, Stage 3 restrictions were imposed, with enforcdeable stay at home measures.
 

### 2020-04-28 R/R_effective.R
add script and functions for modelling timeseries of R_effective

@@ -1,0 +1,159 @@
+# fit a Bayesian model-based estimate of R_effective over time.
+
+# The model assumes that the number of new locally-acquired cases at time $t$ in
+# region $i$ is *conditionally* Poisson-distributed with expectation given by
+# the number of infectious people $lambda_{t,i}$ multiplied by the region- and
+# time-varying reproduction rate $R_{t,i}$. To account for clusters in numbers
+# of cases reported (overdispersion in the data), and fluctuations due to
+# unobserved factors affecting transmission (temporal correlation) $R_{t,i}$ has
+# a random error structure - either temporally-correlated, temporally
+# independent, or some mix of both. This can be expressed via a Gaussian process
+# over time in each region on $log(R_{t,i})$. To account for temporal
+# fluctations common to all regions, we also include a single temporal Gaussian
+# process on $log(\hat{R}_{t})$, where $\hat{R}_{t}$ is the Australia-wide mean
+# time-varying effective reproduction rate. Fixed effects for all regions can
+# also be included via a linear mean function $m_{t,i}$ for these
+# region-specific Gaussian processes. The number of infectious people is
+# computed by disaggregating the number of cases (both locally-and
+# overseas-acquired) over a subsequent period of time, with probability
+# distribution given by a serial interval distribution. The serial interval
+# distribution is computed from a linelist of cases, and is represented by a
+# discretized cumulative density function $p-serial-interval(t')$, giving the
+# probability of a serial interval (average time between subsequent cases) of
+# $t'$. The relative contribution of imported versus locally-acquired cases to
+# subsequent infections in each region can be accounted for via a time-varying
+# contribution rate $import-contribution_{t'}$, which reflects the effectiveness
+# of quarantine measures.
+
+#   local-incidence_{t,i} ~ Poisson(lambda_{t,i} * R_{t,i})
+#   log(R_{t,i}) ~ GP(m_{t,i}, K) + log(\hat{R}_{t})
+#   log(\hat{R}_{t}) ~ GP(0, K)
+#   m_{t,i} = a + b * social-distancing-index_{t,i}
+#   lambda_{t,i} = sum_{t'=1}^t p-serial-interval(t') (local-incidence_{t',i} + imported-incidence_{t',i} import-contribution_{t'})
+
+# - load in Freya's data
+# - get serial interval distribution
[…]

### 2020-04-28 R/functions.R

@@ -534,3 +534,199 @@ plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0
   title(main = title,
         col.main = grey(0.3))
 }
+
+# greta function for the lognormal CDF (equivalent to plnorm(x, meanlog, sdlog))
+# x must not be a greta array, though meanlog and sdlog can be. A greta array of
+# CDF values is returned, equal to 0 wherex was 0 or lower.
+lognormal_cdf <- function(x, meanlog, sdlog) {
+  
+  # filter out any invalid dates, to replace their CDF with 0
+  valid <- x > 0
+  x <- x[valid]
+  
+  p <- iprobit((log(x) - meanlog) / sdlog)
+  
+  # if any were missing, replace their cumulative density with 0
+  if (any(!valid)) {
+    result <- zeros(length(valid))
+    result[valid] <- p
+    p <- result
+  }
+  
+  p
+  
+}
+
+exponential_cdf <- function (x, rate) {
+  
+  # filter out any invalid dates, to replace their CDF with 0
+  valid <- x > 0
+  x <- x[valid]
+  
+  p <- 1 - exp(-rate * x)
+  
+  # if any were missing, replace their cumulative density with 0
+  if (any(!valid)) {
+    result <- zeros(length(valid))
+    result[valid] <- p
+    p <- result
+  }
+  
+  p
+  
+}
+
+# Given samples 'x' of a parameter, approximate the distribution by a (possibly
+# truncated) normal distribution, and return a variable greta array following
+# that distribution
+parameter <- function(x, truncation = c(-Inf, Inf)) {
+  normal(mean(x), sd(x), truncation = truncation)
+}
+
+# given a positive integer vector of (days since a case became symptomatic,
+# compute the probability that the serial interval values falls in that day
+serial_interval_probability <- function(days) {
+  
+  # priors for the parameters of the lognormal distribution over the serial interval from Nishiura et
+  # al., as stored in the EpiNow source code 
+  si_param_samples <- read_csv(
+    file = "https://raw.githubusercontent.com/epiforecasts/EpiNow/758b706599244a545d6b07f7be4c10ffe6c8cf50/data-raw/nishiura-lognormal-truncated.csv",
+    col_types = cols(param1 = col_double(),
+                     param2 = col_double())
+  )
+  
+  # cdf(x) under a lognormal distribution is just pnorm() of (log(x) - mu) / sd ?
+  
[…]

### 2020-04-28 R/R_effective.R
make r_eff code a bit more efficient

@@ -112,9 +112,11 @@ local_cases <- date_by_state %>%
 
 library(greta.gp)
 
+idx <- which(states == "VIC")
+
 # summarise nationally to start with
-local_cases <- as_data(rowSums(local_cases))
-imported_cases <- as_data(rowSums(imported_cases))
+local_cases <- local_cases[, idx, drop = FALSE]
+imported_cases <- imported_cases[, idx, drop = FALSE]
 
 # the relative contribution of imported cases (relative to locally-acquired
 # cases) to transmission. I.e. one minus the effectiveness of quarantine
@@ -125,12 +127,13 @@ import_contribution <- 1
 # the expected number of infectious people in each state and time
 n_dates <- length(dates)
 
-serial_interval_disaggregation <- disaggregation_matrix(seq_len(n_dates),
-                                                        serial_interval_probability)
+serial_interval_disaggregation <- disaggregation_matrix(n_dates,
+                                                        serial_interval_probability,
+                                                        max_days = 46)
 
-imported_infectious <- serial_interval_disaggregation %*% imported_cases
-local_infectious <- serial_interval_disaggregation %*% local_cases
-lambda <- local_infectious + imported_infectious * import_contribution
+case_contribution <- local_cases + imported_cases * import_contribution
+
+lambda <- serial_interval_disaggregation %*% case_contribution
 
 # define a GP on dates, using subset of regressors approximation
 lengthscale <- lognormal(4, 0.5)
@@ -157,3 +160,25 @@ m <- model(log_R_eff)
 
 draws <- mcmc(m, one_by_one = TRUE)
 
+r_hats <- coda::gelman.diag(draws, autoburnin = FALSE, multivariate = FALSE)$psrf[, 1]
+n_eff <- coda::effectiveSize(draws)
+max(r_hats)
+min(n_eff)
+
+R_eff <- exp(log_R_eff)
+
+R_eff_draws <- calculate(R_eff, values = draws)
+R_eff_mat <- as.matrix(R_eff_draws)
+mean <- colMeans(R_eff_mat)
+ci <- apply(R_eff_mat, 2, quantile, c(0.025, 0.975))
+
+plot(mean ~ dates,
+     type = "n",
+     ylim = range(c(ci, 0, 1)))
+polygon(x = c(dates, rev(dates)),
+        y = c(ci[1, ], rev(ci[2, ])),
+        lty = 0,
+        col = "lightskyblue1")
+lines(mean ~ dates,
+      col = "blue",
+      lwd = 2)

### 2020-04-28 R/functions.R

@@ -704,29 +704,32 @@ fake_linelist <- function() {
   
 }
 
-# given a positive integer vector of day differences and a discrete probability
-# function that takes as input day differences and returns a vector of
-# probabilities that a day should actually be assigned that number of days into
-# the future, return a symmetric matrix that can be used in a matrix multiply to
+# given a positive integer 'n_days' of the number of days for which to compute
+# values, a discrete probability function 'probability_function' that takes as
+# input day differences and returns a vector of probabilities that a day should
+# actually be assigned that number of days into the future, and an optional
+# upper limit of days 'max_days' beyond which the probability is assumed to be 0
+# return a symmetric matrix that can be used in a matrix multiply to
 # disaggregate daily count data into the future according to that probability
 # distribution. Doing this disaggregation using a circulant matrix of
 # probabilities with masked lower values, is more efficient in greta than
 # looping since it can easily be parallelised. Note this is the same operation
 # as cases_known_outcome_matrix() in goldingn/australia_covid_ascertainment
-disaggregation_matrix <- function (day_differences, probability_function) {
+disaggregation_matrix <- function (n_days, probability_function, max_days = Inf) {
   
-  n_days <- length(day_differences)
+  # get a range of days ahead to consider
+  days <- seq_len(min(n_days, max_days))
   
-  # get a probability of delaying each of these number of days (starting from 0)
-  disaggregation_probs <- probability_function(day_differences - 1)
+  # get a probability of delaying by each of these number of days (starting from 0)
+  disaggregation_probs <- probability_function(days - 1)
   
   # build an upper-triangular circulant matrix (contribution of each day's cases
   # to each other's)
   mat <- zeros(n_days, n_days)
-  indices <- col(mat) - row(mat) + 1
-  mask <- indices > 0
+  indices <- row(mat) - col(mat) + 1
+  mask <- indices > 0 & indices <= max_days
   mat[mask] <- disaggregation_probs[indices[mask]]
-  t(mat)
+  mat
   
 }
 

### 2020-04-29 R/R_effective.R
implement hierarchical GP multistate model

@@ -26,17 +26,13 @@
 # of quarantine measures.
 
 #   local-incidence_{t,i} ~ Poisson(lambda_{t,i} * R_{t,i})
-#   log(R_{t,i}) ~ GP(m_{t,i}, K) + log(\hat{R}_{t})
-#   log(\hat{R}_{t}) ~ GP(0, K)
-#   m_{t,i} = a + b * social-distancing-index_{t,i}
+#   log(R_{t,i}) = R_0 + b * social-distancing-index_{t,i} + \epsilon_{t,i} + \gamma_t
+#   epsilon_i ~ GP(0, K1)
+#   gamma ~ GP(0, K2)
 #   lambda_{t,i} = sum_{t'=1}^t p-serial-interval(t') (local-incidence_{t',i} + imported-incidence_{t',i} import-contribution_{t'})
+#   log(R_0) ~ N(0.723, 0.465)
 
-# - load in Freya's data
-# - get serial interval distribution
-# - get case count time series by locals and imports
-# - write out model for one region
-# - compare fit with David's outputs
-# - extend to multiple regions.
+# this lognormal prior on R0 has a mean of 2.6 and sd of 2
 
 library(dplyr)
 library(readr)
@@ -45,6 +41,7 @@ source("R/functions.R")
 
 # return a fake linelist based on the real case counts and samples of the
 # reporting delay distribution
+set.seed(2020-04-29)
 linelist <- fake_linelist()
 
 # for now just do single imputation with the mean delay on the date of onset for
@@ -83,6 +80,10 @@ import_statuses <- sort(unique(linelist$import_status))
 states <- sort(unique(linelist$state))
 dates <- seq(min(linelist$date), max(linelist$date), by = 1)
 
+n_states <- length(states)
+n_dates <- length(dates)
+date_nums <- seq_len(n_dates)
+
 # pad this with full set of dates, states, and import statuses
 grid <- expand_grid(
   date = dates,
@@ -112,11 +113,14 @@ local_cases <- date_by_state %>%
 
 library(greta.gp)
 
-idx <- which(states == "VIC")
+# idx <- which(states == "VIC")
+# 
+# # summarise nationally to start with
+# local_cases <- local_cases[, idx, drop = FALSE]
+# imported_cases <- imported_cases[, idx, drop = FALSE]
 
-# summarise nationally to start with
-local_cases <- local_cases[, idx, drop = FALSE]
-imported_cases <- imported_cases[, idx, drop = FALSE]
+# prior on R0
[…]

### 2020-04-29 R/functions.R

@@ -585,7 +585,7 @@ parameter <- function(x, truncation = c(-Inf, Inf)) {
 
 # given a positive integer vector of (days since a case became symptomatic,
 # compute the probability that the serial interval values falls in that day
-serial_interval_probability <- function(days) {
+serial_interval_probability <- function(days, fixed = FALSE) {
   
   # priors for the parameters of the lognormal distribution over the serial interval from Nishiura et
   # al., as stored in the EpiNow source code 
@@ -595,22 +595,30 @@ serial_interval_probability <- function(days) {
                      param2 = col_double())
   )
   
-  # cdf(x) under a lognormal distribution is just pnorm() of (log(x) - mu) / sd ?
-  
-  # create greta arrays for the parameters, with these priors
-  meanlog <- parameter(si_param_samples$param1)
-  sdlog <- parameter(si_param_samples$param2, c(0, Inf))
-  
-  # exclude missing values and replace them later
-  missing <- days <= 0
-  
   # compute the CDF of the lognormal distribution, for this and the next day
   days_lower <- days
   days_upper <- days + 1
   
+  if (fixed) {
+    # if fixed, we can do this in R
+    meanlog <- mean(si_param_samples$param1)
+    sdlog <- mean(si_param_samples$param2)
+    
+    p_lower <- plnorm(days_lower, meanlog, sdlog)
+    p_upper <- plnorm(days_upper, meanlog, sdlog)
+    
+  } else {
+    # otherwise create greta arrays for the parameters, with priors based on
+    # these samples
+    meanlog <- parameter(si_param_samples$param1)
+    sdlog <- parameter(si_param_samples$param2, c(0, Inf))
+    
+    p_lower <- lognormal_cdf(days_lower, meanlog, sdlog)
+    p_upper <- lognormal_cdf(days_upper, meanlog, sdlog)
+  }
+  
+
   # get the integral of the density over this day
-  p_upper <- lognormal_cdf(days_upper, meanlog, sdlog)
-  p_lower <- lognormal_cdf(days_lower, meanlog, sdlog)
   p_upper - p_lower
   
 }
@@ -692,10 +700,7 @@ fake_linelist <- function() {
     mutate(date_onset = ymd(date) - report_delay) %>% 
     rename(date_confirmation = date)
   
-  # remove delay column and throw away some of the true onset dates
-  linelist <- linelist %>%
-    select(-report_delay)
-  
[…]

### 2020-04-29 R/R_effective.R
fix up R_eff plotting to match EpiNow

@@ -113,13 +113,7 @@ local_cases <- date_by_state %>%
 
 library(greta.gp)
 
-# idx <- which(states == "VIC")
-# 
-# # summarise nationally to start with
-# local_cases <- local_cases[, idx, drop = FALSE]
-# imported_cases <- imported_cases[, idx, drop = FALSE]
-
-# prior on R0
+# lognormal prior on R0 (normal prior on log(R0))
 R0_prior <- lognormal_prior(2.6, 2)
 
 # the relative contribution of imported cases (relative to locally-acquired
@@ -143,7 +137,7 @@ lambda <- apply_serial_interval(case_contribution, fixed = TRUE)
 # but since this implies a MVN prior over log(R_eff), we can marginalise all of
 # these separate parameters and make the posterior much more nicely behaved:
 
-#  log(R0) ~ R0_mean + GP(0, bias())
+#  log(R0) ~ R0_mean + GP(0, white())
 #  gamma ~ GP(0, rbf(time))
 #  epsilon_i ~ GP(0, rbf1(time))
 #  epsilon ~ GP(0, rbf2(time) * iid(state))
@@ -152,7 +146,7 @@ lambda <- apply_serial_interval(case_contribution, fixed = TRUE)
 # These GPs can be combined as:
 
 #  log(R0) + b * social-distancing-index + gamma + epsilon = GP(0.723, K)
-#  K = bias() +
+#  K = white() +
 #      lin(social-distancing-index)
 #      rbf1(time) +
 #      rbf2(time) * iid(state) +
@@ -171,23 +165,35 @@ X <- cbind(date_vec,
 # define a GP on dates, using subset of regressors approximation
 l_ntnl <- lognormal(1, 0.5)
 l_state <- lognormal(1, 0.5)
-sigma_ntnl <- lognormal(-1, 1)
-sigma_state <- lognormal(-1, 1)
+
+# magnitude of the national error trend
+sigma_ntnl <- normal(0, 1, truncation = c(0, Inf))
+# magnitude of the state-level (difference from the national) error trend
+sigma_state <- normal(0, 1, truncation = c(0, Inf))
+# degree of variation between state timeseries (shrink towards 0)
 sigma_state_iid <- normal(0, 1, truncation = c(0, Inf))
 
-intercept_kernel <- bias(R0_prior$sd ^ 2)
+# standard deviation of IID noise associated with each day (case clustering)
+# we want the prior marginal variance to approximately match prior on log(R0),
[…]

### 2020-04-29 R/R_effective.R
add time-varying quarantine effectiveness

@@ -118,12 +118,17 @@ R0_prior <- lognormal_prior(2.6, 2)
 
 # the relative contribution of imported cases (relative to locally-acquired
 # cases) to transmission. I.e. one minus the effectiveness of quarantine
-# measures.
-import_contribution <- 1
+# measures. Assume it varies over time
+quarantine_effectiveness <- case_when(
+  dates < as.Date("2020-03-15") ~ 0.2,
+  dates <= as.Date("2020-03-27") ~ 0.5,
+  TRUE ~ 0.99,
+)
+import_contribution <- 1 - quarantine_effectiveness
 
 # disaggregate imported and local cases according to these probabilities to get
 # the expected number of infectious people in each state and time
-case_contribution <- local_cases + imported_cases * import_contribution
+case_contribution <- local_cases + sweep(imported_cases, 1, import_contribution, FUN = "*")
 lambda <- apply_serial_interval(case_contribution, fixed = TRUE)
 
 # NOTE: fixing the SI parameters at their prior means just for testing

### 2020-04-29 R/R_effective.R
add plots of trend and error

@@ -37,6 +37,7 @@
 library(dplyr)
 library(readr)
 library(tidyr)
+library(RColorBrewer)
 source("R/functions.R")
 
 # return a fake linelist based on the real case counts and samples of the
@@ -214,7 +215,7 @@ distribution(local_cases[valid]) <- poisson(expected_infections)
 
 m <- model(log_R_eff)
 
-draws <- mcmc(m, one_by_one = TRUE)
+draws <- mcmc(m, chains = 10, one_by_one = TRUE)
 
 r_hats <- coda::gelman.diag(draws, autoburnin = FALSE, multivariate = FALSE)$psrf[, 1]
 n_eff <- coda::effectiveSize(draws)
@@ -231,7 +232,6 @@ ci50 <- apply(R_eff_mat, 2, quantile, c(0.25, 0.75))
 
 df <- tibble(date = rep(dates, n_states),
              state = rep(states, each = n_dates),
-             social_distance = X[, 3],
              R_eff_mean = mean,
              R_eff_50_lo = ci50[1, ],
              R_eff_50_hi = ci50[2, ],
@@ -284,6 +284,141 @@ df %>%
 ggsave("outputs/figures/R_eff.png",
        width = 10,
        height = 16, scale = 0.8)
+
+# plot fixed effect trend in Rt (common to all states)
+X_one_state <- X[1:n_dates, ]
+distancing_effect <- project(zero_mean_gp,
+                             x_new = X_one_state,
+                             kernel = distancing_kernel)
+
+noise <- normal(0, sigma_noise, dim = n_dates)
+log_R_eff_trend <- R0_prior$mean + noise + distancing_effect
+R_eff_trend <- exp(log_R_eff_trend)
+
+R_eff_trend_sim <- calculate(R_eff_trend, values = draws, nsim = 10000)[[1]][, , 1]
+mean <- colMeans(R_eff_trend_sim)
+ci90 <- apply(R_eff_trend_sim, 2, quantile, c(0.05, 0.95))
+ci50 <- apply(R_eff_trend_sim, 2, quantile, c(0.25, 0.75))
+
+df_trend <- tibble(date = dates,
+                   R_eff_mean = mean,
+                   R_eff_50_lo = ci50[1, ],
+                   R_eff_50_hi = ci50[2, ],
+                   R_eff_90_lo = ci90[1, ],
+                   R_eff_90_hi = ci90[2, ])
+
+base_colour <- brewer.pal(8, "Set2")[1]
[…]

### 2020-04-29 R/R_effective.R
remove hierarchical error structure in the correlated errors

@@ -26,9 +26,8 @@
 # of quarantine measures.
 
 #   local-incidence_{t,i} ~ Poisson(lambda_{t,i} * R_{t,i})
-#   log(R_{t,i}) = R_0 + b * social-distancing-index_{t,i} + \epsilon_{t,i} + \gamma_t
-#   epsilon_i ~ GP(0, K1)
-#   gamma ~ GP(0, K2)
+#   log(R_{t,i}) = R_0 + b * social-distancing-index_{t,i} + \epsilon_{t,i}
+#   epsilon_i ~ GP(0, K)
 #   lambda_{t,i} = sum_{t'=1}^t p-serial-interval(t') (local-incidence_{t',i} + imported-incidence_{t',i} import-contribution_{t'})
 #   log(R_0) ~ N(0.723, 0.465)
 
@@ -135,26 +134,23 @@ lambda <- apply_serial_interval(case_contribution, fixed = TRUE)
 # NOTE: fixing the SI parameters at their prior means just for testing
 
 # define a GP to reflect the function:
-#   log(R_eff) <- log(R_0) + b * social-distancing-index_{t,i} + \epsilon_{t,i} + \gamma_t
+#   log(R_eff) <- log(R_0) + b * social-distancing-index_{t,i} + \epsilon_{t,i}
 #   epsilon_i ~ GP(0, K1)
-#   gamma ~ GP(0, K2)
 #   log(R_0) ~ N(0.723, 0.465)
 
 # but since this implies a MVN prior over log(R_eff), we can marginalise all of
 # these separate parameters and make the posterior much more nicely behaved:
 
 #  log(R0) ~ R0_mean + GP(0, white())
-#  gamma ~ GP(0, rbf(time))
 #  epsilon_i ~ GP(0, rbf1(time))
 #  epsilon ~ GP(0, rbf2(time) * iid(state))
 #  social-distancing-index * b ~ GP(0, lin(social-distancing-index))
 
 # These GPs can be combined as:
 
-#  log(R0) + b * social-distancing-index + gamma + epsilon = GP(0.723, K)
+#  log(R0) + b * social-distancing-index + epsilon = GP(0.723, K)
 #  K = white() +
 #      lin(social-distancing-index)
-#      rbf1(time) +
 #      rbf2(time) * iid(state) +
 
 # need to first put the data into a vector, and take this opportunity to exclude those with no infectious cases
@@ -169,12 +165,10 @@ X <- cbind(date_vec,
            social_distancing_index_vec)
 
[…]

### 2020-04-29 R/functions.R
update datastreams

@@ -5,16 +5,15 @@ library(stringr)
 # read in and tidy up Facebook movement data
 facebook_mobility <- function() {
   
-  file <- "data/fb_data/au_gadm_mobility_statistics.20200420.csv"
+  file <- "data/fb_data/au_gadm_mobility_statistics.20200427.csv"
   data <- readr::read_csv(file) %>%
     dplyr::select(
       state = polygon_name,
       date = ds,
-      "movement range" = all_day_bing_tiles_visited_relative_change,
       "staying still" = all_day_ratio_single_tile_users
     ) %>%
     tidyr::pivot_longer(
-      cols = c("movement range", "staying still"),
+      cols = c("staying still"),
       names_to = "metric",
       values_to = "trend"
     ) %>%
@@ -93,7 +92,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev18/v1/en-us/applemobilitytrends-2020-04-26.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev20/v2/en-us/applemobilitytrends-2020-04-27.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -133,7 +132,7 @@ apple_mobility <- function() {
 citymapper_mobility <- function() {
   
   # get link from: https://citymapper.com/cmi/about
-  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200427.csv"
+  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200429.csv"
   data <- readr::read_csv(url, skip = 3) %>%
     tidyr::pivot_longer(cols = -Date,
                         names_to = "region",

### 2020-04-29 R/latent_social_distancing.R
save latent factors estimates

@@ -576,6 +576,20 @@ ggsave("outputs/figures/state_distancing_waning_warning.png",
        width = 8,
        height = 6)
 
+# output posterior summaries for the different latent factors
+for(i in 1:4) {
+  filename <- latent_names[i] %>%
+    tolower() %>%
+    gsub(" ", "_", .) %>%
+    paste0("_latent.RDS") %>%
+    file.path("outputs", .)
+  summarise_vec_posterior(latents_ntnl[, i],
+                          draws) %>%
+    as_tibble() %>%
+    mutate(date = dates) %>%
+    saveRDS(filename)
+}
+
 # - plot state-by-factor loading plots
 # - pull more model code out into functions
 # - programatically find Apple download link

### 2020-04-29 R/R_effective.R
account for quarantine and social distancing effects

@@ -110,20 +110,45 @@ local_cases <- date_by_state %>%
   select(-import_status) %>%
   as.matrix()
 
-
+# load the social distancing factor
+distancing_file <- "outputs/social_distancing_latent.RDS"
+social_distancing_index <- distancing_file %>%
+  readRDS() %>%
+  select(mean, date) %>%
+  right_join(tibble(date = dates)) %>%
+  replace_na(list(mean = 0)) %>%
+  pull(mean)
+  
 library(greta.gp)
 
 # lognormal prior on R0 (normal prior on log(R0))
-R0_prior <- lognormal_prior(2.6, 2)
+R0_prior <- lognormal_prior(2.6, 1)
 
 # the relative contribution of imported cases (relative to locally-acquired
-# cases) to transmission. I.e. one minus the effectiveness of quarantine
+# cases) to transmission. I.e. one minus sthe effectiveness of quarantine
 # measures. Assume it varies over time
-quarantine_effectiveness <- case_when(
-  dates < as.Date("2020-03-15") ~ 0.2,
-  dates <= as.Date("2020-03-27") ~ 0.5,
-  TRUE ~ 0.99,
+
+# quarantine_effectiveness <- case_when(
+#   dates < as.Date("2020-03-15") ~ 0.2,
+#   dates <= as.Date("2020-03-27") ~ 0.5,
+#   TRUE ~ 0.99,
+# )
+
+quarantine_index <- case_when(
+  dates < as.Date("2020-03-15") ~ 1,
+  dates <= as.Date("2020-03-27") ~ 2,
+  TRUE ~ 3,
 )
+
+q3 <- uniform(0.99, 1)
+ratio1 <- uniform(0.5 / 0.99, 1)
+q2 <- q3 * ratio1
+ratio2 <- uniform(0.2 / 0.5, 1)
+q1 <- q2 * ratio2
+
+quarantine <- c(q1, q2, q3)
+quarantine_effectiveness <- quarantine[quarantine_index]
+
 import_contribution <- 1 - quarantine_effectiveness
 
 # disaggregate imported and local cases according to these probabilities to get
@@ -158,7 +183,7 @@ lambda_vec <- c(lambda)
 
 date_vec <- rep(date_nums, n_states)
 state_vec <- rep(seq_len(n_states), each = n_dates)
-social_distancing_index_vec <- rep(0, n_dates * n_states)
[…]

### 2020-04-30 R/R_effective.R
use infection not onset dates

@@ -69,11 +69,12 @@ linelist <- linelist %>%
 # - when defining the likelihood, marginalise over the sum of the known onsets
 # and probabilistic unknown onsets
 
-# build date-by-state matrices of the counts of new local and impiorted cases and
-# imports by date of onset
+# build date-by-state matrices of the counts of new local and imported cases and
+# imports by assumed date of infection (with an incubation period of 5 days)
 linelist <- linelist %>%
   rename(state = region,
          date = date_onset) %>%
+  mutate(date = date - 5) %>%
   select(-date_confirmation)
 
 import_statuses <- sort(unique(linelist$import_status))

### 2020-04-30 R/R_effective.R
use European prior for R0; reinstate IID noise (overdispersion)

@@ -122,8 +122,9 @@ social_distancing_index <- distancing_file %>%
   
 library(greta.gp)
 
-# lognormal prior on R0 (normal prior on log(R0))
-R0_prior <- lognormal_prior(2.6, 1)
+# lognormal prior on R0 (normal prior on log(R0)) based on estimates for
+# Northern Europe from Flaxman et al. (Imperial report 13)
+R0_europe <- R0_prior()
 
 # the relative contribution of imported cases (relative to locally-acquired
 # cases) to transmission. I.e. one minus sthe effectiveness of quarantine
@@ -195,26 +196,18 @@ X <- cbind(date_vec,
 # wiggliness of the state-level correlated errors
 l_state <- lognormal(2, 0.5)
 # magnitude of the state-level correlated errors
-sigma_state <- normal(0, 1, truncation = c(0, Inf))
+sigma_state <- normal(0, 0.5, truncation = c(0, Inf))
 # degree of variation between state timeseries (shrink towards 0)
-sigma_state_iid <- normal(0, 1, truncation = c(0, Inf))
+sigma_state_iid <- normal(0, 0.5, truncation = c(0, Inf))
+# amount of IID noise in log(R_eff) per day (overdispersion, or case clustering)
+sigma_noise <- normal(0, 0.5, truncation = c(0, Inf))
 
-# standard deviation of IID noise associated with each day (case clustering)
-# we want the prior marginal variance to approximately match prior on log(R0),
-# but also to shrink towards 0, so use a half normal prior with mean matching
-# the required standard deviation
-# sigma_noise_sd <- R0_prior$sd * sqrt(pi) / sqrt(2)
-# sigma_noise <- normal(0, sigma_noise_sd, truncation = c(0, Inf))
-# summary(calculate(sigma_noise, nsim = 10000)[[1]])
-
-
-# noise_kernel <- white(sigma_noise ^ 2)
-# try setting a prior over the true R0 for Australia - maybe the covariate will let it stay high?
-noise_kernel <- bias(R0_prior$sd ^ 2)
+noise_kernel <- white(sigma_noise ^ 2)
+R0_kernel <- bias(R0_europe$sdlog ^ 2)
 distancing_kernel <- linear(1, columns = 3)
 error_kernel <- rbf(l_state, sigma_state ^ 2, columns = 1) * iid(sigma_state_iid ^ 2, columns = 2)
 
[…]

### 2020-04-30 R/functions.R

@@ -797,3 +797,53 @@ tf_iid <- function(X, X_prime, variance, active_dims) {
 
 }
 
+R0_prior <- function() {
+  # Visually estimated values of the posterior distributions  (means and lower
+  # and upper 95% CIs) over R0 in 11 European countries from Flaxman et al.
+  # (Imperial report 13). Rather than use the estimates in the main text, we
+  # usee those fromm the SI under the shortest serial interval - a mean of 5
+  # days, corresponding to a median of 4.49 days. Note, we assumed based on
+  # Figure 7 that the SI gamma distribution parameter values given are the mean
+  # and the rate parameter of the distribution), since no other combination of
+  # mean, sd, shape, scale, or rate could be found to match the figure closely.
+  # We use this lower estimate because it is closest to the estimate of Nishiura
+  # et al. of a median serial interval of 4 days, or 4.6 days taking only the
+  # pairs of cases with greatest confidence. This is also the most conseervative
+  # eestimate of R0 and therefore of the impact of social distancing interventions
+  prior_estimate <- tibble::tribble(
+    ~country, ~lower, ~mean, ~upper,
+    "Denmark", 2, 2.6, 3.3,
+    "Italy", 2.4, 2.6, 2.8,
+    "Germany", 2.5, 3, 3.75,
+    "Spain", 2.7, 3.2, 3.6,
+    "United Kingdom", 2.4, 2.6, 3.1,
+    "France", 2.5, 2.8, 3.2,
+    "Norway", 1.7, 2.4, 3.3,
+    "Belgium", 2.4, 2.9, 3.7,
+    "Austria", 2.3, 2.7, 3.7,
+    "Sweden", 2.3, 2.8, 3.6,
+    "Switzerland", 2.3, 2.7, 3.4
+  ) %>%
+    summarise(lower = mean(lower),
+              mean = mean(mean),
+              upper = mean(upper))
+  
+  # to find a lognormal prior that matches this, find the parameters of a
+  # lognormal distribution that give the minimum mean squared error against
+  # these estiamtes
+  obj <- function(par) {
+    meanlog <- par[1]
+    sdlog <- exp(par[2])
+    mean <- exp(meanlog + sdlog ^ 2 / 2)
+    cis <- qlnorm(c(0.025, 0.975), meanlog, sdlog)
+    mean_error <- mean - prior_estimate$mean
[…]

### 2020-04-30 R/R_effective.R
relax priors on quarantine effectiveness

@@ -128,13 +128,7 @@ R0_europe <- R0_prior()
 
 # the relative contribution of imported cases (relative to locally-acquired
 # cases) to transmission. I.e. one minus sthe effectiveness of quarantine
-# measures. Assume it varies over time
-
-# quarantine_effectiveness <- case_when(
-#   dates < as.Date("2020-03-15") ~ 0.2,
-#   dates <= as.Date("2020-03-27") ~ 0.5,
-#   TRUE ~ 0.99,
-# )
+# measures. Assume it varies over time, based around these federal announcements
 
 quarantine_index <- case_when(
   dates < as.Date("2020-03-15") ~ 1,
@@ -142,10 +136,13 @@ quarantine_index <- case_when(
   TRUE ~ 3,
 )
 
-q3 <- uniform(0.99, 1)
-ratio1 <- uniform(0.5 / 0.99, 1)
+# assume at least 95% imports removed from circulation in the last period
+# (enforced quarantine), but just assume that quarantine efectiveness in earlier
+# periods was increasing
+q3 <- uniform(0.95, 1)
+ratio1 <- uniform(0, 1)
 q2 <- q3 * ratio1
-ratio2 <- uniform(0.2 / 0.5, 1)
+ratio2 <- uniform(0, 1)
 q1 <- q2 * ratio2
 
 quarantine <- c(q1, q2, q3)

### 2020-04-30 R/R_effective.R
increase the resolution of the subset of regressors approximation

@@ -207,7 +207,7 @@ error_kernel <- rbf(l_state, sigma_state ^ 2, columns = 1) * iid(sigma_state_iid
 kernel <- noise_kernel + R0_kernel + distancing_kernel + error_kernel
 
 # build a matrix of inducing points, one every 7 days but with one at the end
-inducing_date_nums <- seq(n_dates, 1, by = -7)
+inducing_date_nums <- seq(n_dates, 1, by = -3)
 inducing_index <- which(X[, 1] %in% inducing_date_nums)
 X_inducing <- X[inducing_index, ]
 
@@ -236,6 +236,7 @@ max(r_hats)
 min(n_eff)
 
 summary(calculate(quarantine, values = draws))
+
 
 R_eff <- exp(log_R_eff)
 
@@ -303,12 +304,11 @@ ggsave("outputs/figures/R_eff.png",
 
 # plot fixed effect trend in Rt (common to all states)
 X_one_state <- X[1:n_dates, ]
-trend_kernel <- noise_kernel + distancing_kernel
+trend_kernel <- R0_kernel + distancing_kernel
 distancing_effect <- project(zero_mean_gp,
                              x_new = X_one_state,
                              kernel = trend_kernel)
 
-# noise <- normal(0, sigma_noise, dim = n_dates)
 log_R_eff_trend <- R0_europe$meanlog + distancing_effect
 R_eff_trend <- exp(log_R_eff_trend)
 
@@ -433,22 +433,10 @@ ggsave("outputs/figures/R_eff_error.png",
        width = 10,
        height = 16, scale = 0.8)
 
+# posterior summary of R0
+R0_draws <- R_eff_trend_sim[, 1]
+mean(R0_draws)
+sd(R0_draws)
 
 # - add a custom greta function for lognormal CDF (using TFP) and try
 #   sampling with that
-
-# # run epinow too
-# epinow_cases <- linelist %>%
-#   rename(date = date_confirmation) %>%
-#   group_by(date, region, import_status) %>%
-#   summarise(cases = n())
-# 
-# epinow_linelist <- linelist %>%
-#   rename(date_confirm = date_confirmations)
-# 
-# EpiNow::regional_rt_pipeline(
-#   cases = epinow_cases,
-#   linelist = epinow_linelist,
-#   case_limit = 20
-# )
-

### 2020-04-30 R/R_effective.R
add code for current R_eff summary

@@ -438,5 +438,10 @@ R0_draws <- R_eff_trend_sim[, 1]
 mean(R0_draws)
 sd(R0_draws)
 
+# posterior summary of R_eff for the llatest date
+R_eff_now_draws <- R_eff_trend_sim[, ncol(R_eff_trend_sim)]
+mean(R_eff_now_draws)
+sd(R_eff_now_draws)
+max(dates)
 # - add a custom greta function for lognormal CDF (using TFP) and try
 #   sampling with that

### 2020-04-30 R/functions.R
drop facebook from mobility analysis

@@ -171,14 +171,14 @@ all_mobility <- function() {
       -region,
       -transportation_type
     )
-  
-  facebook <- facebook_mobility() %>%
-    mutate(
-      datastream = str_c("Facebook: ", metric)
-    ) %>%
-    dplyr::select(
-      -metric
-    )
+
+  # facebook <- facebook_mobility() %>%
+  #   mutate(
+  #     datastream = str_c("Facebook: ", metric)
+  #   ) %>%
+  #   dplyr::select(
+  #     -metric
+  #   )
   
   citymapper <- citymapper_mobility() %>%
     mutate(
@@ -189,7 +189,7 @@ all_mobility <- function() {
   bind_rows(
     google,
     apple,
-    facebook,
+    # facebook,
     citymapper
   )
   

### 2020-04-30 R/latent_social_distancing.R

@@ -288,7 +288,7 @@ for (j in seq_len(n_states)) {
 # same for a subset of datastreams, by all states
 target_datastreams <- c("Apple: directions for driving",
                         "Google: time at residential",
-                        "Facebook: movement range")
+                        "Google: time at grocery and pharmacy")
 
 # get the vector of state-datastreams to plot
 state_datastreams_plot <- mobility %>%

### 2020-04-30 manuscript/ms.Rmd

@@
 
-All of these datastreams are derived from apps that are primarily smartphone-based.
-As such they overrepresent demographics that are more likely to own smartphones and have account with these technology companies.
-Data on the demographics of the users that contributed to these particular datastreams are not available.
-However rough public estimates suggest that 90% of Facebook users are aged between 18 and 64.
-Younger and older demographics are underrepresented, with 4.5% facebook users vs 16% in the general population over the age of 65, and 5.6% vs 20% under the age of 18.
+All but one of these datastreams are derived from primarily smartphone-based apps.
+As such, they overrepresent demographics that are more likely to own smartphones and have accounts with these technology companies.
+In Australia, the penetration of smartphones in the population is close to saturation.
+Detailed data on the demographics of the users that contributed to these particular datastreams are not available for Australia.
+However, rough global estimates suggest that 90% of Facebook users are aged between 18 and 64.
+Younger and older demographics are underrepresented, with 4.5% Facebook users vs 16% in the general population over the age of 65, and 5.6% vs 20% under the age of 18.
 Given that school-age children are underrepresented in COVID-19 cases in Australia, and that those over 65 are likely to be less mobile, these data may be unintentionally biased towards those contributing most strongly to transmission.
-Apple's and Citymapper's direction requests likely represent deliberate visits to a specific location, and a location that the user either does not usually visit (and therefore requires directions) or would benefit from live traffic updates for.
-Whilst Google's index of the time spent in residential places is likely to be a good proxy for people staying at home, it does not exclude the possibility that this time is spent in another person's household.
+Deatils about the nature of data collection and processing of the Apple's and Citymapper's direction requests are also unclear.
+They likely represent deliberate visits to a specific location, and a location that the user either does not usually visit (and therefore requires directions) or forwhich the user would would benefit from live traffic updates.
+Google - through its extensive user base that includes all Android OS phoine users is likely to capture well information across demographics.
+Whilst Google's index of the time spent in residential places is therefore likely to be a good proxy for the amount of people staying at home in the general population, it does not exclude the possibility that this time is spent in another person's household.
+Such effects can, in general, not be well filtered out.
+
+Finally, the data provided by NBN are the most distinct from the rest of the data analysed here -- they are guaranteed to be generated by users taht are static at home.
+They only include domestic users of NBN connections (although a number of small numer of self-emplyed customers may profit from domestic pricing plans).
+While this dataset is unable to provide an indication of mobility, it serves as a good proxy for a change in internet access from home, which is hypothesized to directly correlate with active internet use.
+We assume here that the size of a standard accessed item ( website, movie, music file) has not increased substantially in the monitored period, and therefore any change in data volume transferred relates direclty to active interaction with data.
 
@@
 The large-scale aggregation of these datasets ensures the privacy of users - the smallest population is that of the Northern Territory: over 245,000.
+The NBN dataset is provided exclusively to the University of Melbourne research team (contact M. Tomko).
 

### 2020-04-30 manuscript/ms.Rmd
Adding a first pass of the introduction and a few comments in the discussion.

--- 2020-04-28 manuscript/ms.Rmd
@@
 
-## Gist (to be split up into a paper format)
-
-Social distancing behaviour - advised and enforced by governments around the world - has been a cornerstone of the global response to the COVID-19 pandemic.
-Quantifying how the behaviour of a population has changed in response to these public health measures, is critical to monitoring the ongoing effectiveness of these measures, identifying which are most impactful, and as inputs to modelling to forecast how the measures will aimpact ont he course of the outbreak.
-
-A variety of data sources have are available that can quantify different aspects of social distancing behaviour at a population level.
-Aggregated user-tracking data from technology companies provide timeseries of: how often users leave home, the lengths of journeys they undertake when they do, the types of locations they visited, and the locations of those places.
-
-Whilst these multiple data sources provide varied perspectives on population behaviour, they vary significantly in the subset of the population they reflect and the aspects of behaviour being considered.
-For example, data from route-planning apps is more likely to reflect unfamiliar journeys than trips to places regularly visited, and data on the types of places being visited does not indicate how far users travel to visit them.
-The different, and often non-overlapping spatial scales and time frames of these data sources also hinders direct comparison and synthesis of them.
-
-Moreover, no single one of these types of data fully encompasses the range of behaviours that affect transmission of COVID-19 - those that can be sumamrised in a mathematical modelling framework via the mixing parameter of a transmission model.
-
-The aim of this work is to synthesise these varied measures of population mixing behaviour into a small, quantified set of latent social distancing factors that reflect the underlying changes in behaviour.
-We consider these latent social distancing factors in a semi-mechanistic way; reflecting knowledge and hypotheses of the events that triggered and influenced these behavioural trends.
-To do this, we construct a statistical semi-mechanistic latent factor model that considers dates that events, policy measures, and public announcements are made, and estimates how the behaviour of the population responded to each event by calibration against all of the available metrics of population mixing behaviour.
-
-We apply this model to 12 separate data streams provided by 4 technology companies for Australia, with spatial data varying between national, state and city-levels.
-This reveals two distinct latent population behavioural factors: a strong social distancing effect in response to national announcements and policies, preceeded by a small and short-lived rise in mixing apparently driven by the population 'stocking up' on supplies.
+## Gist
 
@@
 
- - COVID-19 background
- - Social distancing policies
- - Population mixing/mobility studies
- - Timeline of data availability
+The novel coronavirus SARS-CoV-2 first emerged in the city of Wuhan, China, in December 2019.
+Despite early containment efforts, by early March 2020 SARS-CoV-2 had spread to all global regions and overwhelmed some of the world’s strongest health systems.
+As of 29 April 2020, over 2.9 million cases of the disease (known as COVID-19) had been confirmed globally, including more than 200,000 deaths.
+
+In the absence of pharmaceutical interventions, ‘physical distancing’ behaviour —-- advised and enforced by governments around the world --- has been a cornerstone of the global response to COVID-19.
+Many countries have successfully mitigated an initial epidemic wave of COVID-19, using a combination of border, case targeted and physical distancing measures.
+Physical distancing or social distancing measures may include the closure of schools, businesses and workplaces, and limits on the size of public gatherings.
+The goal of these strategies is to reduce normal levels of physical and social population mixing in order to slow the spread of the infection and intensity of the epidemic, thus reducing the risk of overwhelming the health system.
+The specific physical distancing measures, the intensity and timing of their implementation, the public response to them, and the level of government enforcement, has varied vastly across the globe, including within countries.
+
+For example, on 16 March, the Australian Government imposed restrictions on outdoor public gatherings of more than 500 people in order to reduce and prevent further COVID-19 transmission.
+Further restrictions were incrementally imposed over the following 10 days.
+By March 27, all Australians were strongly encouraged to leave their homes only for limited essential activities and all public gatherings were limited to two people.
+These measures were in addition to self-isolation advice for (mild) confirmed cases and their contacts, as well as for returned travellers.
+By late March, daily counts of new cases were declining, and the effective reproduction number was estimated to be below 1, indicating that these measures had successfully mitigated the epidemic.
+
+Australia --- and many other countries with intensive physical distancing measures in place --- is now shifting its focus from reacting to a first epidemic wave of COVID-19 to proactive planning for gradual and targeted relaxation of physical distancing measures.
+Quantifying how the behaviour of the population has changed in response to these measures and the potential impact on disease transmission is critical for the next stage of response planning.
+
+A variety of data sources are available that can quantify different aspects of social distancing behaviour at a population level.
+Aggregated user-tracking data from technology companies provide timeseries of: how often users leave home, the lengths of journeys they undertake when they do, the types of locations they visited, and the locations of those places.
+
+While these multiple data sources provide varied perspectives on population behaviour, they vary significantly in the subset of the population they reflect, and the aspects of behaviour being considered.
+For example, data from route-planning apps is more likely to reflect unfamiliar journeys than trips to places regularly visited, and data on the types of places being visited does not indicate how far users travel to visit them.
+The different, and often non-overlapping spatial scales and time frames of these data sources also hinders direct comparison and synthesis of them.
+
+Moreover, no single one of these types of data fully encompasses the range of behaviours that affect transmission of COVID-19 *i.e.*, those that can be summarised in a mathematical modelling framework via the mixing parameter of a transmission model.
+
+Here we synthesise these varied measures of population mixing behaviour into a small, quantified set of latent physical distancing factors that reflect the underlying changes in behaviour.
+To do so, we construct a statistical latent factor model that considers dates of key events, including COVID-19 policy announcements and public holidays.
+The model estimates how population behaviour changed in response to each event by calibrating against all available metrics of population mixing behaviour.
+We apply this model to 12 separate data streams provided by four technology companies for Australia, with spatial data varying between national, state and city-levels.
+We then use outputs from the latent factor model to inform estimates of the effective reproduction number for each Australian state in March and April, and finally, to quantify the relative effectiveness of key physical distancing measures implemented in late March.
 
@@
 
+Summary of results: This reveals two distinct latent population behavioural factors: a strong social distancing effect in response to national announcements and policies, preceeded by a small and short-lived rise in mixing apparently driven by the population 'stocking up' on supplies.
+
+Our work will enable monitoring of the ongoing effectiveness of physical distancing measures in Australia, and the generation of model inputs for estimating the impact of these measures on the future course of the epidemic.
+
+The impacts of physical distancing efforts have been in addition to extensive case targeted efforts (case isolation and contact quarantine) which cannot be disentangled from our results.
+Case targeted measures will be particularly important for disease management as physical distancing measures are gradually relaxed.
+Further studies, potentially using mathematical modelling frameworks, are required to analyse the relationship between physical distancing behaviour, case targeted interventions, and disease transmission.
+
  - Use for monitoring adherence.

### 2020-05-01 R/R_effective.R
use real linelist

@@ -26,7 +26,7 @@
 # of quarantine measures.
 
 #   local-incidence_{t,i} ~ Poisson(lambda_{t,i} * R_{t,i})
-#   log(R_{t,i}) = R_0 + b * social-distancing-index_{t,i} + \epsilon_{t,i}
+#   log(R_{t,i}) = log(R_0) + b * social-distancing-index_{t,i} + \epsilon_{t,i}
 #   epsilon_i ~ GP(0, K)
 #   lambda_{t,i} = sum_{t'=1}^t p-serial-interval(t') (local-incidence_{t',i} + imported-incidence_{t',i} import-contribution_{t'})
 #   log(R_0) ~ N(0.723, 0.465)
@@ -42,7 +42,7 @@ source("R/functions.R")
 # return a fake linelist based on the real case counts and samples of the
 # reporting delay distribution
 set.seed(2020-04-29)
-linelist <- fake_linelist()
+linelist <- readRDS("~/not_synced/nnds/linelist_formatted.RDS")
 
 # for now just do single imputation with the mean delay on the date of onset for
 # those missing it
@@ -212,7 +212,6 @@ inducing_index <- which(X[, 1] %in% inducing_date_nums)
 X_inducing <- X[inducing_index, ]
 
 zero_mean_gp <- gp(X, kernel, inducing = X_inducing, tol = 0)
-
 log_R_eff <- R0_europe$meanlog + zero_mean_gp
 
 # work out which ones to exclude (because there were no infectious people)
@@ -221,7 +220,6 @@ valid <- which(lambda_vec_vals > 0)
 
 # combine with lambda on log scale for numerical stability - since greta can use
 # this in evaluating the poisson likelihood
-# log_expected_infections <- sweep(log(lambda), 1, log_R_eff, FUN = "+") 
 log_expected_infections <- log(lambda_vec[valid]) + log_R_eff[valid]
 expected_infections <- exp(log_expected_infections)
 distribution(local_cases[valid]) <- poisson(expected_infections)
@@ -234,8 +232,6 @@ r_hats <- coda::gelman.diag(draws, autoburnin = FALSE, multivariate = FALSE)$psr
 n_eff <- coda::effectiveSize(draws)
 max(r_hats)
 min(n_eff)
-
-summary(calculate(quarantine, values = draws))
 
 
 R_eff <- exp(log_R_eff)

### 2020-05-01 R/R_effective.R
switch to rational quadratic kernel

@@ -191,7 +191,8 @@ X <- cbind(date_vec,
 # define a GP on dates, using subset of regressors approximation
 
 # wiggliness of the state-level correlated errors
-l_state <- lognormal(2, 0.5)
+a_state <- lognormal(2, 0.5)
+l_state <- lognormal(4, 0.5)
 # magnitude of the state-level correlated errors
 sigma_state <- normal(0, 0.5, truncation = c(0, Inf))
 # degree of variation between state timeseries (shrink towards 0)
@@ -202,12 +203,18 @@ sigma_noise <- normal(0, 0.5, truncation = c(0, Inf))
 noise_kernel <- white(sigma_noise ^ 2)
 R0_kernel <- bias(R0_europe$sdlog ^ 2)
 distancing_kernel <- linear(1, columns = 3)
-error_kernel <- rbf(l_state, sigma_state ^ 2, columns = 1) * iid(sigma_state_iid ^ 2, columns = 2)
+error_kernel <- rational_quadratic(
+  lengthscales = l_state,
+  variance = sigma_state ^ 2,
+  alpha = a_state,
+  columns = 1) *
+  iid(sigma_state_iid ^ 2,
+      columns = 2)
 
 kernel <- noise_kernel + R0_kernel + distancing_kernel + error_kernel
 
 # build a matrix of inducing points, one every 7 days but with one at the end
-inducing_date_nums <- seq(n_dates, 1, by = -3)
+inducing_date_nums <- seq(n_dates, 1, by = -5)
 inducing_index <- which(X[, 1] %in% inducing_date_nums)
 X_inducing <- X[inducing_index, ]
 

### 2020-05-01 R/R_effective.R
separately estimate the impacts of quarantine and distancing

@@ -123,314 +123,208 @@ social_distancing_index <- distancing_file %>%
 library(greta.gp)
 
 # lognormal prior on R0 (normal prior on log(R0)) based on estimates for
-# Northern Europe from Flaxman et al. (Imperial report 13)
+# Northern Europe from Flaxman et al. (Imperial report 13, lowest R0s from their
+# sensitivity analysis)
 R0_europe <- R0_prior()
+log_R0 <- normal(R0_europe$meanlog, R0_europe$sdlog)
 
-# the relative contribution of imported cases (relative to locally-acquired
-# cases) to transmission. I.e. one minus sthe effectiveness of quarantine
-# measures. Assume it varies over time, based around these federal announcements
+# the reduction from R0 down to R_eff for imported cases due to different
+# quarantine measures each measure applied during a different period. Q_t is
+# R_eff_t / R0 for each time t, modelled as a monotone decreasing step function
+# over three periods with increasingly strict policies
 
-quarantine_index <- case_when(
+q_index <- case_when(
   dates < as.Date("2020-03-15") ~ 1,
   dates <= as.Date("2020-03-27") ~ 2,
   TRUE ~ 3,
 )
 
-# assume at least 95% imports removed from circulation in the last period
-# (enforced quarantine), but just assume that quarantine efectiveness in earlier
-# periods was increasing
-q3 <- uniform(0.95, 1)
-ratio1 <- uniform(0, 1)
-q2 <- q3 * ratio1
-ratio2 <- uniform(0, 1)
-q1 <- q2 * ratio2
+q_raw <- uniform(0, 1, dim = 3)
+q <- cumprod(q_raw)
+q_index_vec <- rep(q_index, n_states)
+log_q <- log(q)
+log_Qt_vec <- log_q[q_index_vec]
+# log_q_raw <- -exponential(1, dim = 3)
+# log_q <- cumsum(log_q_raw)
+# log_Qt <- log_q[q_index]
 
-quarantine <- c(q1, q2, q3)
-quarantine_effectiveness <- quarantine[quarantine_index]
+# the reduction from R0 down to R_eff for locally-acquired cases due to social
+# distancing behaviour, modelled as a being proportional to the change in
+# mobility measured from the mobility datastreams
+beta <- normal(0, 1)
+log_Dt_vec <- beta * rep(social_distancing_index, n_states)
[…]

### 2020-05-01 R/functions.R

@@ -847,3 +847,87 @@ R0_prior <- function() {
   list(meanlog = par[1], sdlog = exp(par[2]))
   
 }
+
+plot_trend <- function(simulations,
+                       base_colour = grey(0.4),
+                       multistate = FALSE,
+                       hline_at = 1,
+                       ylim = c(0, 3),
+                       vline_at = NA,
+                       keep_only_rows = NULL) {
+  
+  library(ggplot2)
+  
+  mean <- colMeans(simulations)
+  ci_90 <- apply(simulations, 2, quantile, c(0.05, 0.95))
+  ci_50 <- apply(simulations, 2, quantile, c(0.25, 0.75))
+  
+  if (multistate) {
+    dates <- rep(dates, n_states)
+    states <- rep(states, each = n_dates)
+  } else {
+    states <- NA
+  }
+  
+  df <- tibble(date = dates,
+               state = states,
+               mean = mean,
+               ci_50_lo = ci_50[1, ],
+               ci_50_hi = ci_50[2, ],
+               ci_90_lo = ci_90[1, ],
+               ci_90_hi = ci_90[2, ])
+  
+  if (!is.null(keep_only_rows)) {
+    df <- df[keep_only_rows, ]
+  }
+  
+  df <- df %>%
+    filter(date >= as.Date("2020-03-01")) %>%
+    mutate(type = "Nowcast")
+  
+  if (is.null(ylim)) {
+    ylim <- c(min(df$ci_90_lo), max(df$ci_90_hi)) 
+  }
+  
+  p <- ggplot(df) + 
+    
+    aes(date, mean, fill = type) +
+    
+    xlab(element_blank()) +
+    
+    coord_cartesian(ylim = ylim) +
+    scale_y_continuous(position = "right") +
+    scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
+    scale_alpha(range = c(0, 0.5)) +
+    scale_fill_manual(values = c("Nowcast" = base_colour)) +
+    
+    geom_ribbon(aes(ymin = ci_90_lo,
+                    ymax = ci_90_hi),
+                alpha = 0.2) +
+    geom_ribbon(aes(ymin = ci_50_lo,
+                    ymax = ci_50_hi),
+                alpha = 0.5) +
+    geom_line(aes(y = ci_90_lo),
+              colour = base_colour,
+              alpha = 0.8) + 
+    geom_line(aes(y = ci_90_hi),
+              colour = base_colour,
[…]

### 2020-05-01 R/R_effective.R
interpolate proportion imported

@@ -132,10 +132,11 @@ log_R0 <- normal(R0_europe$meanlog, R0_europe$sdlog)
 # quarantine measures each measure applied during a different period. Q_t is
 # R_eff_t / R0 for each time t, modelled as a monotone decreasing step function
 # over three periods with increasingly strict policies
+quarantine_dates <- as.Date(c("2020-03-15", "2020-03-28"))
 
 q_index <- case_when(
-  dates < as.Date("2020-03-15") ~ 1,
-  dates <= as.Date("2020-03-27") ~ 2,
+  dates < quarantine_dates[1] ~ 1,
+  dates < quarantine_dates[2] ~ 2,
   TRUE ~ 3,
 )
 
@@ -236,25 +237,40 @@ n_eff <- coda::effectiveSize(draws)
 max(r_hats)
 min(n_eff)
 
-# estimate R effective for all cases as a case-weighted sum of the two?
-all_infectious <- local_infectious + imported_infectious
-p_imported <- imported_infectious / all_infectious
-p_imported[local_infectious == 0]  <- 1
-p_imported[imported_infectious == 0]  <- 0
-keep <- all_infectious > 0
+# estimate the overall R_eff by weighting using the (interpolated) proportion of
+# infectious cases that are imports
+p_imported <- proportion_imported(local_infectious, imported_infectious, X)
 
 R_eff_loc_weighted <- R_eff_loc * (1 - c(p_imported))
 R_eff_imp_weighted <- c(sweep(p_imported, 1, R_eff_imp_trend, FUN = "*"))
-
 R_eff <- R_eff_loc_weighted + R_eff_imp_weighted 
 
-# R_eff of locally-acquired cases
-R_eff_sim <- calculate(R_eff, values = draws, nsim = 1e4)[[1]]
+# simulate from posterior for quantitites of interest
 
-# pass in a masking variable to plot_trend to remove some dates?
+# R_eff of all cases
+R_eff_sim <- posterior_sims(R_eff, draws)
+
+# R_eff of locally-acquired cases
+R_eff_loc_sim <- posterior_sims(R_eff_loc, draws)
+
+# average R_eff of locally-acquired cases
+R_eff_loc_trend <- exp(log_R0 + log_Dt)
+R_eff_loc_trend_sim <- posterior_sims(R_eff_loc_trend, draws)
+
+# average R_eff of overseas-acquired cases
+R_eff_imp_trend <- exp(log_R0 + log_Qt)
[…]

### 2020-05-01 R/functions.R

@@ -452,12 +452,19 @@ state_populations <- function() {
   )
 }
 
+posterior_sims <- function(vector, draws, nsim = NULL) {
+  if (is.null(nsim)) {
+    nsim <- coda::niter(draws) * coda::nchain(draws)
+  }
+  calculate(vector, values = draws, nsim = nsim)[[1]][, , 1]
+}
+
 # summarise the posterior for a vector greta array
 summarise_vec_posterior <- function(vector,
                                     draws,
                                     quantiles = c(0.025, 0.975),
                                     nsim = 1000) {
-  vector_sim <- calculate(vector, values = draws, nsim = nsim)[[1]][, , 1]
+  vector_sim <- posterior_sims(vector, draws, nsim = nsim)[, , 1]
   posterior_mean <- colMeans(vector_sim)
   posterior_ci <- t(apply(vector_sim, 2, quantile, quantiles))
   cbind(mean = posterior_mean, posterior_ci)
@@ -901,6 +908,8 @@ plot_trend <- function(simulations,
     scale_alpha(range = c(0, 0.5)) +
     scale_fill_manual(values = c("Nowcast" = base_colour)) +
     
+    geom_vline(xintercept = vline_at, colour = "grey80") +
+    
     geom_ribbon(aes(ymin = ci_90_lo,
                     ymax = ci_90_hi),
                 alpha = 0.2) +
@@ -914,6 +923,7 @@ plot_trend <- function(simulations,
               colour = base_colour,
               alpha = 0.8) + 
     
+
     geom_hline(yintercept = hline_at, linetype = "dotted") +
     
     cowplot::theme_cowplot() +
@@ -931,3 +941,29 @@ plot_trend <- function(simulations,
   
 }
 
+
+# interpolate proportion of infectious cases that are imports
+proportion_imported <- function (local_infectious, imported_infectious, X) {
+  
+  all_infectious <- local_infectious + imported_infectious
+  prop_imported <- imported_infectious / all_infectious
+  
+  # transform to vector on unconstrained scale
+  prop_imported <- pmax(0.01, prop_imported)
+  prop_imported <- pmin(0.99, prop_imported)
+  q_imported <- c(qlogis(prop_imported))
+  
+  # interpolate state-by-state  
+  df <- as.data.frame(X)
+  df$state <- factor(df$state)
+  model <- mgcv::gam(response_q ~ s(date) +
+                       s(date, by = state, k = 30),
+                     data = df)
[…]

### 2020-05-01 R/R_effective.R
try a combined epsilon

@@ -1,18 +1,69 @@
-# fit a Bayesian model-based estimate of R_effective over time.
+# fit a Bayesian model-based estimate of R_effective over time, quantifying the
+# impacts of both quarantine and physical distancing measures.
 
-# The model assumes that the number of new locally-acquired cases at time $t$ in
-# region $i$ is *conditionally* Poisson-distributed with expectation given by
-# the number of infectious people $lambda_{t,i}$ multiplied by the region- and
-# time-varying reproduction rate $R_{t,i}$. To account for clusters in numbers
+# The basic model assumes that the number of new locally-acquired cases
+# $N_{ti}$ at time $t$ in region $i$ is *conditionally* Poisson-distributed
+# with expectation $lambda_{ti}$ given by the product of the number of
+# infectious cases $I_{ti}$ and the time-varying reproduction rate $R_{ti}$:
+
+#   N_{ti} &\sim Poisson(lambda_{ti})
+#   lambda_{ti} &= I_{ti} R_{ti}
+
+# We extend this model by considering separate reproduction rates for two groups
+# of infectious cases in order to model the effects of different interventions
+# targetted at each group: those with locally-acquired cases $I_{ti}^L$, and
+# those with overseas acquired cases $I_{ti}^O$, with corresponding reproduction
+# rates $R_{ti}^L$ and $R_{ti}^O$:
+
+#   lambda_{ti} &= I_{ti}^L R_{ti}^L + I_{ti}^O R_{ti}^O
+#   R_{ti}^L &= R_0 D_t e^{\epsilon_{ti}^L}
+#   R_{ti}^O &= R_0 Q_t e^{\epsilon_{ti}^O}
+
+# where each of these reproduction rates is modelled as a product of: the
+# reproduction rate under initial conditions and no interventions $R_0$;
+# deterministic functions $D_t$ and $Q_t$ that modify $R_0$ over time to
+# respectively represent the impacts of physical distancing and quarantine
+# interventions at a national level, and correlated timeseries of random effects
+# $\epsilon_{ti}^L$ and $\epsilon_{ti}^O$ to represent stochastic fluctuations
+# in the reporting rate in each state, for example due to the clusters in
[…]

### 2020-05-01 R/functions.R

@@ -956,14 +956,17 @@ proportion_imported <- function (local_infectious, imported_infectious, X) {
   # interpolate state-by-state  
   df <- as.data.frame(X)
   df$state <- factor(df$state)
-  model <- mgcv::gam(response_q ~ s(date) +
+  model <- mgcv::gam(q_imported ~ s(date) +
                        s(date, by = state, k = 30),
                      data = df)
   q_imported <- predict(model, newdata = df)
   
-  # transform back and return
+  # transform back, clamp to numerically stable values, and return
   p_imported <- plogis(q_imported)
   dim(p_imported) <- dim(all_infectious)
+  eps <- sqrt(.Machine$double.eps)
+  p_imported <- pmax(eps, p_imported)
+  p_imported <- pmin(1 - eps, p_imported)
   p_imported
   
 }

### 2020-05-02 R/R_effective.R
use model with two error terms, and write up notation in comments

@@ -2,87 +2,78 @@
 # impacts of both quarantine and physical distancing measures.
 
 # The basic model assumes that the number of new locally-acquired cases
-# $N_{ti}$ at time $t$ in region $i$ is *conditionally* Poisson-distributed
-# with expectation $lambda_{ti}$ given by the product of the number of
-# infectious cases $I_{ti}$ and the time-varying reproduction rate $R_{ti}$:
+# $N_{t,i}^L$ at time $t$ in region $i$ is *conditionally* Poisson-distributed
+# with expectation $lambda_{t,i}$ given by the product of the number of
+# infectious cases $I_{t,i}$ and the time-varying reproduction rate $R_{t,i}$:
 
-#   N_{ti} &\sim Poisson(lambda_{ti})
-#   lambda_{ti} &= I_{ti} R_{ti}
+#   N_{t,i}^L &\sim Poisson(lambda_{t,i})
+#   lambda_{t,i} &= I_{t,i} R_{t,i}
 
 # We extend this model by considering separate reproduction rates for two groups
 # of infectious cases in order to model the effects of different interventions
-# targetted at each group: those with locally-acquired cases $I_{ti}^L$, and
-# those with overseas acquired cases $I_{ti}^O$, with corresponding reproduction
-# rates $R_{ti}^L$ and $R_{ti}^O$:
+# targetted at each group: those with locally-acquired cases $I_{t,i}^L$, and
+# those with overseas acquired cases $I_{t,i}^O$, with corresponding reproduction
+# rates $R_{t,i}^L$ and $R_{t,i}^O$:
 
-#   lambda_{ti} &= I_{ti}^L R_{ti}^L + I_{ti}^O R_{ti}^O
-#   R_{ti}^L &= R_0 D_t e^{\epsilon_{ti}^L}
-#   R_{ti}^O &= R_0 Q_t e^{\epsilon_{ti}^O}
+#   lambda_{t,i} &= I_{t,i}^L R_{t,i}^L + I_{t,i}^O R_{t,i}^O
+#   R_{t,i}^L &= R_0 D_t e^{\epsilon_{t,i}^L}
+#   R_{t,i}^O &= R_0 Q_t e^{\epsilon_{t,i}^O}
 
 # where each of these reproduction rates is modelled as a product of: the
 # reproduction rate under initial conditions and no interventions $R_0$;
 # deterministic functions $D_t$ and $Q_t$ that modify $R_0$ over time to
[…]

### 2020-05-03 R/R_effective.R
rejig R_effective plots

@@ -329,6 +329,22 @@ orange <- brewer.pal(8, "Set2")[2]
 pink <- brewer.pal(8, "Set2")[4]
 
 
+# david does 8.27 x 11.69 (landscape A4) for 3x2 panels
+# aspect ratio of 0.707:1 h:w
+# want A4 *portrait* width (8.27) with same aspect ratio
+
+# get required aspect ratio
+panel_width <- 11.69 / 2
+panel_height <- 8.27 / 3
+panel_ratio <- panel_height / panel_width
+
+# work out dimensions for 4x2 panels for reports
+multi_mfrow <- c(4, 2)
+multi_width <- 8.27
+multi_height <- (multi_width / multi_mfrow[2]) * panel_ratio * multi_mfrow[1]
+# add a bit of space for the title
+multi_height <- multi_height * 1.2
+
 # overall R_eff
 plot_trend(R_eff_sim,
            multistate = TRUE,
@@ -338,8 +354,9 @@ plot_trend(R_eff_sim,
   ylab(expression(R["eff"]))
 
 ggsave("outputs/figures/R_eff_all.png",
-       width = 10,
-       height = 16, scale = 0.8)
+       width = multi_width,
+       height = multi_height,
+       scale = 0.8)
 
 # local R_eff
 plot_trend(R_eff_loc_sim,
@@ -350,8 +367,9 @@ plot_trend(R_eff_loc_sim,
   ylab(expression(R["eff"]~of~"locally-acquired"~cases))
 
 ggsave("outputs/figures/R_eff_local.png",
-       width = 10,
-       height = 16, scale = 0.8)
+       width = multi_width,
+       height = multi_height,
+       scale = 0.8)
 
 # imported R_eff
 plot_trend(R_eff_imp_sim,
@@ -362,8 +380,10 @@ plot_trend(R_eff_imp_sim,
   ylab(expression(R["eff"]~of~"overseas-acquired"~cases))
 
 ggsave("outputs/figures/R_eff_imported.png",
-       width = 10,
-       height = 16, scale = 0.8)
+       width = multi_width,
+       height = multi_height,
+       scale = 0.8)
+
 
 # local average R_eff 
 plot_trend(R_eff_loc_trend_sim,
@@ -371,25 +391,27 @@ plot_trend(R_eff_loc_trend_sim,
            base_colour = green,
            vline_at = intervention_dates()$date) + 
   ggtitle(label = "Impact of social distancing",
[…]

### 2020-05-03 R/functions.R
update latent factor model

@@ -92,7 +92,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2006HotfixDev20/v2/en-us/applemobilitytrends-2020-04-27.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev44/v2/en-us/applemobilitytrends-2020-04-30.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -132,7 +132,7 @@ apple_mobility <- function() {
 citymapper_mobility <- function() {
   
   # get link from: https://citymapper.com/cmi/about
-  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200429.csv"
+  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200501.csv"
   data <- readr::read_csv(url, skip = 3) %>%
     tidyr::pivot_longer(cols = -Date,
                         names_to = "region",
@@ -464,7 +464,7 @@ summarise_vec_posterior <- function(vector,
                                     draws,
                                     quantiles = c(0.025, 0.975),
                                     nsim = 1000) {
-  vector_sim <- posterior_sims(vector, draws, nsim = nsim)[, , 1]
+  vector_sim <- posterior_sims(vector, draws, nsim = nsim)
   posterior_mean <- colMeans(vector_sim)
   posterior_ci <- t(apply(vector_sim, 2, quantile, quantiles))
   cbind(mean = posterior_mean, posterior_ci)

### 2020-05-03 R/latent_social_distancing.R

@@ -61,12 +61,17 @@ distancing <- latent_behaviour_switch(date_num, trigger_date_num)
 # switching back to baseline behaviour or increasing distancing behaviour) Fix
 # the parameters to an exponential function, since it's barely identified.
 last_date_num <- n_dates - 1
-kappa_distancing_change <- 2
-tau_distancing_change <- last_date_num
-distancing_change <- latent_behaviour_switch(date_num,
-                                             tau = tau_distancing_change,
-                                             kappa = kappa_distancing_change)
-distancing_change <- distancing_change * 2
+# kappa_distancing_effect <- 2
+# tau_distancing_effect <- last_date_num - 7
+# distancing_effect <- latent_behaviour_switch(date_num,
+#                                              tau = tau_distancing_effect,
+#                                              kappa = kappa_distancing_effect)
+# distancing_effect <- distancing_effect * 2
+
+# make this a linear increase since one week after the last intervention
+# should be able to fit to date once we get more datastreams showing waning
+effect_period <- as.numeric(max(dates) - max(interventions$date)) - 7
+distancing_effect <- pmax(0, 1 + (date_num - last_date_num) / effect_period)
 
 # latent factor for pre-distancing surge in mobility with a prior that it peaks
 # around the time of the first restriction
@@ -95,7 +100,7 @@ weekday <- day_weights[doy]
 # combine into latent factor matrix
 latents_ntnl <- cbind(bump,
                       distancing,
-                      distancing_change,
+                      distancing_effect,
                       back_to_work,
                       weekday)
 n_latents_ntnl <- ncol(latents_ntnl)
@@ -288,7 +293,7 @@ for (j in seq_len(n_states)) {
 # same for a subset of datastreams, by all states
 target_datastreams <- c("Apple: directions for driving",
                         "Google: time at residential",
-                        "Google: time at grocery and pharmacy")
+                        "Google: time at parks")
 
 # get the vector of state-datastreams to plot
 state_datastreams_plot <- mobility %>%
@@ -589,6 +594,8 @@ for(i in 1:4) {
     mutate(date = dates) %>%
     saveRDS(filename)
 }
+
+save.image("outputs/latent_social_distancing_temp.RData")
 
 # - plot state-by-factor loading plots
 # - pull more model code out into functions

### 2020-05-04 manuscript/ms.Rmd

@@
 Australia --- and many other countries with intensive physical distancing measures in place --- is now shifting its focus from reacting to a first epidemic wave of COVID-19 to proactive planning for gradual and targeted relaxation of physical distancing measures.
-Quantifying how the behaviour of the population has changed in response to these measures and the potential impact on disease transmission is critical for the next stage of response planning.
+Quantifying how the behaviour of the population has changed in response to these measures and the potential impact on disease transmission is critical for the next stage of epidemic planning and response.
 

### 2020-05-04 R/R_effective.R
include social distancing in R_eff in more sensible way; use more stable version of log(1 - x)

@@ -192,11 +192,13 @@ log_Qt_vec <- log_q[q_index_vec]
 # log_Qt <- log_q[q_index]
 
 # the reduction from R0 down to R_eff for locally-acquired cases due to social
-# distancing behaviour, modelled as a being proportional to the change in
-# mobility measured from the mobility datastreams
-beta <- normal(0, 1)
-log_Dt_vec <- beta * rep(social_distancing_index, n_states)
-# log_Dt <- beta * social_distancing_index
+# distancing behaviour, modelled as a being proportional to the reduction in
+# mobility (and therefore contacts) measured from the mobility datastreams. The
+# social distanding index is proportional to the percentage change in mobility,
+# but we can express is as proportional reduction in mobility with: 1 - beta * sdi
+beta <- uniform(0, 1)
+log_Dt_vec <- log1p(-beta * rep(social_distancing_index, n_states))
+# log_Dt <- log1p(-beta * social_distancing_index)
 
 # temporally correlated errors in R_eff for local cases - representing all the
 # stochastic transmission dynamics in the community, such as outbreaks in
@@ -289,7 +291,7 @@ min(n_eff)
 # infectious cases that are imports
 
 p_imported <- proportion_imported(local_infectious, imported_infectious, X)
-log_Dt_weighted <- log_Dt_vec + epsilon_L + log(1 - c(p_imported))
+log_Dt_weighted <- log_Dt_vec + epsilon_L + log1p(-c(p_imported))
 log_Qt_weighted <- log_Qt_vec + epsilon_O + log(c(p_imported))
 log_rel_R_eff <- log(exp(log_Dt_weighted) + exp(log_Qt_weighted))
 log_R_eff <- log_R0 + log_rel_R_eff

### 2020-05-04 R/functions.R

@@ -772,38 +772,6 @@ lognormal_prior <- function(mean, sd) {
   
 }
 
-tf_as_float <- greta:::tf_as_float
-fl <- greta:::fl
-
-greta_kernel <- greta.gp:::greta_kernel
-tf_cols <- greta.gp:::tf_cols
-tf_distance <- greta.gp:::tf_distance
-check_active_dims <- greta.gp:::check_active_dims
-
-
-iid <- function(variance, columns = 1) {
-  greta_kernel("iid",
-               tf_name = "tf_iid",
-               parameters = list(variance = variance),
-               arguments = list(
-                 active_dims = check_active_dims(columns,
-                                                 rep(1, length(columns))
-                 )))
-}
-
-tf_iid <- function(X, X_prime, variance, active_dims) {
-  
-  # pull out active dimensions
-  X <- tf_cols(X, active_dims)
-  X_prime <- tf_cols(X_prime, active_dims)
-  
-  # find where these values match and assign the variance as a covariance there
-  # (else set it to 0)
-  distance <- tf_distance(X, X_prime)
-  tf_as_float(distance < fl(1e-12)) * variance
-
-}
-
 R0_prior <- function() {
   # Visually estimated values of the posterior distributions  (means and lower
   # and upper 95% CIs) over R0 in 11 European countries from Flaxman et al.

### 2020-05-05 R/functions.R
update data; use state-level Apple driving data; estimate peak of distancing

@@ -67,7 +67,7 @@ facebook_mobility <- function() {
 # - remove the grocery and pharmacy category
 # (affected by panic buying, not interventions)
 google_mobility <- function() {
-  # get link from: https://www.google.com/covid19/mobility/index.html?hl=en
+  # get link from: https://www.google.com/covid19/mobility/index.html
   url <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
   data <- readr::read_csv(url, ) %>%
     filter(country_region == "Australia") %>%
@@ -92,7 +92,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev44/v2/en-us/applemobilitytrends-2020-04-30.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev47/v2/en-us/applemobilitytrends-2020-05-03.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -112,17 +112,38 @@ apple_mobility <- function() {
   data <- data %>%
     mutate(trend = trend - 100,
            date = date + 1)
+
+  # pull out the states and cities
+  states <- c("New South Wales", "Victoria", "Queensland", "Western Australia", 
+              "South Australia", "Australian Capital Territory", "Tasmania", 
+              "Northern Territory")
   
-  # add on a state label, rename the transportation type to 'category' to match
-  # google, and add on the data source
+  cities <- c("Sydney", "Melbourne", "Brisbane", "Perth", 
+              "Adelaide", "Canberra", "Hobart", "Darwin")
+  
+  # need to use state-level data for driving, but cities for other metrics
   data <- data %>%
+    filter(
+      case_when(
+        transportation_type == "driving" & region %in% states ~ TRUE,
+        transportation_type != "driving" & region %in% cities ~ TRUE,
+        TRUE ~ FALSE
+      )
+    ) %>%
     mutate(
-      state = case_when(region == "Sydney" ~ "New South Wales",
-                        region == "Melbourne" ~ "Victoria",
-                        region == "Brisbane" ~ "Queensland",
[…]

### 2020-05-05 R/latent_social_distancing.R

@@ -58,20 +58,22 @@ trigger_date_num <- as.numeric(interventions$date - first_date)
 distancing <- latent_behaviour_switch(date_num, trigger_date_num)
 
 # add term for recent change to social distancing (some proportion either
-# switching back to baseline behaviour or increasing distancing behaviour) Fix
-# the parameters to an exponential function, since it's barely identified.
+# switching back to baseline behaviour or increasing distancing behaviour)
+# make this a linear increase since a peak (with prior mean one week after the
+# last intervention, 95% interval in the week around that)
 last_date_num <- n_dates - 1
-# kappa_distancing_effect <- 2
-# tau_distancing_effect <- last_date_num - 7
-# distancing_effect <- latent_behaviour_switch(date_num,
-#                                              tau = tau_distancing_effect,
-#                                              kappa = kappa_distancing_effect)
-# distancing_effect <- distancing_effect * 2
+last_intervention_date_num <- as.numeric(max(interventions$date) - first_date)
+peak_range <- last_date_num - last_intervention_date_num
 
-# make this a linear increase since one week after the last intervention
-# should be able to fit to date once we get more datastreams showing waning
-effect_period <- as.numeric(max(dates) - max(interventions$date)) - 7
-distancing_effect <- pmax(0, 1 + (date_num - last_date_num) / effect_period)
+# how far back from the last date was the peak of distancing?
+peak <- normal(last_intervention_date_num + 7,
+               3.5 / 1.96,
+               truncation = c(last_intervention_date_num, last_date_num))
+effect_period <- last_date_num - peak
+
+distancing_effect <- 1 + ((date_num - last_date_num) / effect_period)
+nullify <- (sign(distancing_effect) + 1) / 2
+distancing_effect <- distancing_effect * nullify
 
 # latent factor for pre-distancing surge in mobility with a prior that it peaks
 # around the time of the first restriction
@@ -174,7 +176,7 @@ m <- model(loadings_ntnl, loadings_holiday)
 draws <- mcmc(m,
               sampler = hmc(Lmin = 20, Lmax = 25),
               chains = 20)
-draws <- extra_samples(draws, 4000)
+draws <- extra_samples(draws, 1000)
 
 # check convergence
 r_hats <- coda::gelman.diag(draws, autoburnin = FALSE, multivariate = FALSE)$psrf[, 1]
@@ -291,9 +293,9 @@ for (j in seq_len(n_states)) {
 }
 
[…]

### 2020-05-05 R/R_effective.R
account for waning distancing in R_eff model

@@ -153,16 +153,29 @@ local_cases <- date_by_state %>%
   select(-import_status) %>%
   as.matrix()
 
-# load the social distancing factor
+library(greta.gp)
+
+# build the social distancing index, with fixed shape for main effect, and informative prior on the amount of waning
 distancing_file <- "outputs/social_distancing_latent.RDS"
-social_distancing_index <- distancing_file %>%
+social_distancing_main_index <- distancing_file %>%
+  readRDS() %>%
+  select(mean, date) %>%
+  right_join(tibble(date = dates)) %>%
+  replace_na(list(mean = 0)) %>%
+  pull(mean)
+
+waning_file <- "outputs/waning_distancing_latent.RDS"
+waning_index <- waning_file %>%
   readRDS() %>%
   select(mean, date) %>%
   right_join(tibble(date = dates)) %>%
   replace_na(list(mean = 0)) %>%
   pull(mean)
   
-library(greta.gp)
+waning_amount_params <- readRDS("outputs/waning_amount_parameters.RDS")
+waning_amount <- do.call(normal, waning_amount_params)
+
+social_distancing_index <- social_distancing_main_index + waning_index * waning_amount
 
 # lognormal prior on R0 (normal prior on log(R0)) based on estimates for
 # Northern Europe from Flaxman et al. (Imperial report 13, lowest R0s from their

### 2020-05-05 R/latent_social_distancing.R

@@ -608,11 +608,77 @@ distancing_change_data %>%
     max = min(value)
   )
 
-vals <- calculate(peak, values = draws, nsim = 20000)[[1]][, 1, 1]
-peak_date <- first_date + vals
-first_date + mean(vals)
-first_date + quantile(vals, c(0.025, 0.975))
+peak_draws <- calculate(peak, values = draws, nsim = 20000)[[1]][, 1, 1]
+peak_mean <- first_date + mean(peak_draws)
+peak_ci <- first_date + quantile(peak_draws, c(0.025, 0.975))
 
+
+# get an index of waned distancing, weighting datastreams according to their
+# representativeness of transmission potential ('riskiness')
+
+transmission_potential <- state_datastream_lookup %>%
+  mutate(risk = case_when(
+    datastream %in% c(
+      "Apple: directions for driving",
+      "Apple: directions for walking",
+      "Google: time at parks",
+      "Google: time at residential"
+    ) ~ "low",
+    datastream %in% c(
+      "Google: time at grocery and pharmacy",
+      "Google: time at retail and recreation"
+    ) ~ "medium",
+    datastream %in% c(
+      "Citymapper: directions",
+      "Apple: directions for transit",
+      "Google: time at transit stations",
+      "Google: time at workplaces"
+    ) ~ "high",
+  )) %>%
+  group_by(risk) %>%
+  mutate(weight = case_when(
+    risk == "low" ~ 0.1,
+    risk == "medium" ~ 0.3,
+    risk == "high" ~ 0.6
+  )) %>%
+  mutate(weight = weight / n_distinct(state_datastream))
+
+# calculate weighted regression of the overall distancing effect, weighted to
+# represent mobility data streams thought to be most representative of
+# transmission pontential
+regression <- distancing_change_relative %*% as_data(transmission_potential$weight)
+regression_vals <- calculate(regression, values = draws, nsim = 10000)[[1]]
+
+# save parameters of posterior on the amount of waning, for use in R_eff model
+waning_amount_params <- list(
+  mean = mean(regression_vals),
+  sd = sd(regression_vals),
+  truncation = c(-1, 0)
+)
+saveRDS(waning_amount_params,
+        file = "outputs/waning_amount_parameters.RDS")
+
[…]

### 2020-05-05 R/R_effective.R
make R_effective model fitting and posterior prediction more efficient

@@ -195,14 +195,14 @@ q_index <- case_when(
   TRUE ~ 3,
 )
 
-q_index_vec <- rep(q_index, n_states)
+# q_index_vec <- rep(q_index, n_states)
 # q_raw <- uniform(0, 1, dim = 3)
 # q <- cumprod(q_raw)
 # log_q <- log(q)
 log_q_raw <- -exponential(1, dim = 3)
 log_q <- cumsum(log_q_raw)
-log_Qt_vec <- log_q[q_index_vec]
-# log_Qt <- log_q[q_index]
+# log_Qt_vec <- log_q[q_index_vec]
+log_Qt <- log_q[q_index]
 
 # the reduction from R0 down to R_eff for locally-acquired cases due to social
 # distancing behaviour, modelled as a being proportional to the reduction in
@@ -210,8 +210,8 @@ log_Qt_vec <- log_q[q_index_vec]
 # social distanding index is proportional to the percentage change in mobility,
 # but we can express is as proportional reduction in mobility with: 1 - beta * sdi
 beta <- uniform(0, 1)
-log_Dt_vec <- log1p(-beta * rep(social_distancing_index, n_states))
-# log_Dt <- log1p(-beta * social_distancing_index)
+# log_Dt_vec <- log1p(-beta * rep(social_distancing_index, n_states))
+log_Dt <- log1p(-beta * social_distancing_index)
 
 # temporally correlated errors in R_eff for local cases - representing all the
 # stochastic transmission dynamics in the community, such as outbreaks in
@@ -252,19 +252,19 @@ noise_kernel_O <- white(sigma_noise_O ^ 2)
 kernel_L <- noise_kernel_L + time_kernel_L * state_variation_kernel_L
 kernel_O <- noise_kernel_O + time_kernel_O * state_variation_kernel_O
 
-# data for the GP to act on
-X <- cbind(date = rep(date_nums, n_states),
-           state = rep(seq_len(n_states), each = n_dates))
+# # data for the GP to act on
+# X <- cbind(date = rep(date_nums, n_states),
+#            state = rep(seq_len(n_states), each = n_dates))
 
 # build a matrix of inducing points, regularly spaced over time but with one on
 # the most recent date
[…]

### 2020-05-05 R/functions.R

@@ -931,7 +931,7 @@ plot_trend <- function(simulations,
 
 
 # interpolate proportion of infectious cases that are imports
-proportion_imported <- function (local_infectious, imported_infectious, X) {
+proportion_imported <- function (local_infectious, imported_infectious) {
   
   all_infectious <- local_infectious + imported_infectious
   prop_imported <- imported_infectious / all_infectious
@@ -941,9 +941,15 @@ proportion_imported <- function (local_infectious, imported_infectious, X) {
   prop_imported <- pmin(0.99, prop_imported)
   q_imported <- c(qlogis(prop_imported))
   
+  n_dates <- nrow(all_infectious)
+  n_states <- ncol(all_infectious)
+  date_nums <- seq_len(n_dates)
+  states <- factor(seq_len(n_states))
   # interpolate state-by-state  
-  df <- as.data.frame(X)
-  df$state <- factor(df$state)
+  df <- data.frame(
+    date = rep(date_nums, n_states),
+    state = rep(states, each = n_dates)
+  )
   model <- mgcv::gam(q_imported ~ s(date) +
                        s(date, by = state, k = 30),
                      data = df)
@@ -951,10 +957,10 @@ proportion_imported <- function (local_infectious, imported_infectious, X) {
   
   # transform back, clamp to numerically stable values, and return
   p_imported <- plogis(q_imported)
-  dim(p_imported) <- dim(all_infectious)
   eps <- sqrt(.Machine$double.eps)
   p_imported <- pmax(eps, p_imported)
   p_imported <- pmin(1 - eps, p_imported)
+  dim(p_imported) <- dim(all_infectious)
   p_imported
   
 }

### 2020-05-06 R/delay_variation.R
estimate change in time to detection of cases

@@ -1,0 +1,146 @@
+# model reporting delays in the linelist to detect impact of the app
+
+library(dplyr)
+library(readr)
+library(tidyr)
+library(RColorBrewer)
+source("R/functions.R")
+
+set.seed(2020-04-29)
+linelist <- readRDS("~/not_synced/nnds/linelist_formatted.RDS")
+
+# Delay is time from date of symptom onset to specimen collection, only consider
+# it valid if 0 or greater. There are a very few, data points in the 2-3 days
+# before symptom onset (when virus detectable) and a number of clearly erroneous
+# ones before this. These are so few that they a) are clearyly not from the same
+# distribution as the positive delays, and b) are so few (<1.5%) that their
+# omission should not substantially bias the estimate of mean time to detection.
+delay_data <- linelist %>%
+  filter(import_status == "local") %>%
+  mutate(date_infection = date_onset - 5) %>%
+  filter(!is.na(date_infection),
+         !is.na(date_detection)) %>%
+  mutate(time_to_detection = as.numeric(date_detection - date_onset),
+         date_num = date_infection - min(date_infection) + 1,
+         date_num = as.numeric(date_num)) %>%
+  filter(time_to_detection >= 0) %>%
+  select(date_infection,
+         date_num,
+         delay = time_to_detection,
+         state = region)
+
+
+n_dates <- max(delay_data$date_num)
+date_nums <- seq_len(n_dates)
+states <- sort(unique(delay_data$state))
+n_states <- length(states)
+
+# fit a hierarchical gp
+library(greta.gp)
+
+kernel <- function(alpha_time = lognormal(2, 0.5),
+                   lengthscale_time = lognormal(4, 0.5),
+                   sigma_time = normal(0, 0.5, truncation = c(0, Inf)),
+                   sigma_bias = NULL) {
+  
+  kernel <- rational_quadratic(
+    lengthscales = lengthscale_time,
+    variance = sigma_time ^ 2,
+    alpha = alpha_time)
+  
+  
+  if (!is.null(sigma_bias)) {
+    bias <- bias(sigma_bias ^ 2)
+    kernel <- kernel + bias
+  }
+  
+  kernel
+  
+  
+}
+
+state_kernel <- kernel()
+national_kernel <- kernel(sigma_bias = 0.1)
+
[…]

### 2020-05-06 R/functions.R

@@ -964,3 +964,25 @@ proportion_imported <- function (local_infectious, imported_infectious) {
   p_imported
   
 }
+
+app_downloads <- function () {
+  tibble::tribble(
+    ~date, ~number,
+    "04-25", 0,
+    "04-26", 1113000,
+    "04-27", 2e6,
+    "04-28", 2440000,
+    "04-29", 3e6,
+    "04-30", 3310000,
+    "05-01", 3500000,
+    "05-02", 4000000,
+    "05-03", 4250000,
+    "05-05", 4950000,
+    "05-06", 5100000
+  ) %>%
+    mutate(
+      date = paste0("2020-", date),
+      date = as.Date(date),
+      uptake = number / 16e6
+    )
+}

### 2020-05-06 R/delay_variation.R
estimate change in R_eff due to surveillance

@@ -144,3 +144,64 @@ ggsave("outputs/figures/time_to_detection_fit.png",
 
 pred_df %>%
   filter(date_num == max(date_num))
+
+# priors for the parameters of the lognormal distribution over the serial interval from Nishiura et
+# al., as stored in the EpiNow source code 
+si_param_samples <- read_csv(
+  file = "https://raw.githubusercontent.com/epiforecasts/EpiNow/758b706599244a545d6b07f7be4c10ffe6c8cf50/data-raw/nishiura-lognormal-truncated.csv",
+  col_types = cols(param1 = col_double(),
+                   param2 = col_double())
+)
+meanlog <- mean(si_param_samples$param1)
+sdlog <- mean(si_param_samples$param2)
+# beta <- function(x) {
+#   plnorm(x + 5, meanlog, sdlog)
+# }
+# gamma <- function(x) {
+#   beta(x) / beta(10)
+# }
+# # 
+# pred_df$prop_R_eff <- gamma(pmin(10, pred_df$delay))
+# par(mfrow = c(4, 2))
+# for (state in states) {
+#   plot(prop_R_eff ~ date, data = pred_df[pred_df$state == state, ],
+#        type = "l",
+#        ylim = c(0, 1))
+#   title(main = state)
+# }
+
+# need to incorporate uncertainty on earlier delay here, so the later delay
+# becomes uncertain
+
+# so need to compute lognormal cdf in greta
+beta_dt <- iprobit((log(mu + 5) - meanlog) / sdlog)
+gamma_dt <- sweep(beta_dt, 2, t(beta_dt[1, ]), FUN = "/")
+gamma_dt_sim <- calculate(gamma_dt, values = draws, nsim = 2000)[[1]]
+gamma_dt_mean <- c(apply(gamma_dt_sim, 2:3, FUN = base::mean))
+gamma_dt_lower <- c(apply(gamma_dt_sim, 2:3, FUN = quantile, 0.025))
+gamma_dt_upper <- c(apply(gamma_dt_sim, 2:3, FUN = quantile, 0.975))
+
+pred_df$rel_reff_mean <- gamma_dt_mean
+pred_df$rel_reff_lower <- gamma_dt_lower
+pred_df$rel_reff_upper <- gamma_dt_upper
+  
+ggplot(pred_df) +
+  aes(date, rel_reff_mean) +
+  
[…]

### 2020-05-13 R/functions.R
update mobility links

@@ -94,7 +94,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev54/v2/en-us/applemobilitytrends-2020-05-09.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev55/v2/en-us/applemobilitytrends-2020-05-10.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -155,7 +155,7 @@ apple_mobility <- function() {
 citymapper_mobility <- function() {
   
   # get link from: https://citymapper.com/cmi/about
-  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200510.csv"
+  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200511.csv"
   data <- readr::read_csv(url, skip = 3) %>%
     tidyr::pivot_longer(cols = -Date,
                         names_to = "region",

### 2020-05-13 R/latent_social_distancing.R

@@ -599,6 +599,7 @@ for(i in 1:4) {
 
 save.image("outputs/latent_social_distancing_temp.RData")
 
+# summarise state-level variation in waning
 distancing_change_data %>%
   mutate(value = pmin(0, value * 100)) %>%
   group_by(datastream) %>%
@@ -607,6 +608,12 @@ distancing_change_data %>%
     min = max(value),
     max = min(value)
   )
+
+distancing_change_data %>%
+  mutate(value = pmin(0, value * 100)) %>%
+  filter(datastream %in% target_datastreams) %>%
+  arrange(datastream, state) %>%
+  as.data.frame()
 
 peak_draws <- calculate(peak, values = draws, nsim = 20000)[[1]][, 1, 1]
 peak_mean <- first_date + mean(peak_draws)
@@ -675,10 +682,6 @@ saveRDS(waning_amount_params,
 # lines(distancing_overall_est[, 2] ~ dates, lty = 2)
 # lines(distancing_overall_est[, 3] ~ dates, lty = 2)
 # lines(distancing_overall_est[, 1] ~ dates, lwd = 3)
-# 
-# # 0.1, 0.1, 0.8: 10.1% (6.6% - 13.3%) 
-# # 0.1, 0.3, 0.6: 10.7% (5.0% - 15.6%) 
-# # 0.2, 0.4, 0.4: 12.5% (5.3% - 18.6%) 
 
 
 # - plot state-by-factor loading plots

### 2020-05-13 R/simulate_case_counts.R
add scrappy stochastic simulation of cases

@@ -1,0 +1,178 @@
+# use a discrete-time convolution with/without Poisson sampling to forecast from the Reff model
+library(dplyr)
+library(tidyr)
+source("R/functions.R")
+
+linelist <- readRDS("~/not_synced/nnds/linelist_formatted.RDS")
+
+# for now just do single imputation with the mean delay on the date of onset for
+# those missing it
+delay_samples <- read_csv(
+  file = "data/cases/sampled_report_delay.csv",
+  col_types = cols(x = col_integer())
+)
+mean_delay <- round(mean(delay_samples$x))
+
+linelist <- linelist %>%
+  mutate(
+    date_onset = case_when(
+      is.na(date_onset) ~ date_confirmation - mean_delay,
+      TRUE ~date_onset
+    )
+  )
+
+# build date-by-state matrices of the counts of new local and imported cases and
+# imports by assumed date of infection (with an incubation period of 5 days)
+linelist <- linelist %>%
+  rename(state = region,
+         date = date_onset) %>%
+  mutate(date = date - 5) %>%
+  select(-date_confirmation)
+
+import_statuses <- sort(unique(linelist$import_status))
+states <- sort(unique(linelist$state))
+dates <- seq(min(linelist$date), max(linelist$date), by = 1)
+
+n_states <- length(states)
+n_dates <- length(dates)
+n_extra <- 21
+date_nums <- seq_len(n_dates + n_extra)
+
+# pad this with full set of dates, states, and import statuses
+grid <- expand_grid(
+  date = dates,
+  import_status = import_statuses,
+  state = states)
+
+# widen into matrices of date by state
+date_by_state <- linelist %>%
+  mutate(cases = 1) %>%
+  right_join(grid) %>%
+  group_by(import_status, state, date) %>%
+  summarise(cases = sum(cases, na.rm = TRUE)) %>%
+  ungroup() %>%
+  pivot_wider(names_from = state, values_from = cases) %>%
+  select(-date)
+
+imported_cases <- date_by_state %>%
+  filter(import_status == "imported") %>%
+  select(-import_status) %>%
+  as.matrix()
+
+local_cases <- date_by_state %>%
+  filter(import_status == "local") %>%
+  select(-import_status) %>%
+  as.matrix()
+
+scenarios <- c("projection",
+               "counterfactual_1",
[…]

### 2020-05-13 R/functions.R
add functions for LSHTM's generation interval distribution

@@ -563,6 +563,10 @@ plot_latent_factor <- function (factor, draws, dates, key_dates, cols = grey(c(0
         col.main = grey(0.3))
 }
 
+gamma_cdf <- function(x, shape, rate) {
+  stop("not yet implemented")
+}
+
 # greta function for the lognormal CDF (equivalent to plnorm(x, meanlog, sdlog))
 # x must not be a greta array, though meanlog and sdlog can be. A greta array of
 # CDF values is returned, equal to 0 wherex was 0 or lower.
@@ -611,7 +615,55 @@ parameter <- function(x, truncation = c(-Inf, Inf)) {
   normal(mean(x), sd(x), truncation = truncation)
 }
 
-# given a positive integer vector of (days since a case became symptomatic,
+# given a positive integer vector of days (since an infection began), compute
+# the probability that the generation interval values falls in that day
+generation_interval_probability <- function(days, fixed = FALSE) {
+  
+  # priors for the parameters of the lognormal distribution over the serial interval from Nishiura et
+  # al., as stored in the EpiNow source code 
+  
+  download.file(
+    "https://github.com/epiforecasts/EpiNow/raw/master/data-raw/gi.rds",
+    (f <- tempfile()),
+    quiet = TRUE
+  )
+  gi_param_samples <- readRDS(f)
+  
+  # compute the CDF of the lognormal distribution, for this and the next day
+  days_lower <- days
+  days_upper <- days + 1
+  
+  if (fixed) {
+    
+    # if fixed, we can do this in R
+    mean <- mean(gi_param_samples$mean)
+    sd <- mean(gi_param_samples$sd)
+    scale <- sd ^ 2 / mean
+    shape <- mean / scale
+    
+    p_lower <- pgamma(days_lower, shape = shape, scale = scale)
+    p_upper <- pgamma(days_upper, shape = shape, scale = scale)
+    
+  } else {
+    # otherwise create greta arrays for the parameters, with priors based on
+    # these samples
+    scale_draws <- gi_param_samples$sd ^ 2 / gi_param_samples$mean
+    shape_draws <- gi_param_samples$mean / scale_draws
+    rate_draws <- 1 / scale_draws
+    
+    rate <- parameter(rate_draws, c(0, Inf))
+    shape <- parameter(shape_draws, c(0, Inf))
+    
+    p_lower <- gamma_cdf(days_lower, shape, rate)
[…]

### 2020-05-14 manuscript/ms.Rmd
add Reff model description to the manuscript

@@
 ---
-title: "Quantifying social distancing behaviour across Australia in response to COVID-19"
+title: "The impacts of public health policy measures on reducing SARS-CoV-2 transmission in Australia"
 author:
@@
 ## Gist
+
+NOTE: this manuscript was initially focussed solely on quantifying the reduction in mobility in Australia in response to social distancing policies, but the focus is now on quantifying the impact of this on transmission, and comparin with the impacts of other measures, including quarantining of overseas imports and improved surveillance.
+However the manuscript has not yet been updated to relfect this.
 
@@
 ## Methods
+
+## Modelling changes in transmission rate
+
+We developed a semi-mechanistic Bayesian statistical model to estimate $R_t$, the effective rate of transmission of of SARS-CoV-2 over time, whilst quantifying the impacts on $R_t$ of a range of policy measures introcudced at national and regional levels in Australia.
+
+### Observation model
+
+A straightforward observation model to relate case counts to the rate of transmission is to assume that the number of new locally-acquired cases $N_{t,i}^L$ at time $t$ in region $i$ is (conditionally on its expectation) Poisson-distributed with mean $lambda_{t,i}$ given by the product of the total infectiousness of infected individuals $I_{t,i}$ and the time-varying reproduction rate $R_{t,i}$:
+
+\begin{align}
+  N_{t,i}^L &\sim Poisson(\lambda_{t,i}) \\
+  \lambda_{t,i} &= I_{t,i} R_{t,i} \\
+  I_{t,i} &= \sum_{t'=1}^t g(t') N_{t',i} \\
+  N_{t',i} &= N_{t,i}^L + N_{t,i}^O
+\end{align}
+
+where the total infectiousness is the sum of all active infections $N_{t,i}^L$ - both locally-acquired and overseas-acquired $N_{t',i}^O$ - initiated at times $t'$ prior to $t$, each weighted by an infectivity function $g(t')$ giving the proportion of new infections that occur $t'$ days post-infection.
+$g(t')$ is equivalent to a discretised probability density function of the generation interval - the probability that the generation interval for an infector-infectee pair is $t'$ days.
+
+This observation model forms the basis of the maximum-likelihood method proposed by White and Pagano (2007) and the variations of that method by Cori et al.
+(2013), and Abbott et al.
+(2020) that have previously been used to estimate time-varying SARS-CoV-2 reproduction numbers in Australia.
+
+We extend this model to consider separate reproduction rates for two groups of infectious cases, in order to model the effects of different interventions targetted at each group: those with locally-acquired cases $I_{t,i}^L$, and those with overseas acquired cases $I_{t,i}^O$, with corresponding reproduction rates $R_{t,i}^L$ and $R_{t,i}^O$.
+These respectively are the rates of transmission from imported cases to locals, and from locally-acquired cases to locals:
+
+\begin{align}
+  N_{t,i}^L &\sim Poisson(\lambda_{t,i}) \\
+  \lambda_{t,i} &= I_{t,i}^L R_{t,i}^L + I_{t,i}^O R_{t,i}^O \\
+  I_{t,i}^L &= \sum_{t'=1}^t g(t') N_{t',i}^L \\
+  I_{t,i}^O &= \sum_{t'=1}^t g(t') N_{t',i}^O
+\end{align}
+
+Note that if data were available on the whether the source of infection for each locally-acquired case was another locally-acquired case or an overseas-acquired cases, we could split this into two separate analyses using the observation model above; one for each transmission source.
+In the absence of such data, the fractions of all transmission attributed to sources of each type is implicitly inferred by the model, with an associated increase in parameter uncertainty.
+
+### Reproduction rate models
+
+We model the reproduction rates of the two subpopulations in a semi-mechanistic way:
+
+\begin{align}
+  R_{t,i}^L &= R_0 D_t e^{\epsilon_{t,i}^L} \\
+  R_{t,i}^O &= R_0 Q_t e^{\epsilon_{t,i}^O}
+\end{align}
+
+with each reproduction rate is modelled as a product of: the reproduction rate under initial conditions and no interventions $R_0$; deterministic functions $D_t$ and $Q_t$ that modify $R_0$ over time to respectively represent the impacts of physical distancing and quarantine interventions at a national level, and correlated timeseries of random effects $\epsilon_{t,i}^L$ and $\epsilon_{t,i}^O$ to represent stochastic fluctuations in the reporting rate in each state.
+These error terms are themselves decomposed into population-wide changes in $R_{t,i}$ not captured by the impacts of the policy measures, and the more stochastic fluctuations in the reproduction rate among cases at each point in time - for example due to the clusters in subpopulations with higher or lower reproduction rates than the general
+population.
+
+### Intervention models
+
+We model the effect of $D_t$ as being proportional (on the log scale) to an index of the proportional change in population mobility in response to physical distancing measures $d_t$ - estimated from multiple datastreams of population mobility as described in the next section - which has initial value 0 before distancing measures were implemented and value 1 at its maximum extent:
+
+\begin{align}
+  D_t &= e^{\beta d_t}
+\end{align}
+
+We model $Q_t$ via a monotone decreasing step function with values constrained to the unit interval, and with steps at the known dates $\tau_1$ and $\tau_2$ of changes in quarantine policy:
+
+\begin{equation}
+ Q_t = \begin{cases}
+  q1 & t < \tau_1 \\
+  q2 & \tau_1 \leq t < \tau_2 \\
+  q3 & \tau_2 \leq t
+ \end{cases}
+\end{equation}
+
+where $q1 > q2 > q3$ and all parameters are constrained to the unit interval.
+
+### Error models
+
+The correlated timeseries of errors in the log of the effective reproduction rate for each group $\epsilon_{t,i}^L$ and $\epsilon_{t,i}^O$ are each modelled as a zero-mean Gaussian process (GP) with covariance structured to reflect temporal correlation in errors within each state (but independent between states).
+The covariance function (kernel) for each Gaussian process is constructed as the sum of covariance functions representing different components of error in the observed reproduction rates.
+Below we follow the notation of additive GP kernels that is more common in the field of probabilistic machine learning than in statistics when discussing correlated error terms.
+This corresponds to the summation of the covariance matrices returned by these functions and equally to the summation of GPs, each following these functions.
+
+The structure for error in locally-acquired cases is as follows:
+
+\begin{align}
+  \epsilon_i^L &\sim GP(\mathbf{0}, k^L(t, t')) \\
+  k^L(t, t') &= k_{SE}(t, t') + k_{RQ}(t, t') + k_{IID}^L(t, t') \\
+  k_{SE}(t, t') &= \sigma_1^2 \, exp \left( -\frac{(t - t')^2}{2 l_1^2} \right) \\
+  k_{RQ}(t, t') &= \sigma_2^2 \, exp \left(1 + \frac{(t - t')^2}{2 \alpha l_1^2} \right)^{-\alpha} \\
+  k_{IID}^L(t, t') &= \begin{cases}
+    \sigma_3^2 & t = t' \\ 
+    0 & t \neq t'
+ \end{cases}
+\end{align}
+
+where $k_{SE}()$ is a squared-exponential kernel encoding smooth, long-term deviations from the mean, $k_{RQ}()$ is a rational-quadratic kernel encoding shorter-term deviations from the mean with varying smoothness, and $k_{IID}()$ is an IID or white-noise kernel encoding independent deviations from the mean on each day.
+The three kernels are used to separately encode three different types of variation.
+The squared exponential kernel represents long-term trends in the population-wide transmission rate $R_{t,i}^L$ that are not captured by the effects of social distancing - such as improvements in surveillance, or behavioural changes not captured by changes in mobility data.
+This kernel also represents time-independent region-level differences in transmission rates.
+The rational quadratic kernel represents the short-term fluctuations in the reproduction rate that are due to large clusters of cases, with infections occurring over a period of days to weeks.
+The IID kernel represents clustering of cases on a single day - overdispersion in the observed numbers of new local infections per day relative to the assumption of independent events implicit in the Poisson observation model.
+This is equivalent to using a Poisson-lognormal observation distribution conditional on the timeseries component of the model; functionally equivalent to a negative binomial observation distribution.
+
+The structure for error in overseas-acquired cases is somewhat simpler:
+
+\begin{align}
+  \epsilon_i^O &\sim GP(\mathbf{0}, k^O(t, t')) \\
+  k^O(t, t') &= k_{bias}(t, t') + k_{IID}^O(t, t') \\
+  k_{bias}(t, t') &= \sigma_4^2 \\ 
+  k_{IID}^O(t, t') &= \begin{cases}
+    \sigma_5^2 & t = t' \\ 
+    0 & t \neq t'
+ \end{cases}
+\end{align}
+
+which consists of a bias kernel representing time-independent region-level differences in transmission rates, and an IID kernel representing daily overdispersion due to clustering in new cases acquired from individuals with overseas-acquired infections.
+Note that this GP formulation is equivalent to the following IID Gaussian random effects model:
+
+\begin{align}
+  \epsilon_{t,i}^O &\sim N(\mu_i, \sigma_5 ^ 2) \\
+  \mu_i &\sim N(0, \sigma_4 ^ 2)
+\end{align}
+
+### Components of the local reproduction rate
+
+We model the rate of transmision from locally acquired cases using the basic reproduction rate, the impact of social distancing, and an error term.
+However it is easier to consider these epidemiologically by decomposing the expression for $R_{t,i}^L$ into three different components:
+
+\begin{enumerate}
+\item the national trend in $R_t$ due to social distancing,
+\item the state-level trends in population-wide $R_{t,i}$, and
+\item the deviation in this statewide average transmission rate due to case clusters.
+\end{enumerate}
+
+Component 1 of the transmission rate for locally-acquired cases is simply given by $R_0D_t$.
+
+The latter two correspond to different sub-components of the correlated error term $\epsilon_{t,i}^L$: $\epsilon_{i}^{L,1}$ which comprises long-term variation in the region-wide transmission rate, characterised by $k_{SE}()$; and $\epsilon_{t,i}^{L,2}$, characterised by $k_{RQ}() + k_{IID}^L()$, which comprises short-term variation in the transmission rate among the infectious cases due to clusters of cases in subpopulations with transmission rates that are higher (or possibly lower) than the population average.
+The additive structure of the kernel enables use to perfom an additive decomposition of $\epsilon_{t,i}^L$ into these two terms.
+
+Component 2 is therefore given by $R_0D_te^{\epsilon_{i}^{L,1}}$, and Component 3 (which we consider on the log-scale for ease of visualisation) is $\epsilon_{t,i}^{L,2}$.
+
+## Modelling changes in population mobility
 

### 2020-05-14 R/R_effective.R

@@ -1,79 +1,8 @@
 # fit a Bayesian model-based estimate of R_effective over time, quantifying the
 # impacts of both quarantine and physical distancing measures.
 
-# The basic model assumes that the number of new locally-acquired cases
-# $N_{t,i}^L$ at time $t$ in region $i$ is *conditionally* Poisson-distributed
-# with expectation $lambda_{t,i}$ given by the product of the number of
-# infectious cases $I_{t,i}$ and the time-varying reproduction rate $R_{t,i}$:
-
-#   N_{t,i}^L &\sim Poisson(lambda_{t,i})
-#   lambda_{t,i} &= I_{t,i} R_{t,i}
-
-# We extend this model by considering separate reproduction rates for two groups
-# of infectious cases in order to model the effects of different interventions
-# targetted at each group: those with locally-acquired cases $I_{t,i}^L$, and
-# those with overseas acquired cases $I_{t,i}^O$, with corresponding reproduction
-# rates $R_{t,i}^L$ and $R_{t,i}^O$:
-
-#   lambda_{t,i} &= I_{t,i}^L R_{t,i}^L + I_{t,i}^O R_{t,i}^O
-#   R_{t,i}^L &= R_0 D_t e^{\epsilon_{t,i}^L}
-#   R_{t,i}^O &= R_0 Q_t e^{\epsilon_{t,i}^O}
-
-# where each of these reproduction rates is modelled as a product of: the
-# reproduction rate under initial conditions and no interventions $R_0$;
-# deterministic functions $D_t$ and $Q_t$ that modify $R_0$ over time to
-# respectively represent the impacts of physical distancing and quarantine
-# interventions at a national level, and correlated timeseries of random effects
-# $\epsilon_{t,i}^L$ and $\epsilon_{t,i}^O$ to represent stochastic fluctuations
-# in the reporting rate in each state, for example due to the clusters in
-# subpopulations with higher or lower reproduction rates than the general
-# population
-
-# We model the effect of $D_t$ as being proportional (on the log scale) to an
-# index of the proportional change in population mobility in response to
-# physical distancing measures $d_t$ (estimated from multiple datastreams of
-# poplulation mobility), which has initial value 0 before distancing measures
-# were implemented and value 1 at its maximum extent:
-
-#   D_t &= e^{\beta d_t}
-
-# We model $Q_t$ via a monotone decreasing step function with values constrained
-# to the unit interval, and with steps at the known dates $\tau_1$ and $\tau_2$ of
-# changes in quarantine policy:
-
[…]

### 2020-05-17 R/R_effective.R
tidy up naming of objects and outputs

@@ -160,27 +160,26 @@ iid_noise_kernel_L <- white(sigma_noise_L ^ 2)
 
 sigma_noise_O <- normal(0, 0.5, truncation = c(0, Inf))
 iid_noise_kernel_O <- white(sigma_noise_O ^ 2)
-sigma_bias_O <- normal(0, 0.5, truncation = c(0, Inf))
-bias_kernel_O <- bias(sigma_bias_O ^ 2)
 
 # combine noise kernels
 noise_kernel_L <- time_noise_kernel_L + iid_noise_kernel_L
-noise_kernel_O <- bias_kernel_O + iid_noise_kernel_O
+noise_kernel_O <- iid_noise_kernel_O
 
-# # for locally-acquired cases, do hierarchical smoother GPs as part of the state-level trends
-# lengthscale_time_signal_state_O <- lognormal(4, 0.5)
-# sigma_time_signal_state_O <- normal(0, 0.5, truncation = c(0, Inf))
+# for for local-local cases, use smooth term for state-level trends, for
+# import-local, use random intercepts
+lengthscale_time_signal_L <- lognormal(4, 0.5)
+sigma_time_signal_L <- normal(0, 0.25, truncation = c(0, Inf))
 
-lengthscale_time_signal_state_L <- lognormal(4, 0.5)
-sigma_time_signal_state_L <- normal(0, 0.25, truncation = c(0, Inf))
-
-signal_kernel_state_L  <- rbf(
-  lengthscales = lengthscale_time_signal_state_L,
-  variance = sigma_time_signal_state_L ^ 2
+signal_kernel_L  <- rbf(
+  lengthscales = lengthscale_time_signal_L,
+  variance = sigma_time_signal_L ^ 2
 )
 
-kernel_state_L <- signal_kernel_state_L + noise_kernel_L
-kernel_state_O <- noise_kernel_O
+sigma_bias_O <- normal(0, 0.5, truncation = c(0, Inf))
+signal_kernel_O <- bias(sigma_bias_O ^ 2)
+
+kernel_L <- signal_kernel_L + noise_kernel_L
+kernel_O <- signal_kernel_O + noise_kernel_O
 
 # variation in the Reff timeseries between the states and territories
 sigma_state_L <- normal(0, 0.25, truncation = c(0, Inf), dim = n_states)
@@ -199,7 +198,7 @@ v_O <- sweep(v_O_raw, 2, sigma_state_O, FUN = "*")
[…]

### 2020-05-17 R/functions.R

@@ -1073,7 +1073,7 @@ as.greta_array <- greta:::as.greta_array
 
 # copy of greta.gp::gp with v manually passed in, enabling different
 # (hierarchical) variance on each gp draw
-multi_gp <- function (x, v, kernel, inducing, tol = 1e-04) {
+multi_gp <- function (x, v, kernel, inducing = NULL, tol = 1e-04) {
   
   sparse <- !is.null(inducing)
   

### 2020-05-17 manuscript/ms.Rmd
add text and tables on parameter values

@@
 ---
+
+```{r setup, echo = FALSE, message = FALSE}
+library(dplyr)
+```
 
@@
 \begin{align}
-  D_t &= e^{\beta d_t}
+  D_t &= 1 - \beta d_t
 \end{align}
@@
  Q_t = \begin{cases}
-  q1 & t < \tau_1 \\
-  q2 & \tau_1 \leq t < \tau_2 \\
-  q3 & \tau_2 \leq t
+  q_1 & t < \tau_1 \\
+  q_2 & \tau_1 \leq t < \tau_2 \\
+  q_3 & \tau_2 \leq t
  \end{cases}
@@
 
-where $q1 > q2 > q3$ and all parameters are constrained to the unit interval.
+where $q_1 > q_2 > q_3$ and all parameters are constrained to the unit interval.
 
@@
   k_{SE}(t, t') &= \sigma_1^2 \, exp \left( -\frac{(t - t')^2}{2 l_1^2} \right) \\
-  k_{RQ}(t, t') &= \sigma_2^2 \, exp \left(1 + \frac{(t - t')^2}{2 \alpha l_1^2} \right)^{-\alpha} \\
+  k_{RQ}(t, t') &= \sigma_2^2 \, exp \left(1 + \frac{(t - t')^2}{2 \alpha l_2^2} \right)^{-\alpha} \\
   k_{IID}^L(t, t') &= \begin{cases}
@@
 Component 2 is therefore given by $R_0D_te^{\epsilon_{i}^{L,1}}$, and Component 3 (which we consider on the log-scale for ease of visualisation) is $\epsilon_{t,i}^{L,2}$.
+
+### Parameters and priors
+
+Tables \@ref(tab:mechanistic_priors) and \@ref(tab:timeseries_priors) give the prior distributions of parameters in the semi-mechanistic and timeseries ($\epsilon^L$ and $\epsilon^O$) parts of the model respectively.
+Table \@ref(tab:mechanistic_parameters) gives fixed parameter values usedin the semi-mechanistic part of the model.
+
+The parameters of the generation interval distribution are the posterior means of the lognormal distribution over the serial interval estimated by Nishiura et al. (2020).
+
+The parameters of lognormal prior over $R_0$ were computed such that the distribution matched the averages of the posterior means and 95% credible intervals for 11 European countries as estimated by Flaxman et al.
+(2020) (Imperial College report 13) in a sensitivity analysis where the mean generation interval was 5 days - similar to the serial interval distribution assumed here.
+This corresponds to a prior mean of 2.79, and a standard deviation of 1.70.
+
+We assumed slightly informative priors over the lengthscales of components 2 and 3 of the timeseries over R effective for local-local transmission (parameters $l_1$ and $l_2$), such that longer lengthscales were more likely for component 2 (timeseries of the state-wide average transmission rate) than for component 3 (timeseries of the average transmission rate over active cases).
+
+```{r mechanistic_priors, echo = FALSE}
+tibble::tribble(
+
+  ~"Prior distribution",
+  ~"Parameter description",
+
+  "$R_0 \\sim \\mathrm{lognormal}(1.02, 0.10^2)$",
+  "Basic reproduction number",
+
+  "$\\beta \\sim \\mathrm{U}(0, 1)$",
+  "Effect of social distancing behaviours",
+
+  "$q_1 \\sim \\mathrm{U}(0, 1)$",
+  "Effect of quarantine of overseas arrivals (phase 1)",
+
+  "$q_2 / q_1 \\sim \\mathrm{U}(0, 1)$",
+  "Relative effect of quarantine (phase 2 vs 1)",
+
+  "$q_3 / q_2 \\sim \\mathrm{U}(0, 1)$",
+  "Relative effect of quarantine (phase 3 vs 2)"
+
+) %>%
+  knitr::kable(caption = "Prior distributions for parameters in the semi-mechanistic part of the time-varying model of $R_eff$", escape = FALSE)
+```
+
+```{r mechanistic_parameters, echo = FALSE}
+tibble::tribble(
+
+  ~"Parameter value",
+  ~"Parameter description",
+
+  "$\\tau_1$ = 2020-03-15",
+  "Date of change from overseas quarantine policy phase 1 to 2",
+
+  "$\\tau_2$ = 2020-03-28",
+  "Date of change from overseas quarantine policy phase 2 to 3",
+
+  "$g(t) = \\int_{t-1}^{t} \\mathrm{lognormal}(t | 1.377, 0.567^2) \\mathrm{d}t$",
+  "Generation interval function",
+
+) %>%
+  knitr::kable(caption = "Values of fixed parameters in the semi-mechanistic part of the time-varying model of $R_eff$", escape = FALSE)
+```
+
+```{r timeseries_priors, echo = FALSE}
+tibble::tribble(
+
+  ~"Prior distribution",
+  ~"Parameter description",
+
+  "$\\sigma_1 \\sim \\mathrm{N}^+(0, 0.25^2)$",
+  "Amplitude of deviation; local $R_{eff}$ statewide",
+
+  "$l_1 \\sim \\mathrm{lognormal}(4, 0.5^2)$",
+  "Temporal correlation; local $R_{eff}$ statewide",
+
+  "$\\sigma_2 \\sim \\mathrm{N}^+(0, 0.25^2)$",
+  "Amplitude of deviation; local $R_{eff}$ active cases",
+
+  "$l_2 \\sim \\mathrm{lognormal}(2, 0.5^2)$",
+  "Temporal correlation; local $R_{eff}$ active cases",
+
+  "$\\alpha\\sim \\mathrm{lognormal}(2, 0.5^2)$",
+  "Correlation mixture weights; local $R_{eff}$ active cases",
+
+  "$\\sigma_3 \\sim \\mathrm{N}^+(0, 0.25^2)$",
+  "Overdispersion; local $R_{eff}$ active cases",
+
+  "$\\sigma_4 \\sim \\mathrm{N}^+(0, 0.5^2)$",
+  "Amplitude of deviation; import $R_{eff}$ statewide",
+
+  "$\\sigma_5\\sim \\mathrm{N}^+(0, 0.5^2)$",
+  "Overdispersion; import $R_{eff}$ active cases",
+
+) %>%
+  knitr::kable(caption = "Prior distributions for parameters used in the timeseries part of the time-varying model of $R_eff$", escape = FALSE)
+```
 

### 2020-05-17 manuscript/ms.Rmd
fix up ms and add modele fitting section

@@
   pdf_document:
+    keep_tex: true
   word_document:
@@
 
-  "$q_2 / q_1 \\sim \\mathrm{U}(0, 1)$",
+  "$q_2 \\times q_1 \\sim \\mathrm{U}(0, 1)$",
   "Relative effect of quarantine (phase 2 vs 1)",
 
-  "$q_3 / q_2 \\sim \\mathrm{U}(0, 1)$",
+  "$q_3 \\times q_2 \\sim \\mathrm{U}(0, 1)$",
   "Relative effect of quarantine (phase 3 vs 2)"
@@
 ```
+
+### Model fitting
+Inference was performed by Hamiltonian Monte Carlo using the R packages greta and greta.gp [@greta, @greta.gp].
+Posterior samples of model parameters were generated by 10 independent chains of a Hamiltonian Monte Carlo sampler, each run for 1000 itrations, after an initial, discarded, 'warmup' period during the sampler step size and diagonal mass matrix was tuned, and the regios of highest density located.
+Convergence was assessed by visual assessment of chains, and ensuring that the potential scale reduction factor for all parameters had values less than 1.1, and that there were at least 1000 effective samples for each parameter.
 

### 2020-05-18 R/R_effective.R
forecast Reff components 1+2 & 1+2+3 6 weeks into the future

@@ -54,7 +54,7 @@ dates <- seq(min(linelist$date), max(linelist$date), by = 1)
 
 n_states <- length(states)
 n_dates <- length(dates)
-n_extra <- 21
+n_extra <- as.numeric(Sys.Date() - max(dates)) + 7 * 6
 date_nums <- seq_len(n_dates + n_extra)
 
 # pad this with full set of dates, states, and import statuses
@@ -243,8 +243,16 @@ rel_new_vec <- exp(log_rel_new_from_loc_vec) + exp(log_rel_new_from_imp_vec)
 
 log_expected_infections_vec <- log_R0 + log(rel_new_vec)
 expected_infections_vec <- exp(log_expected_infections_vec)
-
 distribution(local_cases[valid]) <- poisson(expected_infections_vec)
+
+# # try negative binomial likelihood
+# valid <- which(is.finite(log_loc_infectious + log_imp_infectious), arr.ind = TRUE)
+# sqrt_inv_size <- normal(0, 0.5, truncation = c(0, Inf), dim = n_states)
+# size <- 1 / sqrt(sqrt_inv_size[valid[, 2]])
+# prob <- 1 / (1 + expected_infections_vec / size)
+# distribution(local_cases[valid]) <- negative_binomial(size, prob)
+
+# add correlated errors on the quarantine model too
 
 m <- model(beta, log_q, log_R0, rel_new_vec)
 draws <- mcmc(
@@ -308,7 +316,7 @@ for (type in 1:5) {
     rows <- seq_len(n_dates)
     projection_date <- NA
   } else {
-    last_date <- as.Date("2020-06-08")
+    last_date <- min(dates) + n_date_nums - 1
     n_projected <- n_dates + as.numeric(last_date - max(dates))
     rows <- pmin(n_dates + n_extra, seq_len(n_projected))
     projection_date <- max(dates)
@@ -606,28 +614,27 @@ for (type in 1:5) {
   
   write_csv(
     df_samples,
-    file.path(dir, "r_eff_local_samples.csv")
+    file.path(dir, "r_eff_123_local_samples.csv")
   )
   
-  # CSV of estimates based on these
-  mean <- rowMeans(R_eff_123_samples)
-  median <- apply(R_eff_123_samples, 1, FUN = stats::median)
-  ci90 <- apply(R_eff_123_samples, 1, quantile, c(0.05, 0.95))
[…]

### 2020-05-19 R/functions.R
update mobility analysis

@@ -94,7 +94,7 @@ google_mobility <- function() {
 # download and format Apple's mobility data - will need to update the url regularly
 apple_mobility <- function() {
   # get link from: https://www.apple.com/covid19/mobility
-  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2007HotfixDev55/v2/en-us/applemobilitytrends-2020-05-10.csv"
+  url <- "https://covid19-static.cdn-apple.com/covid19-mobility-data/2008HotfixDev30/v2/en-us/applemobilitytrends-2020-05-17.csv"
   data <- readr::read_csv(url) %>%
     tidyr::pivot_longer(
       cols = starts_with("2020-"),
@@ -155,7 +155,7 @@ apple_mobility <- function() {
 citymapper_mobility <- function() {
   
   # get link from: https://citymapper.com/cmi/about
-  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200511.csv"
+  url <- "https://cdn.citymapper.com/data/cmi/Citymapper_Mobility_Index_20200518.csv"
   data <- readr::read_csv(url, skip = 3) %>%
     tidyr::pivot_longer(cols = -Date,
                         names_to = "region",

### 2020-05-19 R/latent_social_distancing.R

@@ -241,8 +241,9 @@ for (j in seq_len(n_states)) {
   n_state_datastreams_j <- length(state_datastreams_j)
   
   par(mfrow = n2mfrow(n_state_datastreams_j),
-      mar = c(2, 2, 4, 2),
-      oma = c(0, 0, 2, 0))
+      mar = c(2, 2, 4, 3),
+      oma = c(0, 0, 2, 1),
+      las = 1)
   for (i in seq_len(n_state_datastreams_j)) {
     
     state_datastream_i <- state_datastreams_j[i]
@@ -265,7 +266,9 @@ for (j in seq_len(n_states)) {
          type = "n",
          ylim = ylim,
          ylab = "",
-         xlab = "")
+         xlab = "",
+         yaxt = "n")
+    axis(4)
     add_gridlines(interventions$date)
     add_mean_ci(est, dates_plot,
                 col = grey(0.9),
@@ -309,8 +312,8 @@ png("outputs/figures/multistate_model_fit.png",
     pointsize = 70)
 
 par(mfrow = c(8, 3),
-    mar = c(2, 2, 1, 1),
-    oma = c(1, 4, 6, 1),
+    mar = c(2, 2, 1, 3),
+    oma = c(1, 2, 6, 2),
     las = 1)
 
 for(this_state in states) {
@@ -353,7 +356,9 @@ for(this_state in states) {
            type = "n",
            ylim = ylim,
            ylab = "",
-           xlab = "")
+           xlab = "",
+           yaxt = "n")
+      axis(4)
       add_gridlines(interventions$date)
       add_mean_ci(est, dates_plot,
                   col = grey(0.9),
@@ -386,7 +391,7 @@ for(this_state in states) {
       state_name_formatted <- abbreviate_states(this_state)
       mtext(text = state_name_formatted,
             side = 2,
-            line = 3,
+            line = 1,
             cex = 1.6,
             xpd = NA,
             las = 0)
@@ -525,17 +530,30 @@ palette <- go_stop
 # write a function to create these plots
 
 # get positive value for proportional reduction in distancing
-distancing_change_data %>%
+p <- distancing_change_data %>%
   mutate(
     value = pmax(pmin(value, -0.01), -1),
     value = -value,
     col = case_when(
       !significant ~ grey(0.9),
       TRUE ~ palette[2]
-    )
+    ),
+    outer1 = 1.05,
+    outer2 = 1
   ) %>%
   ggplot() +
   # change this back to geom_circle, now there's scale_size?
[…]

### 2020-05-19 R/contacts.R
add analysis of baseline contacts rates

@@ -1,0 +1,124 @@
+# estimate the baseline numbers and durations of contacts with household members
+# and non-household members in Australia using data from Rolls et al.
+
+source("R/functions.R")
+
+rolls <- read_csv(
+  file = "data/contacts/rolls_et_al.csv",
+  col_types = cols(
+    participant_id = col_double(),
+    contact_id = col_double(),
+    contact_duration = col_double(),
+    location_duration = col_double(),
+    hh_member = col_double(),
+    weight = col_double()
+  ),
+  na = c("NULL", "", "NA")
+)
+
+contact_data <- rolls %>%
+  # sum durations for each contact
+  group_by(participant_id, contact_id, hh_member, weight) %>%
+  summarise(
+    contact_duration = sum(contact_duration),
+    location_duration = sum(location_duration)
+  ) %>%
+  # count contacts and average durations in household/non household for each
+  # participant
+  group_by(participant_id, hh_member, weight) %>%
+  summarise(
+    contacts = n_distinct(contact_id),
+    contact_duration = mean(contact_duration, na.rm = TRUE)
+  ) %>%
+  # convert duration to hours
+  mutate(contact_duration = contact_duration / (60)) %>%
+  ungroup() %>%
+  mutate(
+    hh_member = ifelse(hh_member == 1,
+                       "household",
+                       "non_household")
+  )
+  
+# compute weighted mean and standard error of numbers and average durations of contacts,
+# with household and non-household members
+baseline_contact_params <- contact_data %>%
+  group_by(hh_member) %>%
+  summarise(
+    mean_contacts = weighted_mean(
+      contacts,
+      w = weight
+    ),
+    se_mean_contacts = weighted_se(
+      contacts,
+      mean_contacts,
+      w = weight
+    ),
+    mean_avg_duration = weighted_mean(
+      contact_duration,
+      w = weight,
+      na.rm = TRUE
+    ),
+    se_avg_duration = weighted_se(
+      contact_duration,
+      mean_avg_duration,
+      w = weight,
+      na.rm = TRUE
+    )
+  )
+
+saveRDS(baseline_contact_params,
[…]

### 2020-05-19 R/functions.R

@@ -1152,3 +1152,19 @@ get_tests <- function() {
   
   df
 }
+
+# weighted mean and standard error of the weighted mean, computed with a
+# bootstrap
+weighted_mean <- weighted.mean
+
+weighted_se <- function(x, mu, w, na.rm = TRUE) {
+  x_sims <- replicate(10000,
+                      sample(x, length(x), replace = TRUE),
+                      simplify = FALSE)
+  means <- vapply(x_sims,
+                  weighted_mean,
+                  w,
+                  na.rm = na.rm,
+                  FUN.VALUE = numeric(1))
+  sd(means)
+}

### 2021-12-09 Rmd/case_date_autocorrelations.Rmd
autocorrelation of cases by onset and confirmation dates

```{r}
nndss_ll <- get_nndss_linelist()
vic_ll <- get_vic_linelist(file = "~/not_synced/vic/20211206_linelist_reff.csv")
nsw_ll <- get_nsw_linelist()

vl <- vic_ll %>% filter(import_status == "local", date_confirmation >= "2021-08-01")
vn <- nndss_ll %>% filter(import_status == "local", date_confirmation >= "2021-08-01", state == "VIC")

nl <- nsw_ll %>% filter(import_status == "local", date_confirmation >= "2021-08-01")
nn <- nndss_ll %>% filter(import_status == "local", date_confirmation >= "2021-08-01", state == "NSW")

dat <- bind_rows(
  vl %>% mutate(linelist = "state"),
  vn %>% mutate(linelist = "NINDSS"),
  nl %>% mutate(linelist = "state"),
  nn %>% mutate(linelist = "NINDSS"),
)
```

Confirmation dates
```{r}
dat %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_confirmation,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  )
```

Confirmation dates last month
```{r}
dat %>%
  filter(date_confirmation >= "2021-11-01") %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_confirmation,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  )
```

Onset dates
```{r}
dat %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_onset,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  ) +
  coord_cartesian(ylim = c(0,1))
```

```{r}
dat %>%
  ggplot() +
  geom_point(
    aes(x = date_onset, y = date_confirmation)
  ) +
  facet_grid(rows = vars(state), cols = vars(linelist)) +
  scale_x_date(date_breaks = "month") +
  theme(axis.text.x = element_text(angle = 270))

```

```{r}
dat %>%
  filter(date_onset >= "2021-08-01") %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_onset,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  ) 
```

Onset dates last month
```{r}
dat %>%
  filter(date_onset >= "2021-11-01") %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_onset,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  )
```

Detection dates
```{r}
dat %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_detection,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  ) +
  coord_cartesian(ylim = c(0,1))
```

Onset dates last month
```{r}
dat %>%
  filter(date_detection >= "2021-11-01") %>%
  ggplot() +
  geom_bar(
    aes(
      x = date_detection,
      fill = linelist
    ),
    position = "dodge"
  ) +
  facet_grid(
    rows = vars(state)
  )
```

```{r}
dates <- expand_grid(
  date = seq.Date(from = as.Date("2021-08-01"), to = as.Date("2021-12-05"), by = 1),
  state = c("NSW", "VIC"),
  linelist = c("state", "NINDSS")
)

dat_conf <- dat %>%
  group_by(state, linelist, date_confirmation) %>%
  summarise(cases = n(), .groups = "drop") %>%
  filter(!is.na(date_confirmation)) %>%
  full_join(
    dates %>% rename(date_confirmation = date)
  ) %>%
  arrange(state, linelist, date_confirmation) %>%
  mutate(cases = ifelse(is.na(cases), 0, cases))

dat_onset <- dat %>%
  group_by(state, linelist, date_onset) %>%
  summarise(cases = n(), .groups = "drop") %>%
  filter(!is.na(date_onset)) %>%
  full_join(
    dates %>% rename(date_onset = date)
  ) %>%
  arrange(state, linelist, date_onset) %>%
  mutate(cases = ifelse(is.na(cases), 0, cases))

dcn <- dat_conf %>% filter(state == "NSW", linelist == "state", date_confirmation >= "2021-10-01")
dcv <- dat_conf %>% filter(state == "VIC", linelist == "state", date_confirmation >= "2021-10-01")

don <- dat_onset %>% filter(state == "NSW", linelist == "state", date_onset >= "2021-10-01")
dov <- dat_onset %>% filter(state == "VIC", linelist == "state", date_onset >= "2021-10-01")
```

```{r}
library(tseries)
```

Autocorrelation

VIC

confirmation dates

since oct
```{r}
acf(dcv$cases)
```

Since Nov
```{r}
dcv %>% filter(date_confirmation >= "2021-11-01") %>% pull(cases) %>% acf
```

Onset dates

since Oct
```{r}
acf(dov$cases)
```

Since Nov
```{r}
dov %>% filter(date_onset >= "2021-11-01") %>% pull(cases) %>% acf
```

NSW

Confrimation dates 
since oct

```{r}
acf(dcn$cases)
```

since nov
```{r}
dcn %>% filter(date_confirmation >= "2021-11-01") %>% pull(cases) %>% acf
```

onset dates

since oct
```{r}
acf(don$cases)
```

since nov
```{r}
don %>% filter(date_onset >= "2021-11-01") %>% pull(cases) %>% acf
```

### 2021-12-09 R/optimal_partial_delays.R
pull out delay data for ecdf for optimal and partial per dates in national plan (and approximate with a lognormal in graph)

@@ -1,0 +1,64 @@
+
+delays_optimal <- delay_data_from_onset %>%
+  filter(
+    date_confirmation >= "2020-07-01",
+    date_confirmation <= "2021-01-31",
+    state == "NSW"
+  ) %>%
+  pull(delay)
+
+mean(delays_optimal)
+sd(delays_optimal)
+
+tibble(
+  delay = delays_optimal + 5,
+  type = "empirical"
+) %>% 
+  bind_rows(
+    tibble(
+      delay = rlnorm(1000, meanlog = log(7), sdlog = log(1.6)) %>% round,
+      type = "simulated"
+    )
+  ) %>%
+  ggplot() +
+  geom_bar(
+    aes(
+      x = delay
+    )
+  ) +
+  facet_grid(
+    rows = vars(type),
+    scales = "free_y"
+  )
+
+
+delays_partial <- delay_data_from_onset %>%
+  filter(
+    date_confirmation == "2020-08-04",
+    state == "VIC"
+  ) %>%
+  pull(delay)
+
+mean(delays_partial)
+sd(delays_partial)
+
+tibble(
+  delay = delays_partial + 5,
+  type = "empirical"
+) %>% 
+  bind_rows(
+    tibble(
+      delay = rlnorm(1000, meanlog = log(8), sdlog = log(1.8)) %>% round,
+      type = "simulated"
+    )
+  ) %>%
+  ggplot() +
+  geom_bar(
+    aes(
+      x = delay
+    )
+  ) +
+  facet_grid(
+    rows = vars(type),
+    scales = "free_y"
+  )

### 2021-12-09 R/get_sample_quantiles.R
more sample quantile examples

@@ -222,3 +222,128 @@ ggplot(c2_samples) +
   facet_wrap(~state) 
 
 
+c1_old <- read_reff_samples("outputs/projection_oldmatrix/r_eff_1_local_samples.csv") %>%
+  mutate(matrix = "old")
+c1_new <- read_reff_samples("outputs/projection/r_eff_1_local_samples.csv") %>%
+  mutate(matrix = "new")
+
+c12_old <- read_reff_samples("outputs/projection_oldmatrix/r_eff_12_local_samples.csv") %>%
+  mutate(matrix = "old")
+c12_new <- read_reff_samples("outputs/projection/r_eff_12_local_samples.csv") %>%
+  mutate(matrix = "new")
+
+
+c1 <- bind_rows(c1_old, c1_new)
+c12 <- bind_rows(c12_old, c12_new)
+
+ggplot(c1) +
+  geom_ribbon(aes(x = date, ymin = lw5, ymax = up95, col = matrix), alpha = 0) +
+  geom_ribbon(aes(x = date, ymin = lw25, ymax = up75, fill = matrix), alpha = 0.25) +
+  geom_line(aes(x = date, y = med, linetype = matrix), alpha = 0.25) +
+  facet_wrap(~state, ncol = 2) +
+  scale_fill_manual(values = c(green, "grey80")) +
+  scale_colour_manual(values = c(green, "grey80")) +
+  theme_cowplot() +
+  geom_vline(xintercept = as.Date("2021-02-22"), col = "steelblue") +
+  scale_y_continuous(limits = c(0, 5))
+
+ggplot(c12) +
+  geom_ribbon(aes(x = date, ymin = lw5, ymax = up95, col = matrix), alpha = 0) +
+  geom_ribbon(aes(x = date, ymin = lw25, ymax = up75, fill = matrix), alpha = 0.25) +
+  geom_line(aes(x = date, y = med, linetype = matrix), alpha = 0.25) +
+  facet_wrap(~state, ncol = 2) +
+  scale_fill_manual(values = c(green, "grey80")) +
+  scale_colour_manual(values = c(green, "grey80")) +
+  theme_cowplot() +
+  geom_vline(xintercept = as.Date("2021-02-22"), col = "steelblue") +
+  scale_y_continuous(limits = c(0, 5))
+  
+
+vthist <- read_csv("outputs/vaccine_effect_timeseries_2021-11-07.csv") %>%
[…]

### 2021-12-09 R/functions.R
handle vaccine status cols in NINDSS linelist temporary - will need updating when columns get regular ID name

@@ -4414,20 +4414,66 @@ get_nndss_linelist <- function(
       CV_DATE_ENTERED_QUARANTINE = "date"
     )
     
+    col_types_4 <- c(
+      STATE = "text",
+      POSTCODE = "numeric",
+      CONFIRMATION_STATUS = "text",
+      TRUE_ONSET_DATE = "date",
+      SPECIMEN_DATE = "date",
+      NOTIFICATION_DATE = "date",
+      NOTIFICATION_RECEIVE_DATE = "date",
+      Diagnosis_Date = "date",
+      AGE_AT_ONSET = "numeric",
+      SEX = "numeric",
+      DIED = "numeric",
+      PLACE_OF_ACQUISITION = "text",
+      HOSPITALISED = "numeric",
+      CV_ICU = "numeric",
+      CV_VENTILATED = "numeric",
+      OUTBREAK_REF = "text",
+      CASE_FOUND_BY = "numeric",
+      CV_SYMPTOMS = "text",
+      CV_OTHER_SYMPTOMS = "text",
+      CV_COMORBIDITIES = "text",
+      CV_OTHER_COMORBIDITIES = "text",
+      CV_GESTATION = "numeric",
+      #CV_CLOSE_CONTACT = "numeric"
+      CV_EXPOSURE_SETTING = "numeric",
+      CV_SOURCE_INFECTION = "numeric",
+      CV_SYMPTOMS_REPORTED = "numeric",
+      CV_QUARANTINE_STATUS = "numeric",
+      CV_DATE_ENTERED_QUARANTINE = "date",
+      "numeric",
+      "text",
+      "date",
+      "numeric",
+      "text",
+      "date",
+      "numeric",
+      "text",
+      "date",
+      "numeric",
+      "text",
+      "date",
+      "numeric",
+      "text",
+      "date"
+    )
+    
   }
   
   
   ll_date <- data$date_time[[1]]
   
   
-  if(ll_date < "2021-03-08"){
+  if (ll_date < "2021-03-08") {
     col_types <- col_types_1
+  } else if (ll_date < "2021-11-08") {
+    col_types <- col_types_2
+  } else if (ll_date < "2021-12-02") {
+    col_types <- col_types_3
   } else {
-    if(ll_date < "2021-11-08"){
-      col_types <- col_types_2
-    } else {
-        col_types <- col_types_3
-    }
+    col_types <- col_types_4
   }
   
   dat <- readxl::read_xlsx(

### 2021-12-10 R/functions.R
collate greta /tensorflow bits for easy commenting-out on non-greta machines

@@ -1,16 +1,24 @@
 source("R/lib.R")
 
 Sys.setenv(RETICULATE_AUTOCONFIGURE = FALSE)
+library(greta)
+library(greta.gp)
+library(tensorflow)
+
+tfp <- reticulate::import("tensorflow_probability")
+
+module <- greta::.internals$utils$misc$module
+fl <- greta:::fl
+tf_float <- greta:::tf_float
+
+
 library(readr)
 library(dplyr)
 library(stringr)
 library(rjson)
 library(tidyr)
-library(greta)
-library(greta.gp)
 library(readxl)
 library(RColorBrewer)
-library(tensorflow)
 library(purrr)
 library(ggplot2)
 library(R6)
@@ -21,11 +29,7 @@ library(rvest)
 library(magrittr)
 library(conmat)
 
-tfp <- reticulate::import("tensorflow_probability")
 
-module <- greta::.internals$utils$misc$module
-fl <- greta:::fl
-tf_float <- greta:::tf_float
 
 # read in and tidy up Facebook movement data
 facebook_mobility <- function() {

### 2021-12-10 R/functions.R
write out mask results from survey for UoA

@@ -2796,6 +2796,64 @@ format_raw_survey_data <- function(file = NULL, wave = NULL) {
   
   raw <- read_csv(file)
   
+  if ("Q222" %in% names(raw)) {
+    mask <- raw %>%
+      select(
+        state = S3,
+        date = StartDate,
+        face_covering = Q222
+      )%>%
+      mutate(
+        state = case_when(state == "ACT" ~ "Australian Capital Territory",
+                          TRUE ~ state),
+        state = abbreviate_states(state)
+      ) %>%
+      filter(!is.na(state)) %>%
+      mutate(
+        date = as.character(date),
+        date = as.Date(date, format = "%Y%m%d"),
+        date = min(date)
+      ) %>%
+      group_by(state, date) %>%
+      mutate(respondents = n()) %>%
+      group_by(state, date, face_covering, respondents) %>%
+      summarise(
+        count = n(),
+        .groups = "drop"
+      ) %>%
+      select(date,
+             state,
+             face_covering,
+             count,
+             respondents) %>%
+      arrange(state, date, face_covering)
+    
+  } else {
+    date_survey <- raw$StartDate %>%
+      as.character %>%
+      as.Date(format = "%Y%m%d") %>%
+      min
+    
+    expand_grid(
+      date = date_survey,
+      state = states,
+      face_covering = c("Always", "No", "Often", "Rarely", "Sometimes"),
+      count = 0,
+      respondents = 0
+    )
+  }
+  
+  
+  write_csv(
+    mask,
+    paste0(
+      "data/face_covering/face_covering_",
+      wave,
+      "_.csv"
+    )
+  )
+  
+  
   micro <- raw %>%
     select(
       state = S3,

### 2022-05-22 notes.Rmd
summary notes

---
title: "Notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### 10 May 2022
- Run latent_social_distancing.R for social_distancing_latent.RDS. Used under the hood for microdistancing_data()

- Need uom_surveys for under the hood functions in micro & macro distancing change. Eg: parse_all_surveys()

> ~/not_synced/uom_surveys/unlocked/SPSS Updated Data Covid19 Attitudes and Practices labels April2020.xlsx

#### 11 May

Plots look off! data object looks off

# 12 may

commonwealth script needs vaccination_effect.RDS for the reff_model_Data(). 
vaccination_effect.RDS generated in immunity_effect.R however to run that, I need scenario_parameters_omricon.csv which I cannot find

'outputs/scenario_parameters_omicron.csv' ??

'outputs/vaccination_effect.RDS'??

# Monday data

- survey data from mediaflux sit assessment repo

# Tuesday data

- not_synced

  + nndss 
  + nsw
  + pcr rat
  + vic
  + quantium forcasts

# Over view

> Effective reproduction number (Reff) — the average number of new infections
caused by an infected individual in the presence of public health interventions and where no assumption of 100% susceptibility is made.

- Reff > 1 = growing epidemic .

- If control measures, population immunity, or other factors can bring Reff below 1, then the epidemic is estimated to be in decline.

> When case incidence is very low or zero, current methods for Reff is not effective. Therefore ESTIMATING TP !  Additionally, the deviation between the Reff and the transmission potential.

> the ability of the virus, if it were present, to spread in a population, which we define as the ‘transmission potential.

- We separately model transmission from locally acquired cases (local-to-local trans-mission) and from overseas acquired cases (import-to-local transmission).

- We model local-to-local transmission (Reff) using two components:

  + Average trend in transmission driven by interventions that target local cases transmissionsuch as physical distancing and case targeted measures. **Component 1**

      we infer temporal trends in macro- and micro-distancing behaviour (sub-models a and b).

      + macro-distancing : reduction in the rate of non-household contacts;

      + micro-distancing : i.e., reduction in transmission probability per non-household contact.

      + using data on the number of days from symptom onset to case notification for cases, we estimate the proportion of cases that are detected (and thus advised to isolate) by each day post-infection.

      By quantifying the temporal change in the probability density for the time-to-detection (sub-model c), the model estimates how earlier isolation of cases — due to improvements in contact tracing, expanded access to testing, more inclusive case definitions, and other factors impacting detection rates — reduces the ability of SARS-CoV-2 to spread.

   + short-term fluctuations in Reff to capture stochastic dynamics of transmission, such as clusters of cases and short periods of lower-than-expected transmission **(Component 2, the ‘deviation’ between TP and Reff).**

  > Transmission potential (Component 1) reflects the average potential for the virus to spread at the population level. During times of disease activity, Component 2 measures how transmission within the sub-populations that have the most active cases at a given point in time differs compared to that expected from the population-wide TP.

  > When Component 2, the deviation between TP and Reff, is positively biased
(Reff > TP), it may indicate that transmission is concentrated in populations with higher-than- average levels of mixing, such as healthcare workers or meat processing workers. If negatively biased (Reff < TP), it reflects suppressed transmission compared to expectation. This may be due to an effective public health response actively suppressing transmission (e.g., through test, trace, isolation and quarantine), or other factors such as local depletion of susceptible
individuals, and/or the virus circulating in a sub-population with fewer-than-average social contacts.

# Model

## outputs

### mobility change

> gets mobility data and takes holidays, lockdowns and so on into consideration to model the mobility across different states at different places like workplace , residential places and so on.

-  state wise datastream_model
-  multi state model fit figs

### microdistancing

>micro-distancing, defined as the reduction in transmission probability per non-household
contact, and assessed through weekly nationwide surveys from which we estimate the
proportion of the population reporting always keeping 1.5 metre physical distance from
non-household contacts.

### macrodistancing

> macro-distancing, defined as the reduction in the average rate of non-household contacts, and assessed through weekly nationwide surveys of the daily number of non-household
contacts;

- gi_cdf? : generation interval distribution; used SI distribution from Nishiura et al

- cumulative density function 'cdf'

- why SI : An empirical estimate of the serial interval— the time from illness onset in a primary case (infector) to illness onset in a secondary case (infectee)—is needed to understand the
turnover of case generations and transmissibility of the disease.

- prior formed from the daily number of non-household contacts in a
pre-COVID-19 contact survey

- We account for variation among responses by day of the week by modelling
the fraction of the weekly number of contacts falling on each day of the week (the length-seven
vector in each state and time Di(t)) and using this to adjust the expected number of contacts
for each respondent based on the day of the week they completed the survey.

- To account for how
the weekly distribution of contacts has changed over time as a function of mixing restrictions
(e.g., a lower proportion of contacts on weekdays during periods when stay-at-home orders were
in place) we model the weekly distribution of contacts itself as a function of deviation in the
weekly average of the daily number of contacts

- Combining the baseline average daily contact rate NC0, mobility-driven modelled change
in contact rates over time δi(t), and time-varying day of the week effects Di(t) we obtain an
expected number of daily contacts for each survey response NCk

- We model the number of contacts from each survey respondent as a draw from an interval-
censored discrete lognormal distribution.

#### plots show :

- (light purple ribbons = 90% credible intervals; dark purple ribbons = 50% credible
intervals).

- Estimates are informed by state-level data from nationwide surveys (indicated by
the black lines and grey rectangles) and population mobility data.

- Green ticks indicate the dates that public holidays coincided with surveys (when people tend to stay home, biasing down the number of non-household contacts reported on those days).

## Surveillance model

- Generation interval : also referred to as the generation time, is the time lag between infection incidents in an infector–infectee pair. It is a proxy of serial interval (SI) of infectious disease, which represent the time lag between onsets of the symptoms in an infector–infectee pair

- We model the region- and time-varying distributions fi(t, t0) empirically via a time-series of empirical distribution functions computed from all observed infection-to-isolation periods observed within an adaptive moving window around each time t.

- we use 5 days prior to the date of
symptom onset to be the assumed date of infection, and the date of case notification to be the assumed date of isolation.

- For a given date and state/territory, the empirical distribution of delays from symptom onset to notification is computed from cases with symptom onset falling within a time window around that date, with the window selected to be the smallest that will yield at least 500 observations; but constrained to between one and eight weeks.

  + Specifically, if fewer than 100 cases, the national estimate is used, if more than 500 the state estimate is used, and if between 100 and 500 the distribution is a weighted average of state and national estimates.

### plots show

- Estimated trend in distributions of time from symptom onset to notification for locally acquired cases (black line = median; yellow ribbons = 90% distribution quantiles; black dots = time-to-notification of each case). Faded regions
indicate where a national trend is used due to low case counts.
